---
title: "Temperature_Analysis"
author: "Mandar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)
```
```{r}
rm(list = ls())
```


# Reading the data from the directory
```{r}
data <- readRDS(file = "/Users/mandarphatak/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")
data <- data |> filter(date < "2025-01-01")
data 
```
```{r}
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
```

# data tranformation :- We select the te
```{r echo=FALSE}
# selection of temperature data :- Tavg:Tmin
data_tbl <- data |> 
  dplyr::select(date,Tavg:Tmin,rel_hum,prec) 
```


```{r echo=FALSE}
# conversion to monthly data
monthly_tbl <-data_tbl |> 
  mutate(year = year(date),
         month = month(date) ) |> 
  group_by(year, month) |> 
  summarize(
    Tavg = mean(Tavg, na.rm =TRUE),
    Tmax = mean(Tmax, na.rm =TRUE),
    Tmin = mean(Tmin, na.rm =TRUE),
    prec = sum(prec,na.rm =  TRUE),
    rel_hum = mean(rel_hum, na.rm = TRUE),
    
    .groups = "drop"
    
  ) |> drop_na()

# # conversion to annual data
annual_tbl <- monthly_tbl |> 
  group_by(year) |> 
  summarize(
     Tavg = mean(Tavg, na.rm =TRUE),
    Tmax = mean(Tmax, na.rm =TRUE),
    Tmin = mean(Tmin, na.rm =TRUE),
    prec = sum(prec,na.rm =  TRUE),
    rel_hum = mean(rel_hum, na.rm = TRUE),
    .groups = "drop"
  )

annual_tbl |> dplyr:: select(-year) |> summary()
```
### Focusing on  Temperature data


## Data Over View :- We perform the following data manipulation for desired data overview 

```{r echo=FALSE}
annual_temperature_tbl_long <- annual_tbl |> 
  pivot_longer(
    cols = c(Tavg,Tmax,Tmin),
    names_to = "Temperature_Type",
    values_to = "Temperature"
  ) |> 
  filter(year < 2025)
```




```{r fig.width=15,fig.height=10}
annual_temperature_tbl_long$Temperature_Type <- factor(
  annual_temperature_tbl_long$Temperature_Type,
  levels = c("Tmax", "Tavg", "Tmin"))


trends_tbl <- annual_temperature_tbl_long %>%
  group_by(Temperature_Type) %>%
  summarize(trend = coef(lm(Temperature ~ year))[2]) %>%
  ungroup()


trends_tbl <- trends_tbl %>%
  mutate(
    label = case_when(
      Temperature_Type == "Tmax" ~ paste0("Maximum Temperature (Trend: +", round(trend, 4), "°C/year)"),
      Temperature_Type == "Tavg" ~ paste0("Average Temperature (Trend: +", round(trend, 4), "°C/year)"),
      Temperature_Type == "Tmin" ~ paste0("Minimum Temperature (Trend: +", round(trend, 4), "°C/year)")
    ),
    color = case_when(
      Temperature_Type == "Tmax" ~ "#ff1744",
      Temperature_Type == "Tavg" ~ "#fc8d59",
      Temperature_Type == "Tmin" ~ "#4575b4"
    ),
    y = case_when(  # adjust based on your plot’s Y-axis ranges
      Temperature_Type == "Tmax" ~ 15,
      Temperature_Type == "Tavg" ~ 10.55,
      Temperature_Type == "Tmin" ~ 6.5
    ),
    x = 1938  # horizontal position of the labels
  )

```

### Overview of Temperature Variation 1936-2024
```{r fig.width=15,fig.height=10}
ggplot(annual_temperature_tbl_long, aes(x = year, y = Temperature, colour = Temperature_Type)) + 
  geom_line(linewidth = 1) +
  geom_point(aes(fill = Temperature_Type), shape = 21, size = 4, stroke = 1, colour = "white") +
  # Add separate trend lines for each temperature type
  geom_smooth(aes(group = Temperature_Type), method = "lm", se = FALSE, linewidth = 1, 
              show.legend = FALSE, color = "black") +
  geom_text(
    data = trends_tbl,
    aes(x = x, y = y, label = label, color = Temperature_Type),
    fontface = "bold", hjust = 0, size = 7, show.legend = FALSE
  ) +
  
  # Fixed color scales with intuitive temperature colors
  scale_color_manual(
    values = c("Tmax" = "#ff1744",    # Hot red for maximum
               "Tavg" = "#fc8d59",    # Orange for average  
               "Tmin" = "#4575b4"),   # Cool blue for minimum
    labels = c("Tmax" = "Maximum Temperature",
               "Tavg" = "Average Temperature", 
               "Tmin" = "Minimum Temperature"),
    name = "Temperature Type"
  ) +
  scale_fill_manual(
    values = c("Tmax" = "#ff1744",    # Same colors for consistency
               "Tavg" = "#fc8d59",    
               "Tmin" = "#4575b4"),
    guide = "none"  # Hide fill legend since color legend shows the same info
  ) +
  
  scale_x_continuous(breaks = c(seq(1936, 2024, 10), 2024), 
                     expand = c(0.02, 0.02), 
                     limits = c(1936, 2024)) +
  scale_y_continuous(
    limits = c(1, 16),
    breaks = seq(1, 16, by = 2),
    labels = function(x) paste0(format(x, nsmall = 0), "°C"),
    expand = c(0, 0)
  ) +
  
  labs(
    title = "Temperature Variation (1936-2025)",
    subtitle = "Hamburg Fuhlsbüttel",
    x = "",
    y = "Temperature (°C)"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0, face = "bold", colour = "black", size = 30, margin = margin(b = 20)),
    plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black", size = 30, margin = margin(b = 10)),
    axis.line = element_line(color = "black", linewidth = 1),
    axis.title = element_text(face = "bold", color = "black", size = 22),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black", size = 20),
    axis.text.y = element_text(face = "bold", color = "black", size = 20),
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    legend.position = "bottom",  # Better positioning for temperature legend
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
   guides(color = "none", fill = "none")
  
```
### Decadal Changes in Temperature 

```{r}

 annual_temperature_tbl_long <- annual_temperature_tbl_long |>
  mutate(decade_group = case_when(
    year >= 1936 & year <= 1939 ~ "1936–1939",
    year >= 1940 & year <= 1949 ~ "1940–1949",
    year >= 1950 & year <= 1959 ~ "1950–1959",
    year >= 1960 & year <= 1969 ~ "1960–1969",
    year >= 1970 & year <= 1979 ~ "1970–1979",
    year >= 1980 & year <= 1989 ~ "1980–1989",
    year >= 1990 & year <= 1999 ~ "1990–1999",
    year >= 2000 & year <= 2009 ~ "2000–2009",
    year >= 2010 & year <= 2019 ~ "2010–2019",
    year >= 2020 & year <= 2024 ~ "2020–2024",
    TRUE ~ NA_character_
  ))

  decadal_summary_tbl <- annual_temperature_tbl_long %>%
  group_by(decade_group, Temperature_Type) %>%
  summarise(decadal_mean = mean(Temperature, na.rm = TRUE), .groups = "drop")
  
library(ggplot2)

ggplot(decadal_summary_tbl, aes(x = decade_group, y = decadal_mean, fill = Temperature_Type)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  scale_fill_manual(
    values = c("Tmax" = "#ff1744", "Tavg" = "#fc8d59", "Tmin" = "#4575b4"),
    name = "Temperature Type",
    labels = c("Tmax" = "Maximum", "Tavg" = "Average", "Tmin" = "Minimum")
  ) +
  labs(
    title = "Decadal Mean Temperature (1936–2024)",
    subtitle = "Hamburg Fuhlsbüttel",
    x = "Decade",
    y = "Mean Temperature (°C)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold")
  )

```
```{r}
decadal_diff_tbl <- decadal_summary_tbl %>%
  arrange(Temperature_Type, decade_group) %>%
  group_by(Temperature_Type) %>%
  mutate(
    change_from_prev = decadal_mean - lag(decadal_mean),
    percent_change = 100 * change_from_prev / lag(decadal_mean)
  ) %>%
  ungroup()

ggplot(decadal_summary_tbl, aes(x = decade_group, y = decadal_mean, group = Temperature_Type, color = Temperature_Type)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(
    values = c("Tmax" = "#ff1744", "Tavg" = "#fc8d59", "Tmin" = "#4575b4"),
    name = "Temperature Type",
    labels = c("Tmax" = "Maximum", "Tavg" = "Average", "Tmin" = "Minimum")
  ) +
  labs(
    title = "Decadal Temperature Trends (1936–2024)",
    subtitle = "Hamburg Fuhlsbüttel",
    x = "",
    y = "Mean Temperature (°C)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold")
  )

```
## Decadal changes in Temperature 
```{r}
decadal_diff_tbl <- decadal_summary_tbl %>%
  arrange(Temperature_Type, decade_group) %>%
  group_by(Temperature_Type) %>%
  mutate(decadal_diff = decadal_mean - lag(decadal_mean)) %>%
  filter(!is.na(decadal_diff)) %>%
  ungroup()

ggplot(decadal_diff_tbl, aes(x = decade_group, y = decadal_diff, fill = Temperature_Type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_label(aes(label = round(decadal_diff, 2)),
           position = position_dodge(width = 0.7),
           size = 4.5, fontface = "bold",
           show.legend = FALSE) +
  # geom_text(aes(label = round(decadal_diff, 2)), 
  #           position = position_dodge(width = 0.7), 
  #           vjust = ifelse(decadal_diff_tbl$decadal_diff >= 0, -0.3, 1.3), 
  #           size = 5, fontface = "bold") +
  scale_fill_manual(
    values = c("Tmax" = "#ff1744", "Tavg" = "#fc8d59", "Tmin" = "#4575b4"),
    name = "Temperature Type",
    labels = c("Tmax" = "Maximum", "Tavg" = "Average", "Tmin" = "Minimum")
  ) +
  labs(
  title = "Decadal Changes in Temperature at Hamburg-Fuhlsbüttel",
  subtitle = "Average, Minimum, and Maximum Temperatures – Change from Previous Decade (°C)"
) +

  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 20),
    axis.text.x = element_text(angle = 30, hjust = 1),
    axis.text.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold")
  )
```
#### Extreme Events :- We visualize the intensity and duration of heat events from 1936-2024.
```{r}
library(ggplot2)
library(dplyr)
library(scales)
library(ggrepel)  # For repelling labels






identify_heatwaves <- function(date, tmax, threshold, min_duration = 3) {
  # Ensure inputs are valid
  stopifnot(length(date) == length(tmax))
  stopifnot(is.numeric(tmax))
  stopifnot(inherits(date, "Date"))
  
  # Sort and prepare
  data <- data.frame(date = date, tmax = tmax)
  data <- data[order(data$date), ]
  data$above_threshold <- data$tmax > threshold
  
  # Initialize output
  heatwaves <- data.frame(
    start_date = as.Date(character()),
    end_date = as.Date(character()),
    duration = integer(),
    avg_intensity = numeric(),
    total_intensity = numeric(),
    max_intensity = numeric(),
    mean_tmax = numeric(),
    peak_tmax = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Tracking flags
  in_heatwave <- FALSE
  heatwave_start <- NULL
  heatwave_length <- 0
  
  # Loop through
  for (i in seq_len(nrow(data))) {
    if (data$above_threshold[i]) {
      if (!in_heatwave) {
        in_heatwave <- TRUE
        heatwave_start <- data$date[i]
        heatwave_length <- 1
      } else {
        heatwave_length <- heatwave_length + 1
      }
    } else {
      if (in_heatwave && heatwave_length >= min_duration) {
        end_date <- data$date[i - 1]
        heatwave_days <- which(data$date >= heatwave_start & data$date <= end_date)
        exceedance <- data$tmax[heatwave_days] - threshold
        
        heatwaves <- rbind(heatwaves, data.frame(
          start_date = heatwave_start,
          end_date = end_date,
          duration = heatwave_length,
          avg_intensity = mean(exceedance),
          total_intensity = sum(exceedance),
          max_intensity = max(exceedance),
          mean_tmax = mean(data$tmax[heatwave_days]),
          peak_tmax = max(data$tmax[heatwave_days])
        ))
      }
      in_heatwave <- FALSE
      heatwave_length <- 0
    }
  }
  
  # Final check at end
  if (in_heatwave && heatwave_length >= min_duration) {
    end_date <- data$date[nrow(data)]
    heatwave_days <- which(data$date >= heatwave_start & data$date <= end_date)
    exceedance <- data$tmax[heatwave_days] - threshold
    
    heatwaves <- rbind(heatwaves, data.frame(
      start_date = heatwave_start,
      end_date = end_date,
      duration = heatwave_length,
      avg_intensity = mean(exceedance),
      total_intensity = sum(exceedance),
      max_intensity = max(exceedance),
      mean_tmax = mean(data$tmax[heatwave_days]),
      peak_tmax = max(data$tmax[heatwave_days])
    ))
  }
  
  # Add year column
  if (nrow(heatwaves) > 0) {
    heatwaves$year <- as.numeric(format(heatwaves$start_date, "%Y"))
  }
  
  return(heatwaves)
}

```


```{r}
heatwave_events_tbl <- identify_heatwaves(
  date = data$date,
  tmax = data$Tmax,
  threshold = 28,
  min_duration = 3
)
```





```{r}
heatwave_count_per_year <- heatwave_events_tbl %>%
  count(year, name = "num_events")
```


```{r}
heatwave_summary_tbl <- heatwave_events_tbl %>%
  group_by(year) %>%
  summarise(
    num_events = n(),
    total_duration = sum(duration, na.rm = TRUE),
    avg_duration = mean(duration, na.rm = TRUE),
    mean_avg_intensity = mean(avg_intensity, na.rm = TRUE),
    mean_max_intensity = mean(max_intensity, na.rm = TRUE),
    total_intensity = sum(total_intensity, na.rm = TRUE),
    max_peak_tmax = max(peak_tmax, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r}
# Convert date columns
 heatwave_events_tbl$start_date <- as.Date(heatwave_events_tbl$start_date)
heatwave_events_tbl$end_date <- as.Date(heatwave_events_tbl$end_date)

heatwave_events_tbl$year <- as.numeric(heatwave_events_tbl$year)
```


```{r}
# Custom color palette
# Custom color palette
custom_colors <- c("#f8f800", "#fdc70c", "#f3903f", "#ff4d00", "#e60000", "#5e0000")

# Identify recent events since 1990 with avg_intensity > 5
recent_events <- heatwave_events_tbl %>% 
  filter(year > 1990, avg_intensity > 5) %>%
  summarize(count = n()) %>%
  pull(count)

# Total events and percentage
total_events <- nrow(heatwave_events_tbl)
percent_recent <- round((recent_events / total_events) * 100)

# 95th percentile extreme threshold (for avg_intensity)
threshold_95 <- quantile(heatwave_events_tbl$avg_intensity, 0.95, na.rm = TRUE)

# Get max intensity (avg_intensity) and longest duration
max_intensity_value <- max(heatwave_events_tbl$avg_intensity, na.rm = TRUE)
longest_duration_value <- max(heatwave_events_tbl$duration, na.rm = TRUE)

# Label max avg_intensity event
max_intensity_event <- heatwave_events_tbl %>%
  filter(avg_intensity == max_intensity_value) %>%
  mutate(label = paste0("Max intensity: ", round(avg_intensity, 1), "\n", format(start_date, "%b %d, %Y")))

# Label longest duration event
longest_duration_event <- heatwave_events_tbl %>%
  filter(duration == longest_duration_value) %>%
  mutate(label = paste0("Longest: ", duration, " days\n", format(start_date, "%b %Y")))

# Filter extreme events based on avg_intensity, exclude max to avoid duplicate label
extreme_events <- heatwave_events_tbl %>%
  filter(avg_intensity > threshold_95, avg_intensity < max_intensity_value) %>%
  mutate(label = paste0("Extreme: ", round(avg_intensity, 1), "\n", format(start_date, "%Y")))

```


```{r}
p <- ggplot(heatwave_events_tbl, aes(x = as.numeric(year), y = avg_intensity)) +
  
  geom_point(aes(size = duration, color = avg_intensity), alpha = 0.85) +
  
  geom_vline(xintercept = 1990, linetype = "dashed", color = "blue", linewidth = 0.6) +
  annotate("text", x = 1990, y = 6.2, label = "1990", 
           color = "blue", fontface = "bold", size = 3) +
  
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = "darkblue", linetype = "solid", linewidth = 0.8, alpha = 0.6) +
  
  geom_text(data = max_intensity_event, aes(label = label),
            hjust = 1, vjust = 1.2, size = 3, fontface = "bold", color = "#5e0000") +
  
  geom_hline(yintercept = threshold_95, linetype = "dotted", color = "darkred") +
  annotate("text", x = 1940, y = threshold_95, label = "95th Percentile", 
           color = "darkred", size = 3) +
  
  # Annotate longest duration using repel to avoid overlap
  ggrepel::geom_text_repel(data = longest_duration_event, aes(label = label),
                           size = 3, fontface = "bold", color = "#5e0000",
                           max.overlaps = 5, box.padding = 0.4, point.padding = 0.4,
                           hjust = 1) +
  
  # Annotate extreme events
  ggrepel::geom_text_repel(data = extreme_events, aes(label = label),
                           size = 2.8, fontface = "italic", color = "black",
                           max.overlaps = 10, box.padding = 0.4, point.padding = 0.3) +
  
  # Axis scales
  scale_y_continuous(limits = c(0, 6.25), breaks = seq(0, 6.25, 0.5),
                     expand = c(0, 0.05)) +
  
  scale_x_continuous(limits = c(1936, 2024),
                     breaks = c(seq(1936, 2024, by = 10), 2024)) +
  
  scale_color_gradientn(colors = custom_colors,
                        name = "Avg. Intensity",
                        guide = guide_colorbar(barwidth = 1, barheight = 10,
                                               title.position = "top", title.hjust = 0.5)) +
  
  scale_size_continuous(range = c(2, 10),
                        name = "Duration (days)",
                        breaks = c(3, 6, 9, 12),
                        guide = guide_legend(title.position = "top",
                                             title.hjust = 0.5,
                                             override.aes = list(color = "darkblue"))) +
  
  # Labels
  labs(
    title = "Heatwave Events in Hamburg Fuhlsbüttel (1936–2025)",
    subtitle = paste0("Based on threshold of 28°C for minimum 3 consecutive days\n",
                      percent_recent, "% of all events occurred since 1990"),
    x = "",
    y = "Avg. Intensity (°C above threshold)",
    caption = "Source: Deutscher Wetterdienst (DWD)"
  ) +
  
  # Theme
  theme_minimal(base_size = 12) +
  theme_minimal(base_size = 12) +
theme(
  plot.title.position = "plot",
  plot.title = element_text(face = "bold", size = 18, margin = margin(b = 15)),
  plot.subtitle = element_text(size = 15, color = "gray30", margin = margin(b = 15), face = "bold"),
  axis.title.y = element_text(margin = margin(r = 10), face = "bold"),
  axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold", color = "gray30", size = 12),
  axis.text.y  = element_text(face = "bold", color = "gray30", size = 12),
  legend.position = "right",  # ✅ Only once
  legend.box = "vertical",
  legend.margin = margin(l = 10),
  legend.title = element_text(face = "bold"),
  panel.grid.minor = element_blank(),
  panel.grid.major.x = element_blank(),
  panel.grid.major.y = element_line(color = "gray80", linetype = "dotted")
)


p

```
### Max Intensity
```{r}
# ===== MAX INTENSITY ANALYSIS =====
recent_events_max <- heatwave_events_tbl %>%
  filter(year > 1990, max_intensity > 5) %>%
  summarize(count = n()) %>%
  pull(count)

percent_recent_max <- round((recent_events_max / total_events) * 100)

threshold_95_max <- quantile(heatwave_events_tbl$max_intensity, 0.95, na.rm = TRUE)

max_max_intensity <- max(heatwave_events_tbl$max_intensity, na.rm = TRUE)

max_max_event <- heatwave_events_tbl %>%
  filter(max_intensity == max_max_intensity) %>%
  mutate(label = paste0("Max exceedance: ", round(max_intensity, 1), "\n", format(start_date, "%b %d, %Y")))

extreme_events_max <- heatwave_events_tbl %>%
  filter(max_intensity > threshold_95_max, max_intensity < max_max_intensity) %>%
  mutate(label = paste0("Extreme: ", round(max_intensity, 1), "\n", format(start_date, "%Y")))

```
```{r}
p_max <- ggplot(heatwave_events_tbl, aes(x = as.numeric(year), y = max_intensity)) +
  
  geom_point(aes(size = duration, color = max_intensity), alpha = 0.85) +
  
  geom_vline(xintercept = 1990, linetype = "dashed", color = "blue", linewidth = 0.6) +
  annotate("text", x = 1990, y = 12, label = "1990", 
           color = "blue", fontface = "bold", size = 3) +
  
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = "darkblue", linetype = "solid", linewidth = 0.8, alpha = 0.6) +
  
  geom_text(data = max_max_event, aes(label = label),
            hjust = 1, vjust = 1.2, size = 3, fontface = "bold", color = "#5e0000") +
  
  geom_hline(yintercept = threshold_95_max, linetype = "dotted", color = "darkred") +
  annotate("text", x = 1940, y = threshold_95_max, label = "95th Percentile", 
           color = "darkred", size = 3) +
  
  ggrepel::geom_text_repel(data = extreme_events_max, aes(label = label),
                           size = 2.8, fontface = "italic", color = "black",
                           max.overlaps = 10, box.padding = 0.4, point.padding = 0.3) +

  scale_y_continuous(limits = c(0, 12), breaks = seq(0, 12, 0.5),
                     expand = c(0, 0.05)) +
  
  scale_x_continuous(limits = c(1936, 2024),
                     breaks = c(seq(1936, 2024, by = 10), 2024)) +
  
  scale_color_gradientn(colors = custom_colors,
                        name = "Max Intensity",
                        guide = guide_colorbar(barwidth = 1, barheight = 10,
                                               title.position = "top", title.hjust = 0.5)) +
  
  scale_size_continuous(range = c(2, 10),
                        name = "Duration (days)",
                        breaks = c(3, 6, 9, 12),
                        guide = guide_legend(title.position = "top",
                                             title.hjust = 0.5,
                                             override.aes = list(color = "darkblue"))) +
  
  labs(
    title = "Max Daily Exceedance in Heatwave Events (1936–2025)",
    subtitle = paste0("Based on max daily exceedance above 28°C threshold\n",
                      percent_recent_max, "% of all events occurred since 1990"),
    x = "",
    y = "Max Intensity (°C above threshold)",
    caption = "Source: Deutscher Wetterdienst (DWD)"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 18, margin = margin(b = 15)),
    plot.subtitle = element_text(size = 15, color = "gray30", margin = margin(b = 15), face = "bold"),
    axis.title.y = element_text(margin = margin(r = 10), face = "bold"),
    axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold", color = "gray30", size = 12),
    axis.text.y  = element_text(face = "bold", color = "gray30", size = 12),
    legend.position = "right",
    legend.box = "vertical",
    legend.margin = margin(l = 10),
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray80", linetype = "dotted")
  )

p_max

```


```{r}
heatwave_events_tbl[heatwave_events_tbl$max_intensity > 5,]

```

##### heatwavR #### library
```{r}
library(dplyr)
library(lubridate)
library(extRemes)

data <- data %>% filter(!is.na(Tmax))  # Clean NA

start_year <- 1936
end_year <- 2024
window_length <- 32
```


```{r}
results <- list()

for (year_to_analyze in (start_year + window_length):end_year) {
  
  # Define reference period: 32 years before the year_to_analyze
  ref_start <- year_to_analyze - window_length
  ref_end <- year_to_analyze - 1
  
  # Extract reference Tmax
  Tref <- data %>%
    filter(year(date) >= ref_start, year(date) <= ref_end) %>%
    pull(Tmax)
  
  # Extract analysis year Tmax
  Temp <- data %>%
    filter(year(date) == year_to_analyze) %>%
    pull(Tmax)
  
  # Skip if data missing or incomplete
  if(length(Tref) < 365 * window_length || length(Temp) < 365){
    results[[as.character(year_to_analyze)]] <- NA
    next
  }
  
  # Run hwmid
  results[[as.character(year_to_analyze)]] <- hwmid(
    yTref = ref_start,
    Tref = Tref,
    yTemp = year_to_analyze,
    Temp = Temp
  )
  
  cat("Processed year:", year_to_analyze, "\n")
}

```
```{r}
results

```
```{r}
data <- data |> mutate(year = year(date), month = month(date))
```

```{r}
 #Define reference period (must be exactly 32 years)
ref_start_year <- 1958
ref_end_year <- 1989  # 1981 + 32 - 1 = 2012

# Filter data for 32-year reference period
ref_data <- data[data$year >= ref_start_year & data$year <= ref_end_year, ]
Tref <- ref_data$Tmax

# Define detection period 
detect_start_year <- 1990  # or whatever year you want to start detection

# Filter data for detection period
detect_data <- data[data$year >= detect_start_year, ]
Temp <- detect_data$Tmax

# Run hwmid with single year values (not vectors)
heatwaves <- hwmid(yTref = ref_start_year,    # Single value: 1981
                   Tref = Tref,               # 32 years of daily Tmax
                   yTemp = detect_start_year, # Single value: 1936  
                   Temp = Temp)               # All years for detection
#The key fix: yTref and yTemp should be single year values, not vectors of years. Try this corrected approach!
```

```{r}
# Install required packages (uncomment if needed)
# install.packages(c("plotly", "dplyr", "ggplot2", "htmlwidgets", "DT"))

# Load libraries
library(plotly)
library(dplyr)
library(ggplot2)
library(htmlwidgets)
library(DT)

# ============================================================================
# DATA PREPARATION
# ============================================================================

# Your heatwave data (from hwmid analysis)
magnitude <- c(0.0000000, 0.2654867, 1.3451327, 0.0000000, 7.4513274, 1.9292035, 
               2.0353982, 1.7699115, 0.0000000, 1.3274336, 4.3008850, 0.5132743,
               0.9911504, 6.3362832, 0.5309735, 1.3274336, 3.5752212, 1.2566372,
               2.4424779, 0.4424779, 5.2212389, 0.0000000, 1.6460177, 1.1327434,
               1.2389381, 3.3628319, 0.9557522, 0.0000000, 11.2389381, 4.0176991,
               8.3008850, 3.3805310, 4.9557522, 0.4247788, 0.7964602)

days <- c(0, 3, 4, 0, 8, 7, 3, 5, 0, 7, 4, 4, 5, 13, 7, 3, 4, 3, 5, 3, 5, 0, 
          3, 4, 3, 5, 3, 0, 12, 6, 13, 5, 3, 7, 3)

peak_day <- c(0, 187, 179, 0, 202, 229, 158, 236, 0, 211, 170, 235, 209, 212, 
              217, 146, 198, 195, 207, 182, 189, 0, 231, 202, 199, 182, 237, 0, 
              205, 204, 218, 167, 199, 248, 177)

# Create main dataset
years <- 1990:2024
heatwave_data <- data.frame(
  year = years,
  magnitude = magnitude,
  days = days,
  peak_day = ifelse(peak_day == 0, NA, peak_day),
  has_heatwave = magnitude > 0,
  decade = paste0(floor((years - 1990) / 10) * 10 + 1990, "s")
)

# ============================================================================
# SUMMARY STATISTICS
# ============================================================================

# Calculate key statistics
stats <- list(
  total_years = nrow(heatwave_data),
  years_with_heatwaves = sum(heatwave_data$has_heatwave),
  frequency_pct = round(sum(heatwave_data$has_heatwave) / nrow(heatwave_data) * 100, 1),
  max_magnitude = max(heatwave_data$magnitude),
  max_magnitude_year = heatwave_data$year[which.max(heatwave_data$magnitude)],
  max_days = max(heatwave_data$days),
  avg_magnitude = round(mean(heatwave_data$magnitude[heatwave_data$magnitude > 0]), 2),
  avg_days = round(mean(heatwave_data$days[heatwave_data$days > 0]), 1)
)

# Print summary
cat("=== HEATWAVE ANALYSIS SUMMARY (1990-2024) ===\n")
cat("Reference Period: 1958-1989\n")
cat("Detection Period: 1990-2024\n\n")
cat("Years with heatwaves:", stats$years_with_heatwaves, "/", stats$total_years, 
    "(", stats$frequency_pct, "%)\n")
cat("Maximum magnitude:", stats$max_magnitude, "in", stats$max_magnitude_year, "\n")
cat("Maximum days:", stats$max_days, "\n")
cat("Average magnitude (non-zero years):", stats$avg_magnitude, "\n")
cat("Average days (non-zero years):", stats$avg_days, "\n\n")

# ============================================================================
# VISUALIZATION 1: MAGNITUDE TIME SERIES
# ============================================================================

p1 <- heatwave_data %>%
  plot_ly(x = ~year, y = ~magnitude, type = 'scatter', mode = 'lines+markers',
          line = list(color = '#dc2626', width = 2),
          marker = list(color = '#dc2626', size = 6),
          hovertemplate = '<b>Year:</b> %{x}<br><b>Magnitude:</b> %{y:.2f}<extra></extra>') %>%
  add_trace(y = rep(stats$avg_magnitude, length(years)), 
            type = 'scatter', mode = 'lines',
            line = list(color = '#fbbf24', width = 2, dash = 'dash'),
            name = paste('Average (', stats$avg_magnitude, ')'),
            hovertemplate = '<b>Average Magnitude:</b> %{y:.2f}<extra></extra>') %>%
  layout(title = list(text = '<b>Heatwave Magnitude Index Over Time</b>', 
                      font = list(size = 16)),
         xaxis = list(title = 'Year'),
         yaxis = list(title = 'Magnitude Index'),
         showlegend = TRUE,
         hovermode = 'x unified')

# Display plot
p1

# ============================================================================
# VISUALIZATION 2: HEATWAVE DAYS BAR CHART
# ============================================================================

p2 <- heatwave_data %>%
  plot_ly(x = ~year, y = ~days, type = 'bar',
          marker = list(color = ~ifelse(days > 10, '#dc2626', '#ef4444')),
          hovertemplate = '<b>Year:</b> %{x}<br><b>Days:</b> %{y}<extra></extra>') %>%
  layout(title = list(text = '<b>Number of Heatwave Days per Year</b>', 
                      font = list(size = 16)),
         xaxis = list(title = 'Year'),
         yaxis = list(title = 'Number of Days'),
         showlegend = FALSE)

# Display plot
p2

# ============================================================================
# VISUALIZATION 3: MAGNITUDE vs DAYS SCATTER
# ============================================================================

p3 <- heatwave_data %>%
  filter(has_heatwave) %>%
  plot_ly(x = ~magnitude, y = ~days, type = 'scatter', mode = 'markers',
          marker = list(size = 10, color = '#dc2626', opacity = 0.7),
          text = ~paste("Year:", year),
          hovertemplate = '<b>Year:</b> %{text}<br><b>Magnitude:</b> %{x:.2f}<br><b>Days:</b> %{y}<extra></extra>') %>%
  layout(title = list(text = '<b>Heatwave Magnitude vs Duration</b>', 
                      font = list(size = 16)),
         xaxis = list(title = 'Magnitude Index'),
         yaxis = list(title = 'Number of Days'))

# Display plot
p3

# ============================================================================
# VISUALIZATION 4: DECADE COMPARISON
# ============================================================================

# Calculate decade statistics
decade_stats <- heatwave_data %>%
  mutate(decade = case_when(
    year >= 1990 & year < 2000 ~ "1990s",
    year >= 2000 & year < 2010 ~ "2000s", 
    year >= 2010 & year < 2020 ~ "2010s",
    year >= 2020 ~ "2020s"
  )) %>%
  group_by(decade) %>%
  summarise(
    frequency = round(sum(has_heatwave) / n() * 100, 1),
    avg_magnitude = round(mean(magnitude[magnitude > 0]), 2),
    total_days = sum(days),
    max_magnitude = max(magnitude),
    .groups = 'drop'
  )

# Create decade comparison plot
p4 <- decade_stats %>%
  plot_ly() %>%
  add_trace(x = ~decade, y = ~frequency, type = 'bar', name = 'Frequency (%)',
            marker = list(color = '#dc2626'),
            yaxis = 'y', offsetgroup = 1) %>%
  add_trace(x = ~decade, y = ~avg_magnitude, type = 'bar', name = 'Avg Magnitude',
            marker = list(color = '#fbbf24'),
            yaxis = 'y2', offsetgroup = 2) %>%
  layout(title = list(text = '<b>Decade Comparison</b>', font = list(size = 16)),
         xaxis = list(title = 'Decade'),
         yaxis = list(side = 'left', title = 'Frequency (%)', color = '#dc2626'),
         yaxis2 = list(side = 'right', overlaying = 'y', title = 'Average Magnitude', color = '#fbbf24'),
         barmode = 'group')

# Display plot  
p4

# ============================================================================
# VISUALIZATION 5: SEASONAL TIMING (Peak Day Analysis)
# ============================================================================

# Convert peak day to month
heatwave_data$peak_month <- ifelse(is.na(heatwave_data$peak_day), NA,
                                   ceiling(heatwave_data$peak_day / 30.44))

# Create month labels
month_labels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

seasonal_data <- heatwave_data %>%
  filter(!is.na(peak_month)) %>%
  mutate(peak_month_label = month_labels[peak_month])

p5 <- seasonal_data %>%
  count(peak_month_label) %>%
  plot_ly(x = ~peak_month_label, y = ~n, type = 'bar',
          marker = list(color = '#dc2626'),
          hovertemplate = '<b>Month:</b> %{x}<br><b>Count:</b> %{y}<extra></extra>') %>%
  layout(title = list(text = '<b>Peak Heatwave Timing by Month</b>', 
                      font = list(size = 16)),
         xaxis = list(title = 'Month'),
         yaxis = list(title = 'Number of Peak Events'))

# Display plot
p5

# ============================================================================
# TREND ANALYSIS
# ============================================================================

# Linear regression for trends
magnitude_trend <- lm(magnitude ~ year, data = heatwave_data)
days_trend <- lm(days ~ year, data = heatwave_data)

cat("=== TREND ANALYSIS ===\n")
cat("Magnitude trend: ", round(coef(magnitude_trend)[2], 4), " per year\n")
cat("Days trend: ", round(coef(days_trend)[2], 4), " per year\n")
cat("Magnitude R-squared: ", round(summary(magnitude_trend)$r.squared, 3), "\n")
cat("Days R-squared: ", round(summary(days_trend)$r.squared, 3), "\n\n")

# ============================================================================
# DATA TABLE
# ============================================================================

# Create interactive data table
heatwave_table <- heatwave_data %>%
  select(year, magnitude, days, peak_day, has_heatwave) %>%
  mutate(
    magnitude = round(magnitude, 2),
    peak_day = ifelse(is.na(peak_day), "-", as.character(peak_day)),
    has_heatwave = ifelse(has_heatwave, "Yes", "No")
  ) %>%
  rename(
    Year = year,
    `Magnitude Index` = magnitude,
    `Heatwave Days` = days,
    `Peak Day` = peak_day,
    `Has Heatwave` = has_heatwave
  )

# Display interactive table
datatable(heatwave_table, 
          options = list(pageLength = 15, scrollX = TRUE),
          caption = "Complete Heatwave Dataset (1990-2024)")

# ============================================================================
# SAVE PLOTS
# ============================================================================

# Save individual plots as HTML files
htmlwidgets::saveWidget(p1, "heatwave_magnitude_timeseries.html")
htmlwidgets::saveWidget(p2, "heatwave_days_barchart.html") 
htmlwidgets::saveWidget(p3, "magnitude_vs_days_scatter.html")
htmlwidgets::saveWidget(p4, "decade_comparison.html")
htmlwidgets::saveWidget(p5, "seasonal_timing.html")

cat("=== ANALYSIS COMPLETE ===\n")
cat("Interactive plots saved as HTML files in your working directory\n")
cat("Files created:\n")
cat("- heatwave_magnitude_timeseries.html\n")
cat("- heatwave_days_barchart.html\n") 
cat("- magnitude_vs_days_scatter.html\n")
cat("- decade_comparison.html\n")
cat("- seasonal_timing.html\n")
```

#### heatwaveR library
```{r}
library(ggplot2)
library(dplyr)

# Add a column to detect_data marking heatwave days
detect_data$heatwave <- FALSE

# Mark heatwave days using start index and duration
for (i in 1:nrow(heatwaves$hwmid)) {
  duration <- heatwaves$hwmid[i, 2]  # duration (column 2)
  start_idx <- heatwaves$hwmid[i, 3] # start index (column 3)
  
  if (duration > 0 && start_idx > 0) {
    indices <- start_idx:(start_idx + duration - 1)
    # Ensure indices stay within bounds
    indices <- indices[indices <= nrow(detect_data)]
    detect_data$heatwave[indices] <- TRUE
  }
}

# Plot
ggplot(detect_data, aes(x = date, y = Tmax)) +
  geom_line(color = "black") +
  geom_point(data = detect_data %>% filter(heatwave), 
             aes(x = date, y = Tmax), 
             color = "red", size = 0.5) +
  labs(title = "Daily Tmax with Heatwaves Highlighted",
       x = "Date", y = "Tmax (°C)") +
  theme_minimal()

```



