---
title: "Hamburg_Drought_Indices_Analysis"
author: "MANDAR PRIYA PHATAK"
date: "2025-07-18"
output:
   html_document: default
   pdf_document:
    latex_engine: xelatex
   word_document: default
  
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)

```



```{r}
rm(list = ls())
```



```{r}
library(DT)
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
library(lubridate)
library(fitdistrplus)
```
### Data upload from the folder
```{r}
data <- readRDS("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")

```
```{r}
monthly_tbl <- data |> 
  dplyr::mutate(year = year(date),
         month = month(date)) |> 
  group_by(year, month) |> 
  summarize(
    prec = sum(prec, na.rm = TRUE),
    Tavg = mean(Tavg, na.rm = TRUE),
    Tmin = mean(Tmin, na.rm = TRUE),
    Tmax = mean(Tmax, na.rm = TRUE),
    .groups = "drop"
  ) |> 
 dplyr:: mutate(date = make_date(year, month)) |> 
  dplyr::select(year, month, date, prec:Tmax)

monthly_tbl <- monthly_tbl |> drop_na()
```
## Beofore we compute RDI & SPEI Index, we want to confirm selection for method for computing PET
```{r}

## PET computation 
PET_hargreaves <- hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)

PET_thornthwaite <- thornthwaite(Tave = monthly_tbl$Tavg,lat = 53.6332)

PET_hargreaves |> as_tibble()
PET_thornthwaite |> as_tibble()

cbind(PET_hargreaves |> as_tibble() , PET_thornthwaite |> as_tibble())
```
### We select Hargreaves 
Key Observations from Your Data:

Thornthwaite shows many zero values - This is problematic as it indicates the method fails during winter months when temperatures are low
Hargreaves provides continuous values - More realistic for year-round analysis
Hargreaves generally shows higher PET values - This is typical for coastal areas

Recommendation for Hamburg Fühlsbuttel:
Use Hargreaves method for the following reasons:
Climate-Specific Advantages:

Coastal location: Hargreaves performs better in maritime climates like Hamburg
Temperature range: Better suited for the moderate temperature variations in northern Germany
Humidity considerations: While it doesn't directly use humidity, it's calibrated for regions with moderate humidity

Technical Advantages:

No zero values: Ensures continuous drought index calculation
Widely validated: Extensively tested in European climates
Robust performance: Less sensitive to extreme temperature variations

For Your Location Specifically:

Hamburg's maritime climate (Köppen Cfb) is well-suited for Hargreaves
The method works well for locations with moderate temperature ranges (like northern Germany)
Less prone to computational issues in winter 

## Computing PET
```{r}
monthly_tbl <- monthly_tbl |> 
  mutate(PET = hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)) |> 
  rename(PREC = prec)
```

### Reconaissance Drought Index
```{r}
calculate_monthly_rdi <- function(data, scale) {
  require(dplyr)
  require(zoo)
  
  data <- data %>%
    arrange(date) %>%
    filter(PET > 0, PREC >= 0) %>%
    mutate(
      PREC_sum = rollsum(PREC, k = scale, fill = NA, align = "right"),
      PET_sum = rollsum(PET, k = scale, fill = NA, align = "right"),
      alpha = ifelse(PET_sum > 0, PREC_sum / PET_sum, NA_real_)
    )
  
  valid_alphas <- data$alpha[!is.na(data$alpha) & is.finite(data$alpha) & data$alpha > 0]
  
  if (length(valid_alphas) > 30) {
    log_alphas <- log(valid_alphas)
    y_bar <- mean(log_alphas, na.rm = TRUE)
    sigma_y <- sd(log_alphas, na.rm = TRUE)
    
    data <- data %>%
      mutate(
        RDIst = ifelse(alpha > 0, (log(alpha) - y_bar) / sigma_y, NA_real_)
      )
  } else {
    data$RDIst <- NA_real_
  }
  
  # Only return date and RDIst
  return(data |>  dplyr::select(date, RDIst))
}

```


```{r}
calculate_rdist_for_scales <- function(data, scales = c(3,6,9,12,24,36,48)) {
  require(dplyr)
  
  result <- data |> dplyr:: select(date)  # Start with date column
  
  for (k in scales) {
    rdi_k <- calculate_monthly_rdi(data, scale = k) %>%
      rename(!!paste0("RDI_", k) := RDIst)
    
    result <- left_join(result, rdi_k, by = "date")
  }
  
  return(result)
}

```
```{r}
monthly_rdist_tbl <- calculate_rdist_for_scales(data = monthly_tbl)
monthly_rdist_tbl |> head()
```







```{r}
# Corrected SPEI calculation (remains the same)
monthly_tbl$BAL = monthly_tbl$PREC - monthly_tbl$PET
monthly_ts <- ts(monthly_tbl[,-c(1,2,3,5,6,7)],start = c(1936,01), end = c(2025,05), frequency = 12)

### SPEI Series
spei3 <- spei(monthly_ts[,"BAL"],3)
spei6 <- spei(monthly_ts[,"BAL"],6)
spei9 <- spei(monthly_ts[,"BAL"],9)
spei12 <- spei(monthly_ts[,"BAL"],12)
spei24 <- spei(monthly_ts[,"BAL"],24)
spei36 <- spei(monthly_ts[,"BAL"],36)
spei48 <- spei(monthly_ts[,"BAL"],48)
```


```{r}
date_seq <- seq.Date(from = as.Date("1936-01-01"), 
                     to = as.Date("2025-05-01"), 
                     by = "month")
spei3_tbl <- spei3$fitted |> as_tibble() |> 
  rename(spei_3 = x) 
spei6_tbl <- spei6$fitted |> as_tibble() |> 
  rename(spei_6 = x) 
spei9_tbl <- spei9$fitted |> as_tibble() |> 
  rename(spei_9 = x) 
spei12_tbl <- spei12$fitted |> as_tibble() |> 
  rename(spei_12 = x) 
spei24_tbl <- spei24$fitted |> as_tibble() |> 
  rename(spei_24 = x) 
spei36_tbl <- spei36$fitted |> as_tibble() |> 
  rename(spei_36 = x) 
spei48_tbl <- spei48$fitted |> as_tibble() |> 
  rename(spei_48 = x)

spei3 <- data.frame(date = date_seq,Value = spei3_tbl)
spei6 <- data.frame(date = date_seq,Value = spei6_tbl)
spei9 <- data.frame(date = date_seq,Value = spei9_tbl)
spei12 <- data.frame(date = date_seq,Value = spei12_tbl)
spei24 <- data.frame(date = date_seq,Value = spei24_tbl)
spei36 <- data.frame(date = date_seq,Value = spei36_tbl)
spei48 <- data.frame(date = date_seq,Value = spei48_tbl)

```
```{r}
spei_df <- spei3 |> 
  left_join(spei6, by ="date") |> 
  left_join(spei9, by ="date") |> 
  left_join(spei12, by ="date") |> 
  left_join(spei24, by ="date") |> 
  left_join(spei36, by ="date") |> 
  left_join(spei48, by ="date")

```
```{r}
library(tidyverse)
combined_drought_tbl <- monthly_rdist_tbl |> 
  left_join(spei_df, by ="date")
```


```{r}
plot_tbl <- combined_drought_tbl %>%
  pivot_longer(cols = -date,
               names_to = "Index_Scale",
               values_to = "Value") |> 
mutate(
    Index = case_when(
      grepl("RDI", Index_Scale) ~ "RDIst",
      grepl("spei", Index_Scale) ~ "SPEI"
    ),
    Scale = gsub("[^0-9]", "", Index_Scale) |> as.numeric()
  )
```

```{r fig.height=10, fig.width=15}
ggplot(plot_tbl, aes(x = date, y = Value, color = Index)) +
  geom_line() +
  facet_wrap(~ Scale, scales = "free_y", ncol = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  labs(title = "Comparison of RDIst and SPEI over Time",
       x = "Year", y = "Index Value") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("SPEI" = "#1b9e77", "RDIst" = "#d95f02"))
```
```{r}
ggplot(plot_tbl %>% filter(Scale == 6), aes(x = date, y = Value, color = Index)) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray30") +
  labs(title = "6-Month Drought Indices: RDIst vs SPEI",
       x = "Year", y = "Index Value") +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("SPEI" = "#1b9e77", "RDIst" = "#d95f02"))
```




