---
title: "Hamburg_Drought_Indices_Analysis"
author: "MANDAR PRIYA PHATAK"
date: "2025-07-18"
output:
   html_document: default
   pdf_document:
    latex_engine: xelatex
   word_document: default
  
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)

```



```{r}
rm(list = ls())
```



```{r}
library(DT)
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
library(lubridate)
library(fitdistrplus)
```
### Data upload from the folder
```{r}
data <- readRDS("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")

```
```{r}
monthly_tbl <- data |> 
  dplyr::mutate(year = year(date),
         month = month(date)) |> 
  group_by(year, month) |> 
  summarize(
    prec = sum(prec, na.rm = TRUE),
    Tavg = mean(Tavg, na.rm = TRUE),
    Tmin = mean(Tmin, na.rm = TRUE),
    Tmax = mean(Tmax, na.rm = TRUE),
    .groups = "drop"
  ) |> 
 dplyr:: mutate(date = make_date(year, month)) |> 
  dplyr::select(year, month, date, prec:Tmax)

monthly_tbl <- monthly_tbl |> drop_na()
```
## Beofore we compute RDI & SPEI Index, we want to confirm selection for method for computing PET
```{r}

## PET computation 
PET_hargreaves <- hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)

PET_thornthwaite <- thornthwaite(Tave = monthly_tbl$Tavg,lat = 53.6332)

PET_hargreaves |> as_tibble()
PET_thornthwaite |> as_tibble()

cbind(PET_hargreaves |> as_tibble() , PET_thornthwaite |> as_tibble())
```
### We select Hargreaves 
Key Observations from Your Data:

Thornthwaite shows many zero values - This is problematic as it indicates the method fails during winter months when temperatures are low
Hargreaves provides continuous values - More realistic for year-round analysis
Hargreaves generally shows higher PET values - This is typical for coastal areas

Recommendation for Hamburg Fühlsbuttel:
Use Hargreaves method for the following reasons:
Climate-Specific Advantages:

Coastal location: Hargreaves performs better in maritime climates like Hamburg
Temperature range: Better suited for the moderate temperature variations in northern Germany
Humidity considerations: While it doesn't directly use humidity, it's calibrated for regions with moderate humidity

Technical Advantages:

No zero values: Ensures continuous drought index calculation
Widely validated: Extensively tested in European climates
Robust performance: Less sensitive to extreme temperature variations

For Your Location Specifically:

Hamburg's maritime climate (Köppen Cfb) is well-suited for Hargreaves
The method works well for locations with moderate temperature ranges (like northern Germany)
Less prone to computational issues in winter 

## Computing PET
```{r}
monthly_tbl <- monthly_tbl |> 
  mutate(PET = hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)) |> 
  rename(PREC = prec)
```

### Reconaissance Drought Index
```{r}
calculate_monthly_rdi <- function(data, scale) {
  require(dplyr)
  require(zoo)
  
  data <- data %>%
    arrange(date) %>%
    filter(PET > 0, PREC >= 0) %>%
    mutate(
      PREC_sum = rollsum(PREC, k = scale, fill = NA, align = "right"),
      PET_sum = rollsum(PET, k = scale, fill = NA, align = "right"),
      alpha = ifelse(PET_sum > 0, PREC_sum / PET_sum, NA_real_)
    )
  
  valid_alphas <- data$alpha[!is.na(data$alpha) & is.finite(data$alpha) & data$alpha > 0]
  
  if (length(valid_alphas) > 30) {
    log_alphas <- log(valid_alphas)
    y_bar <- mean(log_alphas, na.rm = TRUE)
    sigma_y <- sd(log_alphas, na.rm = TRUE)
    
    data <- data %>%
      mutate(
        RDIst = ifelse(alpha > 0, (log(alpha) - y_bar) / sigma_y, NA_real_)
      )
  } else {
    data$RDIst <- NA_real_
  }
  
  # Only return date and RDIst
  return(data |>  dplyr::select(date, RDIst))
}

```


```{r}
calculate_rdist_for_scales <- function(data, scales = c(3,6,9,12,24,36,48)) {
  require(dplyr)
  
  result <- data |> dplyr:: select(date)  # Start with date column
  
  for (k in scales) {
    rdi_k <- calculate_monthly_rdi(data, scale = k) %>%
      rename(!!paste0("RDI_", k) := RDIst)
    
    result <- left_join(result, rdi_k, by = "date")
  }
  
  return(result)
}

```
```{r}
monthly_rdist_tbl <- calculate_rdist_for_scales(data = monthly_tbl)
monthly_rdist_tbl |> head()
```
```{r fig.height=10,fig.width=15}
rdi_24tbl <- monthly_rdist_tbl |> 
  select(date,RDI_24) |> 
  drop_na() 

rdi_24tbl <- rdi_24tbl |> 
  mutate(sign = ifelse(RDI_24 >0,"pos","neg")) |> filter(date>"1989-12-01")

ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = sign)) +
  geom_col(show.legend = FALSE, width = 20) +  # Increased width for monthly data
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_x_date(
    date_breaks = "10 years",
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-12-31")),  # Extended limits
    expand = expansion(mult = c(0.01, 0.01))  # Small padding
  ) + 
  scale_y_continuous(
    breaks = seq(-2.5, 3, 0.5),  # Adjust based on your RDI range
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  scale_fill_manual(values = c("pos" = "#2f5952", "neg" = "#d95f02")) +
  labs(title = "24-Month Reconnaissance Drought Index (RDI)",
       subtitle = "Hamburg Fuhlsbüttel (1936-2025)",
       caption = "Source: DWD",
       x = "", 
       y = "") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
  
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text( face = "bold",size = 22, margin = margin(t=10,b=15)),
    plot.subtitle = element_text(color = "grey60",size = 18),
    axis.text.x = element_text(angle = 45, hjust = 1,size = 18),
    axis.text.y = element_text(hjust = 1,size = 18)
  )
```

```{r fig.height=10,fig.width=15}
# Data preparation
rdi_24tbl <- monthly_rdist_tbl |> 
  select(date, RDI_24) |> 
  drop_na() |> 
  filter(date > "1989-12-01") |> 
  mutate(sign = ifelse(RDI_24 > 0, "pos", "neg"))

# Plot
ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = sign)) +
  geom_col(show.legend = FALSE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  # Add horizontal reference lines for drought categories (optional)
  geom_hline(yintercept = c(-2, -1.5, -1, 1, 1.5, 2), 
             color = "grey70", linewidth = 0.3, linetype = "dashed", alpha = 0.7) +
  scale_x_date(
    date_breaks = "5 years",  # Changed to 5 years for better resolution
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-12-31")),
    expand = expansion(mult = c(0.01, 0.01))
  ) + 
  scale_y_continuous(
    breaks = seq(-2.5, 3, 0.5),
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  scale_fill_manual(values = c("pos" = "#2f5952", "neg" = "#d95f02")) +
  labs(
    title = "24-Month Reconnaissance Drought Index (RDI)",
    subtitle = "Hamburg Fuhlsbüttel (1990-2025)",
    caption = "Source: DWD | Negative values indicate drought conditions",
    x = "", 
    y = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 5)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
    plot.caption = element_text(color = "grey60", size = 12, hjust = 0),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )
```
```{r fig.align='center', fig.height=10, fig.width=15}
library(ggplot2)
library(dplyr)
library(scales)

# Enhanced data preparation with drought categories
rdi_24tbl <- monthly_rdist_tbl |> 
  select(date, RDI_24) |> 
  drop_na() |> 
  filter(date > "1989-12-01") |> 
  mutate(
    # Original sign classification
    sign = ifelse(RDI_24 > 0, "pos", "neg"),
    
    # Enhanced drought classification
    drought_category = case_when(
      RDI_24 >= 2.0 ~ "Extremely Wet",
      RDI_24 >= 1.5 ~ "Very Wet", 
      RDI_24 >= 1.0 ~ "Moderately Wet",
      RDI_24 >= 0.5 ~ "Slightly Wet",
      RDI_24 >= -0.5 ~ "Normal",
      RDI_24 >= -1.0 ~ "Mild Drought",
      RDI_24 >= -1.5 ~ "Moderate Drought", 
      RDI_24 >= -2.0 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Enhanced plot
p1 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  
  # Add reference lines with labels
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  
  # Annotations for key thresholds (optional)
  annotate("text", x = as.Date("2023-01-01"), y = 2.1, label = "Extremely Wet", 
           size = 3, color = "grey40", hjust = 1) +
  annotate("text", x = as.Date("2023-01-01"), y = -2.1, label = "Extreme Drought", 
           size = 3, color = "grey40", hjust = 1) +
  
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-12-31")),
    expand = expansion(mult = c(0.01, 0.01))
  ) + 
  
  scale_y_continuous(
    breaks = seq(-2.5, 3, 0.5),
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE  # Show wet conditions on left, dry on right
    )
  ) +
  
  labs(
    title = "24-Month Reconnaissance Drought Index (RDI)",
    subtitle = "Hamburg Fuhlsbüttel (1990-2025) - Drought Classification",
    caption = "Source: DWD | Negative values indicate drought conditions",
    x = "", 
    y = "RDI Value"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 10)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
    plot.caption = element_text(color = "grey60", size = 12, hjust = 1),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    legend.position = "bottom",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.8, "cm"),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )

# Display the plot
print(p1)
```


```{r fig.align='center', fig.dpi=450, fig.height=10, fig.width=15}
# Optional: Create summary statistics
drought_summary <- rdi_24tbl |>
  count(drought_category, name = "months") |>
  mutate(
    percentage = round(months / sum(months) * 100, 1),
    total_months = sum(months)
  ) |>
  arrange(desc(months))

print("Drought Category Distribution:")
print(drought_summary)

# Alternative version with simplified color scheme (closer to your original)
p2 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24)) +
  geom_col(aes(fill = case_when(
    RDI_24 >= 1.5 ~ "Very Wet",
    RDI_24 >= 0.5 ~ "Wet", 
    RDI_24 >= -0.5 ~ "Normal",
    RDI_24 >= -1.5 ~ "Drought",
    TRUE ~ "Severe Drought"
  )), show.legend = TRUE, width = 20) +
  
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  geom_hline(yintercept = c(-1.5, -0.5, 0.5, 1.5), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  
  scale_fill_manual(
    name = "Condition",
    values = c(
      "Severe Drought" = "#CC0000",
      "Drought" = "#FF6633", 
      "Normal" = "#CCCCCC",
      "Wet" = "#3399FF",
      "Very Wet" = "#0066CC"
    )
  ) +
  
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y", 
    limits = as.Date(c("1990-01-01", "2025-12-31")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  
  scale_y_continuous(
    breaks = seq(-2.5, 3, 0.5),
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  labs(
    title = "24-Month Reconnaissance Drought Index (RDI)",
    subtitle = "Hamburg Fuhlsbüttel (1990-2025) - Simplified Classification", 
    caption = "Source: DWD | Negative values indicate drought conditions",
    x = "",
    y = "RDI Value"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(), 
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 5)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
    plot.caption = element_text(color = "grey60", size = 12, hjust = 0),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    legend.position = "bottom", 
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )

# Display alternative plot
print(p2)
```
```{r fig.align='center',fig.height=10,fig.width=15}
library(ggplot2)
library(dplyr)
library(scales)

# Enhanced data preparation with drought categories
rdi_24tbl <- monthly_rdist_tbl |> 
  select(date, RDI_24) |> 
  drop_na() |> 
  filter(date > "1989-12-01") |> 
  mutate(
    # Original sign classification
    sign = ifelse(RDI_24 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_24 >= 2.0 ~ "Extremely Wet",
      RDI_24 >= 1.5 & RDI_24 < 2.0 ~ "Very Wet", 
      RDI_24 >= 1.0 & RDI_24 < 1.5 ~ "Moderately Wet",
      RDI_24 >= 0.5 & RDI_24 < 1.0 ~ "Slightly Wet",
      RDI_24 >= -0.5 & RDI_24 < 0.5 ~ "Normal",
      RDI_24 >= -1.0 & RDI_24 < -0.5 ~ "Mild Drought",
      RDI_24 >= -1.5 & RDI_24 < -1.0 ~ "Moderate Drought",
      RDI_24 >= -2.0 & RDI_24 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_24 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_24tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_24 < -0.5),
    severe_drought_months = sum(RDI_24 < -1.5),
    wet_months = sum(RDI_24 > 0.5),
    min_rdi = min(RDI_24),
    max_rdi = max(RDI_24),
    min_date = date[which.min(RDI_24)],
    max_date = date[which.max(RDI_24)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Enhanced plot with statistical insights
p1 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  
  # Add reference lines with labels
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  
  # Statistical annotation box
  annotate("rect", 
           xmin = as.Date("2000-01-01"), xmax = as.Date("2008-01-01"),
           ymin = 2.3, ymax = 3.2, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.3) +
  
  annotate("text", x = as.Date("2000-01-01"), y = 3.0, 
           label = paste0("KEY INSIGHTS (", min(year(rdi_24tbl$date)), "-", max(year(rdi_24tbl$date)), ")"),
           size = 3.5, fontface = "bold", hjust = 0) +
  
  annotate("text", x = as.Date("2000-01-01"), y = 2.8, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "% of period"),
           size = 3, hjust = 0, color = "#CC0000") +
  
  annotate("text", x = as.Date("2000-01-01"), y = 2.6, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "% of period"),
           size = 3, hjust = 0, color = "#8B0000") +
  
  annotate("text", x = as.Date("2000-01-01"), y = 2.4, 
           label = paste0("• Extreme event: ", format(drought_stats$min_date, "%b %Y"), " (RDI: ", round(drought_stats$min_rdi, 1), ")"),
           size = 3, hjust = 0, color = "grey30") +
  
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-12-31")),
    expand = expansion(mult = c(0.01, 0.01))
  ) + 
  
  scale_y_continuous(
    breaks = seq(-2.5, 3.5, 0.5),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE  # Show wet conditions on left, dry on right
    )
  ) +
  
  labs(
    title = "24-Month Reconnaissance Drought Index (RDI) - Hamburg Fuhlsbüttel",
    subtitle = stats_text,
    caption = "Source: DWD | Statistical summary shows percentage of time in each condition category",
    x = "", 
    y = "RDI Value"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 20, margin = margin(t = 10, b = 5)),
    plot.subtitle = element_text(color = "grey60", size = 12, margin = margin(b = 15)),
    plot.caption = element_text(color = "grey60", size = 10, hjust = 0),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key.size = unit(0.7, "cm"),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )

# Display the plot
print(p1)

# Create detailed summary table
drought_summary <- rdi_24tbl |>
  count(drought_category, name = "months") |>
  mutate(
    percentage = round(months / sum(months) * 100, 1),
    total_months = sum(months)
  ) |>
  arrange(match(drought_category, levels(rdi_24tbl$drought_category)))

# Print summary statistics
cat("\n=== DROUGHT ANALYSIS SUMMARY ===\n")
cat("Period:", min(rdi_24tbl$date), "to", max(rdi_24tbl$date), "\n")
cat("Total months analyzed:", nrow(rdi_24tbl), "\n\n")

cat("EXTREME EVENTS:\n")
cat("• Driest period:", format(drought_stats$min_date, "%B %Y"), 
    "with RDI =", round(drought_stats$min_rdi, 2), "\n")
cat("• Wettest period:", format(drought_stats$max_date, "%B %Y"), 
    "with RDI =", round(drought_stats$max_rdi, 2), "\n\n")

cat("CONDITION DISTRIBUTION:\n")
print(drought_summary)

cat("\nKEY INSIGHTS:\n")
cat("• Drought conditions (RDI < -0.5):", drought_stats$drought_pct, "% of the period\n")
cat("• Severe drought (RDI < -1.5):", drought_stats$severe_drought_pct, "% of the period\n")
cat("• Wet conditions (RDI > 0.5):", drought_stats$wet_pct, "% of the period\n")
```

```{r fig.align='center',fig.height=10, fig.width=15}
library(ggplot2)
library(dplyr)
library(scales)
library(lubridate)

# Plot
p1 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_text(aes(x = drought_stats$min_date, y = drought_stats$min_rdi - 0.3), 
            label = paste0("Driest: ", format(drought_stats$min_date, "%b %Y")), 
            angle = 90, hjust = 1.1, size = 3.3, color = "#8B0000") +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +
  geom_text(aes(x = drought_stats$max_date, y = drought_stats$max_rdi + 0.3), 
            label = paste0("Wettest: ", format(drought_stats$max_date, "%b %Y")), 
            angle = 90, hjust = -0.1, size = 3.3, color = "#00008B") +

  # KEY INSIGHTS annotation box
  annotate("rect", 
           xmin = as.Date("2004-05-01"), xmax = as.Date("2013-03-01"),
           ymin = 2.3, ymax = 3.2, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = as.Date("2005-01-01"), y = 3.0, 
           label = paste0("KEY INSIGHTS (", min(year(rdi_24tbl$date)), "-", max(year(rdi_24tbl$date)), ")"),
           size = 5, fontface = "bold", hjust = -0.1,) +

  annotate("text", x = as.Date("2004-05-01"), y = 2.8, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#CC0000") +

  annotate("text", x = as.Date("2004-05-01"), y = 2.6, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#8B0000",fontface="bold") +

  annotate("text", x = as.Date("2020-01-01"), y = -2.4, 
           label = paste0(" Event Extreme : ", format(drought_stats$min_date, "%b %Y"), 
                          " (RDI: ", round(drought_stats$min_rdi, 1), ") •"),
           size = 5, hjust = 1, color = "grey30",fontface="bold") +
annotate("text", x = drought_stats$max_date, y = drought_stats$max_rdi + 0.2, 
         label = paste0("• Wettest event:",format(drought_stats$max_date, "%b %Y"),
                        " (RDI: ",round(drought_stats$max_rdi, 1),")"),
         size = 3, hjust = 0, color = "grey30", fontface="bold") +
  # Start and end of analysis period
  annotate("text", x = min(rdi_24tbl$date), y = 3.4, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40",fontface="bold") +
  annotate("text", x = max(rdi_24tbl$date), y = 3.4, 
           label = "End of analysis", hjust = .85, size = 5, color = "grey50",fontface="bold") +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-05-28")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    breaks = seq(-2.5, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and theme
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1990–2024)",
    subtitle = stats_text,
    caption = "Source: DWD | Statistical summary shows percentage of time in each condition category",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t=10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15, colour = "grey10",face = "bold"),
    axis.text.y = element_text(size = 15, colour = "gray10",face="bold"),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key.size = unit(1, "cm"),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )

# Print the final plot
print(p1)

ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/RDI_24_Hamburg.png", plot = p1, width = 15, height = 10, dpi = 450)
```



```{r fig.align='center',fig.height=10,fig.width=15}
# Alternative version with simplified color scheme (closer to your original)
p2 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24)) +
  geom_col(aes(fill = case_when(
    RDI_24 >= 1.5 ~ "Very Wet",
    RDI_24 >= 0.5 ~ "Wet", 
    RDI_24 >= -0.5 ~ "Normal",
    RDI_24 >= -1.5 ~ "Drought",
    TRUE ~ "Severe Drought"
  )), show.legend = TRUE, width = 20) +
  
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  geom_hline(yintercept = c(-1.5, -0.5, 0.5, 1.5), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  
  scale_fill_manual(
    name = "Condition",
    values = c(
      "Severe Drought" = "#CC0000",
      "Drought" = "#FF6633", 
      "Normal" = "#CCCCCC",
      "Wet" = "#3399FF",
      "Very Wet" = "#0066CC"
    )
  ) +
  
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y", 
    limits = as.Date(c("1990-01-01", "2025-12-31")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  
  scale_y_continuous(
    breaks = seq(-2.5, 3, 0.5),
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  labs(
    title = "24-Month Reconnaissance Drought Index (RDI)",
    subtitle = "Hamburg Fuhlsbüttel (1990-2025) - Simplified Classification", 
    caption = "Source: DWD | Negative values indicate drought conditions",
    x = "",
    y = "RDI Value"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(), 
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 5)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
    plot.caption = element_text(color = "grey60", size = 12, hjust = 0),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    legend.position = "bottom", 
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )

# Display alternative plot
print(p2)

```


```{r fig.height=10,fig.width=15}
# # Alternative version with smoother date breaks for monthly data
# ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = sign)) +
#   geom_col(show.legend = FALSE, width = 25) +
#   geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
#   scale_x_date(
#     date_breaks = "2 years",  # Even finer resolution
#     date_labels = "'%y",      # Shortened year format
#     limits = as.Date(c("1990-01-01", "2025-12-31")),
#     expand = expansion(mult = c(0.01, 0.01))
#   ) + 
#   scale_y_continuous(
#     breaks = seq(-2.5, 3, 0.5),
#     expand = expansion(mult = c(0.05, 0.05))
#   ) +
#   scale_fill_manual(values = c("pos" = "#2f5952", "neg" = "#d95f02")) +
#   labs(
#     title = "24-Month Reconnaissance Drought Index (RDI)",
#     subtitle = "Hamburg Fuhlsbüttel (1990-2025)",
#     caption = "Source: DWD",
#     x = "", 
#     y = ""
#   ) +
#   theme_minimal(base_size = 12) +
#   theme(
#     plot.title.position = "plot",
#     panel.grid.minor = element_blank(),
#     panel.grid.major.x = element_line(color = "grey90", linewidth = 0.3),
#     panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
#     plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 5)),
#     plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
#     plot.caption = element_text(color = "grey60", size = 12, hjust = 0),
#     axis.text.x = element_text(size = 16),
#     axis.text.y = element_text(size = 18),
#     plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
#   )
```


```{r}
# For creating all three plots (24, 36, 48 months) with consistent styling:
create_rdi_plot <- function(data, rdi_col, title_months) {
  data |> 
    select(date, all_of(rdi_col)) |> 
    drop_na() |> 
    filter(date > "1989-12-01") |> 
    mutate(sign = ifelse(!!sym(rdi_col) > 0, "pos", "neg")) |> 
    ggplot(aes(x = date, y = !!sym(rdi_col), fill = sign)) +
    geom_col(show.legend = FALSE, width = 20) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    scale_x_date(
      date_breaks = "5 years",
      date_labels = "%Y",
      limits = as.Date(c("1990-01-01", "2025-12-31")),
      expand = expansion(mult = c(0.01, 0.01))
    ) + 
    scale_y_continuous(
      breaks = seq(-2.5, 3, 0.5),
      expand = expansion(mult = c(0.05, 0.05))
    ) +
    scale_fill_manual(values = c("pos" = "#2f5952", "neg" = "#d95f02")) +
    labs(
      title = paste0(title_months, "-Month Reconnaissance Drought Index (RDI)"),
      subtitle = "Hamburg Fuhlsbüttel (1990-2025)",
      caption = "Source: DWD",
      x = "", y = ""
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title.position = "plot",
      panel.grid.minor.x = element_blank(),
      panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
      panel.grid.minor.y = element_blank(),
      plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 5)),
      plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
      plot.caption = element_text(color = "grey60", size = 12, hjust = 0),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
      axis.text.y = element_text(size = 18),
      plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
    )
}

# Usage examples:
# plot_24 <- create_rdi_plot(monthly_rdist_tbl, "RDI_24", "24")
# plot_36 <- create_rdi_plot(monthly_rdist_tbl, "RDI_36", "36")  
# plot_48 <- create_rdi_plot(monthly_rdist_tbl, "RDI_48", "48")
```






```{r}
# Corrected SPEI calculation (remains the same)
monthly_tbl$BAL = monthly_tbl$PREC - monthly_tbl$PET
monthly_ts <- ts(monthly_tbl[,-c(1,2,3,5,6,7)],start = c(1936,01), end = c(2025,05), frequency = 12)

### SPEI Series
spei3 <- spei(monthly_ts[,"BAL"],3)
spei6 <- spei(monthly_ts[,"BAL"],6)
spei9 <- spei(monthly_ts[,"BAL"],9)
spei12 <- spei(monthly_ts[,"BAL"],12)
spei24 <- spei(monthly_ts[,"BAL"],24)
spei36 <- spei(monthly_ts[,"BAL"],36)
spei48 <- spei(monthly_ts[,"BAL"],48)
```


```{r}
date_seq <- seq.Date(from = as.Date("1936-01-01"), 
                     to = as.Date("2025-05-01"), 
                     by = "month")
spei3_tbl <- spei3$fitted |> as_tibble() |> 
  rename(spei_3 = x) 
spei6_tbl <- spei6$fitted |> as_tibble() |> 
  rename(spei_6 = x) 
spei9_tbl <- spei9$fitted |> as_tibble() |> 
  rename(spei_9 = x) 
spei12_tbl <- spei12$fitted |> as_tibble() |> 
  rename(spei_12 = x) 
spei24_tbl <- spei24$fitted |> as_tibble() |> 
  rename(spei_24 = x) 
spei36_tbl <- spei36$fitted |> as_tibble() |> 
  rename(spei_36 = x) 
spei48_tbl <- spei48$fitted |> as_tibble() |> 
  rename(spei_48 = x)

spei3 <- data.frame(date = date_seq,Value = spei3_tbl)
spei6 <- data.frame(date = date_seq,Value = spei6_tbl)
spei9 <- data.frame(date = date_seq,Value = spei9_tbl)
spei12 <- data.frame(date = date_seq,Value = spei12_tbl)
spei24 <- data.frame(date = date_seq,Value = spei24_tbl)
spei36 <- data.frame(date = date_seq,Value = spei36_tbl)
spei48 <- data.frame(date = date_seq,Value = spei48_tbl)

```









