---
title: "Temperature_Analysis"
author: "Mandar"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---


```{r}
rm(list = ls())
```
```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)
```



# Reading the data from the directory
```{r}

## Daily data
data <- readRDS(file = "/Users/mandarphatak/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")
data <- data |> dplyr::filter(date < "2025-01-01")

```
```{r}
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
```

# data tranformation :- We select the te



```{r echo=FALSE}
# conversion to monthly data ( makes it easier for analysis)
monthly_tbl <-data |> 
  mutate(year = year(date),
         month = month(date) ) |> 
  group_by(year, month) |> 
  summarize(
    Tavg = mean(Tavg, na.rm =TRUE),
    Tmax = mean(Tmax, na.rm =TRUE),
    Tmin = mean(Tmin, na.rm =TRUE),
    prec = sum(prec,na.rm =  TRUE),
    rel_hum = mean(rel_hum, na.rm = TRUE),
    
    .groups = "drop"
    
  ) |> drop_na()

# # conversion to annual data (for over all  time-period)
annual_tbl <- monthly_tbl |> 
  group_by(year) |> 
  summarize(
     Tavg = mean(Tavg, na.rm =TRUE),
    Tmax = mean(Tmax, na.rm =TRUE),
    Tmin = mean(Tmin, na.rm =TRUE),
    prec = sum(prec,na.rm =  TRUE),
    rel_hum = mean(rel_hum, na.rm = TRUE),
    .groups = "drop"
  )

annual_tbl |> dplyr:: select(-year) |> summary()
```
### Focusing on  Temperature data


## Data Over View :- We perform the following data manipulation for desired data overview 

```{r echo=FALSE}
## Only Temperature 
annual_temperature_tbl_long <- annual_tbl |> 
  pivot_longer(
    cols = c(Tavg,Tmax,Tmin),
    names_to = "Temperature_Type",
    values_to = "Temperature"
  ) |> 
  filter(year < 2025)
```


#### Temperature Variation 

```{r fig.width=15,fig.height=10}
annual_temperature_tbl_long$Temperature_Type <- factor(
  annual_temperature_tbl_long$Temperature_Type,
  levels = c("Tmax", "Tavg", "Tmin"))


trends_tbl <- annual_temperature_tbl_long %>%
  group_by(Temperature_Type) %>%
  summarize(trend = coef(lm(Temperature ~ year))[2]) %>%
  ungroup()


trends_tbl <- trends_tbl %>%
  mutate(
    label = case_when(
      Temperature_Type == "Tmax" ~ paste0("Maximum Temperature (Trend: +", round(trend, 4), "°C/year)"),
      Temperature_Type == "Tavg" ~ paste0("Average Temperature (Trend: +", round(trend, 4), "°C/year)"),
      Temperature_Type == "Tmin" ~ paste0("Minimum Temperature (Trend: +", round(trend, 4), "°C/year)")
    ),
    color = case_when(
      Temperature_Type == "Tmax" ~ "#ff1744",
      Temperature_Type == "Tavg" ~ "#fc8d59",
      Temperature_Type == "Tmin" ~ "#4575b4"
    ),
    y = case_when(  # adjust based on your plot’s Y-axis ranges
      Temperature_Type == "Tmax" ~ 15,
      Temperature_Type == "Tavg" ~ 10.55,
      Temperature_Type == "Tmin" ~ 6.5
    ),
    x = 1938  # horizontal position of the labels
  )

```

### Overview of Temperature Variation 1936-2024
```{r fig.width=15,fig.height=10}
ggplot(annual_temperature_tbl_long, aes(x = year, y = Temperature, colour = Temperature_Type)) + 
  geom_line(linewidth = 1) +
  geom_point(aes(fill = Temperature_Type), shape = 21, size = 4, stroke = 1, colour = "white") +
  # Add separate trend lines for each temperature type
  geom_smooth(aes(group = Temperature_Type), method = "lm", se = FALSE, linewidth = 1, 
              show.legend = FALSE, color = "black") +
  geom_text(
    data = trends_tbl,
    aes(x = x, y = y, label = label, color = Temperature_Type),
    fontface = "bold", hjust = 0, size = 7, show.legend = FALSE
  ) +
  
  # Fixed color scales with intuitive temperature colors
  scale_color_manual(
    values = c("Tmax" = "#ff1744",    # Hot red for maximum
               "Tavg" = "#fc8d59",    # Orange for average  
               "Tmin" = "#4575b4"),   # Cool blue for minimum
    labels = c("Tmax" = "Maximum Temperature",
               "Tavg" = "Average Temperature", 
               "Tmin" = "Minimum Temperature"),
    name = "Temperature Type"
  ) +
  scale_fill_manual(
    values = c("Tmax" = "#ff1744",    # Same colors for consistency
               "Tavg" = "#fc8d59",    
               "Tmin" = "#4575b4"),
    guide = "none"  # Hide fill legend since color legend shows the same info
  ) +
  
  scale_x_continuous(breaks = c(seq(1936, 2024, 10), 2024), 
                     expand = c(0.02, 0.02), 
                     limits = c(1936, 2024)) +
  scale_y_continuous(
    limits = c(1, 16),
    breaks = seq(1, 16, by = 2),
    labels = function(x) paste0(format(x, nsmall = 0), "°C"),
    expand = c(0, 0)
  ) +
  
  labs(
    title = "Temperature Variation (1936-2025)",
    subtitle = "Hamburg Fuhlsbüttel",
    caption = "Source:- DWD",
    x = "",
    y = "Temperature (°C)"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0, face = "bold", colour = "black", size = 30, margin = margin(b = 20)),
    plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black", size = 30, margin = margin(b = 10)),
    plot.caption = element_text(size = 18, face = "bold", color = "gray50", hjust = 1),  #
    axis.line = element_line(color = "black", linewidth = 1),
    axis.title = element_text(face = "bold", color = "black", size = 22),
    axis.ticks = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black", size = 20),
    axis.text.y = element_text(face = "bold", color = "black", size = 20),
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    legend.position = "bottom",  # Better positioning for temperature legend
    panel.background = element_blank(),
    plot.background = element_blank()
  ) +
   guides(color = "none", fill = "none")
  

```
### Decadal Changes in Temperature 

```{r fig.width=15,fig.height=10}

 annual_temperature_tbl_long <- annual_temperature_tbl_long |>
  mutate(decade_group = case_when(
    year >= 1936 & year <= 1939 ~ "1936–1939",
    year >= 1940 & year <= 1949 ~ "1940–1949",
    year >= 1950 & year <= 1959 ~ "1950–1959",
    year >= 1960 & year <= 1969 ~ "1960–1969",
    year >= 1970 & year <= 1979 ~ "1970–1979",
    year >= 1980 & year <= 1989 ~ "1980–1989",
    year >= 1990 & year <= 1999 ~ "1990–1999",
    year >= 2000 & year <= 2009 ~ "2000–2009",
    year >= 2010 & year <= 2019 ~ "2010–2019",
    year >= 2020 & year <= 2024 ~ "2020–2024",
    TRUE ~ NA_character_
  ))

  decadal_summary_tbl <- annual_temperature_tbl_long %>%
  group_by(decade_group, Temperature_Type) %>%
  summarise(decadal_mean = mean(Temperature, na.rm = TRUE), .groups = "drop")
  
library(ggplot2)

ggplot(decadal_summary_tbl, aes(x = decade_group, y = decadal_mean, fill = Temperature_Type)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  scale_fill_manual(
    values = c("Tmax" = "#ff1744", "Tavg" = "#fc8d59", "Tmin" = "#4575b4"),
    name = "Temperature Type",
    labels = c("Tmax" = "Maximum", "Tavg" = "Average", "Tmin" = "Minimum")
  ) +
  labs(
    title = "Decadal Mean Temperature (1936–2024)",
    subtitle = "Hamburg Fuhlsbüttel",
    x = "",
    y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold",size = 15),
    axis.text.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold")
  )

```
## Variation in Temperature across decade
```{r fig.width=15,fig.height=10}


ggplot(decadal_summary_tbl, aes(x = decade_group, y = decadal_mean, group = Temperature_Type, color = Temperature_Type)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(
    values = c("Tmax" = "#ff1744", "Tavg" = "#fc8d59", "Tmin" = "#4575b4"),
    name = "Temperature Type",
    labels = c("Tmax" = "Maximum", "Tavg" = "Average", "Tmin" = "Minimum")
  ) +
  labs(
    title = "Decadal Temperature Trends (1936–2024)",
    subtitle = "Hamburg Fuhlsbüttel",
    x = "",
    y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 20),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black", size = 20),
    axis.text.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.position = "none"
    
  ) +
  
# Add custom legend inside the plot
annotate("point", x = 2.5, y = 13.5, size = 5, shape = 21, fill = "#ff1744") +
annotate("text", x = 2.6, y = 13.5, label = "Maximum Temperature", hjust = 0, size = 6, fontface = "bold") +

annotate("point", x = 2.5, y = 9.5, size = 5, shape = 21, fill = "#fc8d59") +
annotate("text", x = 2.6, y = 9.5, label = "Average Temperature", hjust = 0, size = 6, fontface = "bold") +

annotate("point", x = 2.5, y = 6, size = 5, shape = 21, fill = "#4575b4") +
annotate("text", x = 2.6, y = 6, label = "Minimum Temperature", hjust = 0, size = 6, fontface = "bold")


```



## Incremental  changes in Temperature across Decades
```{r fig.width=15,fig.height=10}
decadal_diff_tbl <- decadal_summary_tbl %>%
  arrange(Temperature_Type, decade_group) %>%
  group_by(Temperature_Type) %>%
  mutate(decadal_diff = decadal_mean - lag(decadal_mean)) %>%
  filter(!is.na(decadal_diff)) %>%
  ungroup()

ggplot(decadal_diff_tbl, aes(x = decade_group, y = decadal_diff, fill = Temperature_Type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_label(aes(label = round(decadal_diff, 2)),
           position = position_dodge(width = 0.7),
           size = 4.5, fontface = "bold",
           show.legend = FALSE) +
  # geom_text(aes(label = round(decadal_diff, 2)), 
  #           position = position_dodge(width = 0.7), 
  #           vjust = ifelse(decadal_diff_tbl$decadal_diff >= 0, -0.3, 1.3), 
  #           size = 5, fontface = "bold") +
  scale_fill_manual(
    values = c("Tmax" = "#ff1744", "Tavg" = "#fc8d59", "Tmin" = "#4575b4"),
    name = "Temperature Type",
    labels = c("Tmax" = "Maximum", "Tavg" = "Average", "Tmin" = "Minimum")
  ) +
  labs(
  title = "Decadal Changes in Temperature at Hamburg-Fuhlsbüttel",
  subtitle = "Average, Minimum, and Maximum Temperatures – Change from Previous Decade (°C)"
) +

  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 20),
    axis.text.x = element_text(angle = 30, hjust = 1),
    axis.text.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold")
  )
```
```{r fig.width=15,fig.height=10}
ggplot(decadal_diff_tbl, aes(x = decade_group, y = decadal_diff, fill = Temperature_Type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.5) +
  geom_label(aes(label = round(decadal_diff, 2)),
             position = position_dodge(width = 0.8),
             size = 4.5, fontface = "bold",
             show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1, linetype = "dashed") +

  # Manual fill colors
  scale_fill_manual(
    values = c("Tmax" = "#ff1744", "Tavg" = "#fc8d59", "Tmin" = "#4575b4")
  ) +
 scale_y_continuous(
  limits = c(-1, 2),
  breaks = seq(-2, 2, by = 0.25),
  labels = function(x) sprintf("%.2f", x)
) +
  # Titles
  labs(
    title = "Decadal Changes in Temperature at Hamburg-Fuhlsbüttel",
    subtitle = "Average, Minimum, and Maximum Temperatures – Change from Previous Decade (°C)",
    x = "", y = "Temperature Change (°C)"
  ) +

  # Theme & custom key
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 20),
    axis.text.x = element_text(angle = 30, hjust = 1, size = 16, face = "bold"),
    axis.text.y = element_text(face = "bold", size = 14),
    legend.position = "none"
  ) +

  # Custom legend inside plot

  # Custom legend inside plot for y-axis from -1 to 2
annotate("rect", xmin = 7.5, xmax = 9.5, ymin = 1.2, ymax = 2, fill = "white", alpha = 0.7) +  # background box
annotate("rect", xmin = 7.5, xmax = 9.5, ymin = 1.1, ymax = 2, colour = "black", fill = NA) +  # border

annotate("rect", xmin = 7.6, xmax = 7.8, ymin = 1.55, ymax = 1.75, fill = "#ff1744") +
annotate("text", x = 7.85, y = 1.6, label = "Maximum", hjust = 0, size = 5, fontface = "bold") +

annotate("rect", xmin = 7.6, xmax = 7.8, ymin = 1.35, ymax = 1.55, fill = "#fc8d59") +
annotate("text", x = 7.85, y = 1.4, label = "Average", hjust = 0, size = 5, fontface = "bold") +

annotate("rect", xmin = 7.6, xmax = 7.8, ymin = 1.15, ymax = 1.35, fill = "#4575b4") +
annotate("text", x = 7.85, y = 1.2, label = "Minimum", hjust = 0, size = 5, fontface = "bold")

```



#### Heatwave Events  :- We visualize the intensity and duration of heat events from 1936-2024.
```{r}
library(ggplot2)
library(dplyr)
library(scales)
library(ggrepel)  # For repelling labels






identify_heatwaves <- function(date, tmax, threshold, min_duration = 3) {
  # Ensure inputs are valid
  stopifnot(length(date) == length(tmax))
  stopifnot(is.numeric(tmax))
  stopifnot(inherits(date, "Date"))
  
  # Sort and prepare
  data <- data.frame(date = date, tmax = tmax)
  data <- data[order(data$date), ]
  data$above_threshold <- data$tmax > threshold
  
  # Initialize output
  heatwaves <- data.frame(
    start_date = as.Date(character()),
    end_date = as.Date(character()),
    duration = integer(),
    avg_intensity = numeric(),
    total_intensity = numeric(),
    max_intensity = numeric(),
    mean_tmax = numeric(),
    peak_tmax = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Tracking flags
  in_heatwave <- FALSE
  heatwave_start <- NULL
  heatwave_length <- 0
  
  # Loop through
  for (i in seq_len(nrow(data))) {
    if (data$above_threshold[i]) {
      if (!in_heatwave) {
        in_heatwave <- TRUE
        heatwave_start <- data$date[i]
        heatwave_length <- 1
      } else {
        heatwave_length <- heatwave_length + 1
      }
    } else {
      if (in_heatwave && heatwave_length >= min_duration) {
        end_date <- data$date[i - 1]
        heatwave_days <- which(data$date >= heatwave_start & data$date <= end_date)
        exceedance <- data$tmax[heatwave_days] - threshold
        
        heatwaves <- rbind(heatwaves, data.frame(
          start_date = heatwave_start,
          end_date = end_date,
          duration = heatwave_length,
          avg_intensity = mean(exceedance),
          total_intensity = sum(exceedance),
          max_intensity = max(exceedance),
          mean_tmax = mean(data$tmax[heatwave_days]),
          peak_tmax = max(data$tmax[heatwave_days])
        ))
      }
      in_heatwave <- FALSE
      heatwave_length <- 0
    }
  }
  
  # Final check at end
  if (in_heatwave && heatwave_length >= min_duration) {
    end_date <- data$date[nrow(data)]
    heatwave_days <- which(data$date >= heatwave_start & data$date <= end_date)
    exceedance <- data$tmax[heatwave_days] - threshold
    
    heatwaves <- rbind(heatwaves, data.frame(
      start_date = heatwave_start,
      end_date = end_date,
      duration = heatwave_length,
      avg_intensity = mean(exceedance),
      total_intensity = sum(exceedance),
      max_intensity = max(exceedance),
      mean_tmax = mean(data$tmax[heatwave_days]),
      peak_tmax = max(data$tmax[heatwave_days])
    ))
  }
  
  # Add year column
  if (nrow(heatwaves) > 0) {
    heatwaves$year <- as.numeric(format(heatwaves$start_date, "%Y"))
  }
  
  return(heatwaves)
}


heatwave_events_tbl <- identify_heatwaves(
  date = data$date,
  tmax = data$Tmax,
  threshold = 28,
  min_duration = 3
)

heatwave_count_per_year <- heatwave_events_tbl %>%
  count(year, name = "num_events")

heatwave_summary_tbl <- heatwave_events_tbl %>%
  group_by(year) %>%
  summarise(
    num_events = n(),
    total_duration = sum(duration, na.rm = TRUE),
    avg_duration = mean(duration, na.rm = TRUE),
    mean_avg_intensity = mean(avg_intensity, na.rm = TRUE),
    mean_max_intensity = mean(max_intensity, na.rm = TRUE),
    total_intensity = sum(total_intensity, na.rm = TRUE),
    max_peak_tmax = max(peak_tmax, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r}
# Convert date columns
 heatwave_events_tbl$start_date <- as.Date(heatwave_events_tbl$start_date)
heatwave_events_tbl$end_date <- as.Date(heatwave_events_tbl$end_date)

heatwave_events_tbl$year <- as.numeric(heatwave_events_tbl$year)
# 
# Custom color palette
custom_colors <- c("#f8f800", "#fdc70c", "#f3903f", "#ff4d00", "#e60000", "#5e0000")

# Identify recent events since 1990 with avg_intensity > 5
recent_events <- heatwave_events_tbl %>% 
  filter(year > 1990, avg_intensity > 5) %>%
  summarize(count = n()) %>%
  pull(count)

# Total events and percentage
total_events <- nrow(heatwave_events_tbl)
percent_recent <- round((recent_events / total_events) * 100)

# 95th percentile extreme threshold (for avg_intensity)
threshold_95 <- quantile(heatwave_events_tbl$avg_intensity, 0.95, na.rm = TRUE)

# Get max intensity (avg_intensity) and longest duration
max_intensity_value <- max(heatwave_events_tbl$avg_intensity, na.rm = TRUE)
longest_duration_value <- max(heatwave_events_tbl$duration, na.rm = TRUE)

# Label max avg_intensity event
max_intensity_event <- heatwave_events_tbl %>%
  filter(avg_intensity == max_intensity_value) %>%
  mutate(label = paste0("Max intensity: ", round(avg_intensity, 1), "\n", format(start_date, "%b %d, %Y")))

# Label longest duration event
longest_duration_event <- heatwave_events_tbl %>%
  filter(duration == longest_duration_value) %>%
  mutate(label = paste0("Longest: ", duration, " days\n", format(start_date, "%b %Y")))

# Filter extreme events based on avg_intensity, exclude max to avoid duplicate label
extreme_events <- heatwave_events_tbl %>%
  filter(avg_intensity > threshold_95, avg_intensity < max_intensity_value) %>%
  mutate(label = paste0("Extreme: ", round(avg_intensity, 1), "\n", format(start_date, "%Y")))

```


```{r fig.width=15,fig.height=10}
p <- ggplot(heatwave_events_tbl, aes(x = as.numeric(year), y = avg_intensity)) +
  
  geom_point(aes(size = duration, color = avg_intensity), alpha = 0.85) +
  
  geom_vline(xintercept = 1990, linetype = "dashed", color = "blue", linewidth = 0.6) +
  annotate("text", x = 1990, y = 6.25, label = "1990", 
           color = "blue", fontface = "bold", size = 3) +
  
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = "darkblue", linetype = "solid", linewidth = 0.8, alpha = 0.6) +
  
  geom_text(data = max_intensity_event, aes(label = label),
            hjust = 1, vjust = 1.2, size = 3, fontface = "bold", color = "#5e0000") +
  
  geom_hline(yintercept = threshold_95, linetype = "dotted", color = "darkred") +
  annotate("text", x = 1940, y = threshold_95, label = "95th Percentile", 
           color = "darkred", size = 3) +
  
  # Annotate longest duration using repel to avoid overlap
  ggrepel::geom_text_repel(data = longest_duration_event, aes(label = label),
                           size = 3, fontface = "bold", color = "#5e0000",
                           max.overlaps = 5, box.padding = 0.4, point.padding = 0.4,
                           hjust = 1) +
  
  # Annotate extreme events
  ggrepel::geom_text_repel(data = extreme_events, aes(label = label),
                           size = 2.8, fontface = "italic", color = "black",
                           max.overlaps = 10, box.padding = 0.4, point.padding = 0.3) +
  
  # Axis scales
  scale_y_continuous(limits = c(0, 6.25), breaks = seq(0, 6.25, 0.5),
                     expand = c(0, 0.05)) +
  
  scale_x_continuous(limits = c(1936, 2024),
                     breaks = c(seq(1936, 2024, by = 10), 2024)) +
  
  scale_color_gradientn(colors = custom_colors,
                        name = "Avg. Intensity",
                        guide = guide_colorbar(barwidth = 1, barheight = 10,
                                               title.position = "top", title.hjust = 0.5)) +
  
  scale_size_continuous(range = c(2, 10),
                        name = "Duration (days)",
                        breaks = c(3, 6, 9, 12),
                        guide = guide_legend(title.position = "top",
                                             title.hjust = 0.5,
                                             override.aes = list(color = "darkblue"))) +
  
  # Labels
  labs(
    title = "Heatwave Events in Hamburg Fuhlsbüttel (1936–2024)",
    subtitle = paste0("Based on threshold of 28°C for minimum 3 consecutive days\n",
                      percent_recent, "% of all events occurred since 1990"),
    x = "",
    y = "Intensity (°C above threshold)",
    caption = "Source: Deutscher Wetterdienst (DWD)"
  ) +
  
  # Theme
  theme_minimal(base_size = 12) +
  theme_minimal(base_size = 12) +
theme(
  plot.title.position = "plot",
  plot.title = element_text(face = "bold", size = 20, margin = margin(b = 15)),
  plot.subtitle = element_text(size = 15, color = "gray30", margin = margin(b = 15), face = "bold"),
  axis.title.y = element_text(margin = margin(r = 10), face = "bold", size = 12),
  axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold", color = "gray30", size = 12),
  axis.text.y  = element_text(face = "bold", color = "gray30", size = 12),
  legend.position = "right",  # ✅ Only once
  legend.box = "vertical",
  legend.margin = margin(l = 10),
  legend.title = element_text(face = "bold"),
  panel.grid.minor = element_blank(),
  panel.grid.major.x = element_blank(),
  panel.grid.major.y = element_line(color = "gray80", linetype = "dotted")
)


p
ggsave("/Users/mandarphatak/Climate-Analysis/heat_wave_event_plot/HamburgFuhlsbüttel_Heatwave_Events.png", plot = p, width = 15, height = 10, dpi = 360, bg = "white")

```
### Extreme Heatwave using Maximum Intensity 
```{r}
# ===== MAX INTENSITY ANALYSIS =====
recent_events_max <- heatwave_events_tbl %>%
  filter(year > 1990, max_intensity > 5) %>%
  summarize(count = n()) %>%
  pull(count)

percent_recent_max <- round((recent_events_max / total_events) * 100)

threshold_95_max <- quantile(heatwave_events_tbl$max_intensity, 0.95, na.rm = TRUE)

max_max_intensity <- max(heatwave_events_tbl$max_intensity, na.rm = TRUE)

max_max_event <- heatwave_events_tbl %>%
  filter(max_intensity == max_max_intensity) %>%
  mutate(label = paste0("Max exceedance: ", round(max_intensity, 1), "\n", format(start_date, "%b %d, %Y")))

extreme_events_max <- heatwave_events_tbl %>%
  filter(max_intensity > threshold_95_max, max_intensity < max_max_intensity) %>%
  mutate(label = paste0("Extreme: ", round(max_intensity, 1), "\n", format(start_date, "%Y")))

```
```{r fig.width= 15, fig.height= 10}
p_max <- ggplot(heatwave_events_tbl, aes(x = as.numeric(year), y = max_intensity)) +
  
  geom_point(aes(size = duration, color = max_intensity), alpha = 0.85) +
  
  geom_vline(xintercept = 1990, linetype = "dashed", color = "blue", linewidth = 0.6) +
  annotate("text", x = 1990, y = 12, label = "1990", 
           color = "blue", fontface = "bold", size = 3) +
  
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = "darkblue", linetype = "solid", linewidth = 0.8, alpha = 0.6) +
  
  geom_text(data = max_max_event, aes(label = label),
            hjust = 1, vjust = 1.2, size = 3, fontface = "bold", color = "#5e0000") +
  
  geom_hline(yintercept = threshold_95_max, linetype = "dotted", color = "darkred") +
  annotate("text", x = 1940, y = threshold_95_max, label = "95th Percentile", 
           color = "darkred", size = 3) +
  
  ggrepel::geom_text_repel(data = extreme_events_max, aes(label = label),
                           size = 2.8, fontface = "italic", color = "black",
                           max.overlaps = 10, box.padding = 0.4, point.padding = 0.3) +

  scale_y_continuous(limits = c(0, 12), breaks = seq(0, 12, 0.5),
                     expand = c(0, 0.05)) +
  
  scale_x_continuous(limits = c(1936, 2024),
                     breaks = c(seq(1936, 2024, by = 10), 2024)) +
  
  scale_color_gradientn(colors = custom_colors,
                        name = "Max Intensity",
                        guide = guide_colorbar(barwidth = 1, barheight = 10,
                                               title.position = "top", title.hjust = 0.5)) +
  
  scale_size_continuous(range = c(2, 10),
                        name = "Duration (days)",
                        breaks = c(3, 6, 9, 12),
                        guide = guide_legend(title.position = "top",
                                             title.hjust = 0.5,
                                             override.aes = list(color = "darkblue"))) +
  
  labs(
    title = "Max Daily Exceedance in Hamburg Fuhlsbüttel Heatwave Events (1936–2024)",
    subtitle = paste0("Based on max daily exceedance above 28°C threshold\n",
                      percent_recent_max, "% of all events occurred since 1990"),
    x = "",
    y = "Max Intensity (°C above threshold)",
    caption = "Source: Deutscher Wetterdienst (DWD)"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", size = 22, margin = margin(b = 15)),
    plot.subtitle = element_text(size = 15, color = "gray30", margin = margin(b = 15), face = "bold"),
    axis.title.y = element_text(margin = margin(r = 10), face = "bold", size = 16),
    axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold", color = "gray30", size = 14),
    axis.text.y  = element_text(face = "bold", color = "gray30", size = 14),
    legend.position = "right",
    legend.box = "vertical",
    legend.margin = margin(l = 10),
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray80", linetype = "dotted")
  )

p_max

ggsave("/Users/mandarphatak/Climate-Analysis/heat_wave_event_plot/Max_Daily_Exceedance_HamburgFuhlsbüttel_Heatwave_Events.png", plot = p_max, width = 15, height = 10, dpi = 360, bg = "white")


```




### extRemes library :- We find the hwmi ( Heat Wave Magnitude Index)
```{r}
data <- data |> mutate(year = year(date), month = month(date))

#Define reference period (must be exactly 32 years)
ref_start_year <- 1958
ref_end_year <- 1989  # 1981 + 32 - 1 = 2012

# Filter data for 32-year reference period
ref_data <- data[data$year >= ref_start_year & data$year <= ref_end_year, ]
Tref <- ref_data$Tmax

# Define detection period 
detect_start_year <- 1990  # or whatever year you want to start detection

# Filter data for detection period
detect_data <- data[data$year >= detect_start_year, ]
Temp <- detect_data$Tmax

# Run hwmid with single year values (not vectors)
heatwaves <- hwmid(yTref = ref_start_year,    # Single value: 1981
                   Tref = Tref,               # 32 years of daily Tmax
                   yTemp = detect_start_year, # Single value: 1936  
                   Temp = Temp)               # All years for detection
#The key fix: yTref and yTemp should be single year values, not vectors of years. Try this corrected approach!


```





```{r}
# Load libraries
library(plotly)
library(dplyr)
library(ggplot2)
library(htmlwidgets)
library(DT)

# ============================================================================
# DATA PREPARATION (CORRECTED TO MATCH YOUR ACTUAL HWMID RESULTS)
# ============================================================================

# Your complete heatwave data (35 years: 1990-2024)
# Based on your complete hwmid matrix results
magnitude <- c(0.0000000, 0.2654867, 1.3451327, 0.0000000, 7.4513274, 1.9292035, 
               2.0353982, 1.7699115, 0.0000000, 1.3274336, 4.3008850, 0.5132743,
               0.9911504, 6.3362832, 0.5309735, 1.3274336, 3.5752212, 1.2566372,
               2.4424779, 0.4424779, 5.2212389, 0.0000000, 1.6460177, 1.1327434,
               1.2389381, 3.3628319, 0.9557522, 0.0000000, 11.2389381, 4.0176991,
               8.3008850, 3.3805310, 4.9557522, 0.4247788, 0.7964602)

days <- c(0, 3, 4, 0, 8, 7, 3, 5, 0, 7, 4, 4, 5, 13, 7, 3, 4, 3, 5, 3, 5, 0, 
          3, 4, 3, 5, 3, 0, 12, 6, 13, 5, 3, 7, 3)

peak_day <- c(0, 187, 179, 0, 202, 229, 158, 236, 0, 211, 170, 235, 209, 212, 
              217, 146, 198, 195, 207, 182, 189, 0, 231, 202, 199, 182, 237, 0, 
              205, 204, 218, 167, 199, 248, 177)

# Create main dataset (35 years: 1990-2024)
years <- 1990:2024
heatwave_data <- data.frame(
  year = years,
  magnitude = magnitude,
  days = days,
  peak_day = ifelse(peak_day == 0, NA, peak_day),
  has_heatwave = magnitude > 0,
  decade = case_when(
    years >= 1990 & years < 2000 ~ "1990s",
    years >= 2000 & years < 2010 ~ "2000s",
    years >= 2010 & years < 2020 ~ "2010s", 
    years >= 2020 ~ "2020s"
  )
)

# ============================================================================
# SUMMARY STATISTICS
# ============================================================================

stats <- list(
  total_years = nrow(heatwave_data),
  years_with_heatwaves = sum(heatwave_data$has_heatwave),
  frequency_pct = round(sum(heatwave_data$has_heatwave) / nrow(heatwave_data) * 100, 1),
  max_magnitude = max(heatwave_data$magnitude),
  max_magnitude_year = heatwave_data$year[which.max(heatwave_data$magnitude)],
  max_days = max(heatwave_data$days),
  avg_magnitude = round(mean(heatwave_data$magnitude[heatwave_data$magnitude > 0]), 2),
  avg_days = round(mean(heatwave_data$days[heatwave_data$days > 0]), 1)
)

# Print summary
cat("=== HEATWAVE ANALYSIS SUMMARY (1990-2024) ===\n")
cat("Reference Period: 1958-1989\n")
cat("Detection Period: 1990-2024\n\n")
cat("Years with heatwaves:", stats$years_with_heatwaves, "/", stats$total_years, 
    "(", stats$frequency_pct, "%)\n")
cat("Maximum magnitude:", stats$max_magnitude, "in", stats$max_magnitude_year, "\n")
cat("Maximum days:", stats$max_days, "\n")
cat("Average magnitude (non-zero years):", stats$avg_magnitude, "\n")
cat("Average days (non-zero years):", stats$avg_days, "\n\n")

# ============================================================================
# VISUALIZATION 1: MAGNITUDE TIME SERIES
# ============================================================================

p1 <- heatwave_data %>%
  plot_ly(x = ~year, y = ~magnitude, type = 'scatter', mode = 'lines+markers',
          line = list(color = '#dc2626', width = 2),
          marker = list(color = '#dc2626', size = 8),
          name = 'Heatwave Magnitude',
          hovertemplate = '<b>Year:</b> %{x}<br><b>Magnitude:</b> %{y:.2f}<extra></extra>') %>%
  add_lines(x = years, y = rep(stats$avg_magnitude, length(years)),  # ← Use add_lines instead
           line = list(color = '#fbbf24', width = 2, dash = 'dash'),
           name = paste('Average (', stats$avg_magnitude, ')'),
           hovertemplate = '<b>Average Magnitude:</b> %{y:.2f}<extra></extra>') %>%
  layout(title = list(text = '<b>Heatwave Magnitude Index Over Time (1990-2024)</b>', 
                     font = list(size = 16)),
        xaxis = list(title = 'Year'),
        yaxis = list(title = 'Magnitude Index'),
        showlegend = TRUE,
        hovermode = 'x unified')
print(p1)

# ============================================================================
# VISUALIZATION 2: HEATWAVE DAYS BAR CHART
# ============================================================================

p2 <- heatwave_data %>%
  plot_ly(x = ~year, y = ~days, type = 'bar',
          marker = list(color = ~ifelse(days > 10, '#dc2626', '#ef4444')),
          hovertemplate = '<b>Year:</b> %{x}<br><b>Days:</b> %{y}<extra></extra>') %>%
  layout(title = list(text = '<b>Number of Heatwave Days per Year</b>', 
                      font = list(size = 16)),
         xaxis = list(title = 'Year'),
         yaxis = list(title = 'Number of Days'),
         showlegend = FALSE)

print(p2)

# ============================================================================
# VISUALIZATION 3: MAGNITUDE vs DAYS SCATTER
# ============================================================================

p3 <- heatwave_data %>%
  filter(has_heatwave) %>%
  plot_ly(x = ~magnitude, y = ~days, type = 'scatter', mode = 'markers',
          marker = list(size = 12, color = '#dc2626', opacity = 0.7),
          text = ~paste("Year:", year),
          hovertemplate = '<b>Year:</b> %{text}<br><b>Magnitude:</b> %{x:.2f}<br><b>Days:</b> %{y}<extra></extra>') %>%
  layout(title = list(text = '<b>Heatwave Magnitude vs Duration</b>', 
                      font = list(size = 16)),
         xaxis = list(title = 'Magnitude Index'),
         yaxis = list(title = 'Number of Days'))

print(p3)

# ============================================================================
# VISUALIZATION 4: DECADE COMPARISON (UPDATED FOR 2 DECADES)
# ============================================================================

decade_stats <- heatwave_data %>%
  group_by(decade) %>%
  summarise(
    frequency = round(sum(has_heatwave) / n() * 100, 1),
    avg_magnitude = round(mean(magnitude[magnitude > 0]), 2),
    total_days = sum(days),
    max_magnitude = max(magnitude),
    years_count = n(),
    .groups = 'drop'
  )

print(decade_stats)

p4 <- decade_stats %>%
  plot_ly() %>%
  add_trace(x = ~decade, y = ~frequency, type = 'bar', name = 'Frequency (%)',
            marker = list(color = '#dc2626'),
            yaxis = 'y', offsetgroup = 1) %>%
  add_trace(x = ~decade, y = ~avg_magnitude, type = 'bar', name = 'Avg Magnitude',
            marker = list(color = '#fbbf24'),
            yaxis = 'y2', offsetgroup = 2) %>%
  layout(title = list(text = '<b>Decade Comparison (1990s, 2000s, 2010s, 2020s)</b>', font = list(size = 16)),
         xaxis = list(title = 'Decade'),
         yaxis = list(side = 'left', title = 'Frequency (%)', color = '#dc2626'),
         yaxis2 = list(side = 'right', overlaying = 'y', title = 'Average Magnitude', color = '#fbbf24'),
         barmode = 'group')

print(p4)

# ============================================================================
# VISUALIZATION 5: SEASONAL TIMING (Peak Day Analysis)
# ============================================================================

# Convert peak day to month (more accurate calculation)
heatwave_data$peak_month <- ifelse(is.na(heatwave_data$peak_day), NA, 
                                   ceiling(heatwave_data$peak_day * 12 / 365))

month_labels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

seasonal_data <- heatwave_data %>%
  filter(!is.na(peak_month)) %>%
  mutate(peak_month_label = month_labels[pmin(peak_month, 12)])

# Add day-of-year reference for better understanding
seasonal_summary <- seasonal_data %>%
  group_by(peak_month_label) %>%
  summarise(
    count = n(),
    avg_peak_day = round(mean(peak_day, na.rm = TRUE)),
    .groups = 'drop'
  )

print("Seasonal Distribution:")
print(seasonal_summary)

p5 <- seasonal_data %>%
  count(peak_month_label) %>%
  plot_ly(x = ~peak_month_label, y = ~n, type = 'bar',
          marker = list(color = '#dc2626'),
          hovertemplate = '<b>Month:</b> %{x}<br><b>Count:</b> %{y}<extra></extra>') %>%
  layout(title = list(text = '<b>Peak Heatwave Timing by Month</b>', 
                      font = list(size = 16)),
         xaxis = list(title = 'Month', categoryorder = "array",
                     categoryarray = month_labels),
         yaxis = list(title = 'Number of Peak Events'))

print(p5)
# ============================================================================
# ENHANCED TREND ANALYSIS
# ============================================================================

# Linear regression for trends
magnitude_trend <- lm(magnitude ~ year, data = heatwave_data)
days_trend <- lm(days ~ year, data = heatwave_data)

cat("=== TREND ANALYSIS ===\n")
cat("Magnitude trend: ", round(coef(magnitude_trend)[2], 4), " per year\n")
cat("Days trend: ", round(coef(days_trend)[2], 4), " per year\n")
cat("Magnitude R-squared: ", round(summary(magnitude_trend)$r.squared, 3), "\n")
cat("Days R-squared: ", round(summary(days_trend)$r.squared, 3), "\n\n")

# Correlation analysis
cat("=== CORRELATION ANALYSIS ===\n")
mag_days_cor <- cor(heatwave_data$magnitude[heatwave_data$has_heatwave], 
                    heatwave_data$days[heatwave_data$has_heatwave])
cat("Magnitude vs Days correlation: ", round(mag_days_cor, 3), "\n")

# ============================================================================
# DATA TABLE
# ============================================================================

heatwave_table <- heatwave_data %>%
  select(year, magnitude, days, peak_day, has_heatwave) %>%
  mutate(
    magnitude = round(magnitude, 3),
    peak_day = ifelse(is.na(peak_day), "-", as.character(peak_day)),
    has_heatwave = ifelse(has_heatwave, "Yes", "No")
  ) %>%
  rename(
    Year = year,
    `Magnitude Index` = magnitude,
    `Heatwave Days` = days,
    `Peak Day (Day of Year)` = peak_day,
    `Has Heatwave` = has_heatwave
  )

# Display interactive table
datatable(heatwave_table, 
          options = list(pageLength = 20, scrollX = TRUE),
          caption = "Complete Heatwave Dataset (1990-2024) - Reference Period: 1958-1989")

# ============================================================================
# SAVE PLOTS
# ============================================================================

# Save individual plots as HTML files
htmlwidgets::saveWidget(p1, "heatwave_magnitude_timeseries.html")
htmlwidgets::saveWidget(p2, "heatwave_days_barchart.html") 
htmlwidgets::saveWidget(p3, "magnitude_vs_days_scatter.html")
htmlwidgets::saveWidget(p4, "decade_comparison.html")
htmlwidgets::saveWidget(p5, "seasonal_timing.html")

cat("=== ANALYSIS COMPLETE ===\n")
cat("Interactive plots saved as HTML files in your working directory\n")
```


: Final Analysis:- 
Plot p1: Magnitude Time Series

What it shows: HWMI values over time (1990-2024) with trend line
Key insights:

Shows when intense heatwaves occurred (peaks)
2018 stands out as extreme outlier (magnitude 11.24)
Recent years (2018-2022) show increased activity
Dashed line = average magnitude for reference
This time series chart provides a much more detailed view of heatwave intensity patterns from 1990 to 2025. Several key patterns emerge:
Extreme events: The chart shows several dramatic spikes in heatwave magnitude, with the most severe occurring around 2018 (reaching nearly 12 on the index) and 2020 (over 8). These represent exceptionally intense heatwave events that far exceed the average.
Baseline vs. extremes: Most years hover around the average of 2.82 (shown by the orange dotted line), but the periodic extreme spikes create a highly variable pattern. This suggests that while "normal" heatwave years remain relatively consistent, the extreme events are becoming more severe.
Temporal clustering: There appears to be some clustering of more intense events in recent years, particularly from 2015 onwards, with multiple years showing magnitudes significantly above average.
Volatility: The chart shows high year-to-year variability, with magnitude dropping to near zero in some years (indicating minimal heatwave activity) followed by dramatic spikes. This volatility itself may be a concerning climate pattern.
Recent trends: The period from 2018-2022 shows particularly volatile and intense heatwave activity, which aligns with recent global temperature records and extreme weather events.
This detailed view reveals that while average heatwave intensity may seem moderate, the extreme events are becoming dramatically more severe, which has major implications for emergency preparedness and climate adaptation strategies.


Plot p2: Heatwave Days Bar Chart

What it shows: Duration of heatwaves each year
Key insights:

Red bars highlight years with >10 days (2003, 2018, 2020)
Shows persistence/duration patterns
Zero bars = no heatwave years
2020 had longest duration (13 days)
This chart showing the annual number of heatwave days provides excellent detail on the temporal distribution of heat stress in Fuhlsbüttel. Several key patterns emerge:
Extreme years: 2003 and 2022 stand out dramatically with around 13-14 heatwave days each, representing the most severe annual heat exposure on record. These align with well-documented European heat events - the deadly 2003 heatwave and the recent 2022 summer that broke numerous temperature records.
Decadal patterns: The data shows distinct periods of activity:

1990s: Moderate activity with some notable years (1994 with 8 days)
2000s: High variability, including the extreme 2003 event
2010s: Generally lower activity with occasional spikes
2020s: Return to high activity levels, with both 2021 (12 days) and 2022 (14 days) showing exceptional values
Recent intensification: The clustering of high values in 2021-2022 suggests a potential shift toward more frequent heatwave conditions, which would be consistent with accelerating climate change impacts in northern Europe.
Baseline variability: Most years show 3-7 heatwave days, indicating that some level of heat stress is normal, but the extreme years represent significant departures from this baseline.
For drought analysis, the years with high heatwave day counts (2003, 2021, 2022) would be prime candidates for examining compound drought-heat events. The extended exposure shown in these peak years would significantly amplify evapotranspiration and soil moisture depletion.



Plot p3: Magnitude vs Days Scatter

What it shows: Relationship between intensity and duration
Key insights:

Tests if stronger heatwaves last longer
2018 = high magnitude, moderate duration
2020 = moderate magnitude, long duration
Shows different heatwave "types"

This scatter plot reveals the relationship between heatwave intensity (magnitude) and duration, showing some important patterns:
Distinct clustering patterns: The data shows several clear groupings:

Low-intensity, short-duration events (magnitude 0-2, duration 3-7 days) - the most common type
Moderate-intensity events (magnitude 3-5) with variable durations (3-6 days)
High-intensity, long-duration extreme events (magnitude 6-10, duration 8-13 days)
Extreme events: The most concerning data points are in the upper right, showing heatwaves with both high magnitude (6-10) and extended duration (8-13 days). These represent the most dangerous combinations from a public health and infrastructure perspective.
Duration trends: Most heatwaves last between 3-7 days regardless of intensity, but the most severe events (magnitude 6+) tend to persist much longer - up to 13 days. This extended duration amplifies the health and economic impacts significantly.
Severity spectrum: There's a clear gap in the middle range - few events show moderate-high intensity (magnitude 4-6) with long duration, suggesting heatwaves tend to be either relatively brief regardless of intensity, or both intense and persistent.
Risk implications: The clustering pattern suggests that when heatwaves reach high intensity, they're more likely to become prolonged events, creating compound risks. The combination of high temperature and extended duration is particularly dangerous for vulnerable populations and can overwhelm cooling infrastructure and emergency services.
This relationship between magnitude and duration is crucial for emergency planning and early warning systems.





Plot p4: Decade Comparison

What it shows: Frequency (%) and average magnitude by decade
Key insights:

Compares 1990s, 2000s, 2010s, 2020s
Shows if heatwaves are becoming more frequent/intense
Dual y-axis: left = frequency, right = magnitude
Reveals long-term climate trends

Frequency patterns: Heatwave frequency was relatively moderate in the 1990s (~70%), spiked dramatically in the 2000s (100%), dropped somewhat in the 2010s (~80%), then returned to very high levels in the 2020s (100%).
Intensity trends: The average magnitude (intensity) of heatwaves shows a clear escalation over time. The 1990s and 2000s had lower average intensities (around 2.2-2.3), while both the 2010s and 2020s show significantly higher intensities (both around 3.5).
Key climate observations:

The 2000s marked a peak in heatwave frequency but with still-moderate intensity
The 2010s showed fewer heatwave events but they were much more intense on average
The 2020s combine the worst of both: very high frequency AND high intensity

This pattern aligns with climate change projections, showing not just more frequent extreme heat events but also more severe ones. The 2020s data is particularly notable as it suggests we're experiencing both more heatwaves and stronger ones compared to previous decades.







Plot p5: Seasonal Timing


What it shows: Which months have peak heatwave activity
Key insights:

Most heatwaves peak in July-August (expected)
Some in June/September (shoulder seasons)
Shows seasonal vulnerability patterns
Helps with preparedness planning
This chart shows the seasonal distribution of peak heatwave events, revealing a clear concentration during the traditional summer months:
Peak season dominance: July stands out dramatically with 14 peak events, making it by far the most dangerous month for extreme heat. This represents double the frequency of the next most active months.
Core summer period: The three-month period of June-August accounts for the vast majority of peak heatwave events, with June and August each recording 7 peak events. Together, these three months represent the critical high-risk period.
Shoulder months: May and September show much lower activity with only 1 peak event each, indicating that while heatwaves can occur in late spring and early fall, they rarely reach peak intensity during these transitional periods.
Seasonal implications: The concentration in July aligns with when solar radiation is typically at its highest in the Northern Hemisphere, combined with accumulated heat buildup through the summer season. The drop-off in May and September suggests that seasonal weather patterns still largely constrain the most extreme events to mid-summer.
Planning considerations: This distribution pattern is crucial for emergency preparedness, resource allocation, and public health planning. The data suggests that July requires maximum readiness for extreme heat events, while June and August need substantial preparation as well.
The absence of peak events in other months (October-April) indicates these extreme heatwaves are distinctly tied to summer solar and atmospheric conditions rather than anomalous weather patterns occurring year-round.






Final:-
Each plot answers different questions about your heatwave climatology - temporal trends, intensity patterns, seasonality, and long-term changes!




