---
title: "Hamburg_Drought_Indices_Analysis"
author: "MANDAR PRIYA PHATAK"
date: "2025-07-18"
output:
   html_document: default
   pdf_document:
    latex_engine: xelatex
   word_document: default
  
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)

```



```{r}
rm(list = ls())
```



```{r}
library(DT)
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
library(lubridate)
library(fitdistrplus)
```
### Data upload from the folder
```{r}
data <- readRDS("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")

```
```{r}
monthly_tbl <- data |> 
  dplyr::mutate(year = year(date),
         month = month(date)) |> 
  group_by(year, month) |> 
  summarize(
    prec = sum(prec, na.rm = TRUE),
    Tavg = mean(Tavg, na.rm = TRUE),
    Tmin = mean(Tmin, na.rm = TRUE),
    Tmax = mean(Tmax, na.rm = TRUE),
    .groups = "drop"
  ) |> 
 dplyr:: mutate(date = make_date(year, month)) |> 
  dplyr::select(year, month, date, prec:Tmax)

monthly_tbl <- monthly_tbl |> drop_na()
```
## Beofore we compute RDI & SPEI Index, we want to confirm selection for method for computing PET
```{r}

## PET computation 
PET_hargreaves <- hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)

PET_thornthwaite <- thornthwaite(Tave = monthly_tbl$Tavg,lat = 53.6332)

PET_hargreaves |> as_tibble()
PET_thornthwaite |> as_tibble()

cbind(PET_hargreaves |> as_tibble() , PET_thornthwaite |> as_tibble())
```
### We select Hargreaves 
Key Observations from Your Data:

Thornthwaite shows many zero values - This is problematic as it indicates the method fails during winter months when temperatures are low
Hargreaves provides continuous values - More realistic for year-round analysis
Hargreaves generally shows higher PET values - This is typical for coastal areas

Recommendation for Hamburg Fühlsbuttel:
Use Hargreaves method for the following reasons:
Climate-Specific Advantages:

Coastal location: Hargreaves performs better in maritime climates like Hamburg
Temperature range: Better suited for the moderate temperature variations in northern Germany
Humidity considerations: While it doesn't directly use humidity, it's calibrated for regions with moderate humidity

Technical Advantages:

No zero values: Ensures continuous drought index calculation
Widely validated: Extensively tested in European climates
Robust performance: Less sensitive to extreme temperature variations

For Your Location Specifically:

Hamburg's maritime climate (Köppen Cfb) is well-suited for Hargreaves
The method works well for locations with moderate temperature ranges (like northern Germany)
Less prone to computational issues in winter 

## Computing PET
```{r}
monthly_tbl <- monthly_tbl |> 
  mutate(PET = hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)) |> 
  rename(PREC = prec)
```

### Reconaissance Drought Index
```{r}
calculate_monthly_rdi <- function(data, scale) {
  require(dplyr)
  require(zoo)
  
  data <- data %>%
    arrange(date) %>%
    filter(PET > 0, PREC >= 0) %>%
    mutate(
      PREC_sum = rollsum(PREC, k = scale, fill = NA, align = "right"),
      PET_sum = rollsum(PET, k = scale, fill = NA, align = "right"),
      alpha = ifelse(PET_sum > 0, PREC_sum / PET_sum, NA_real_)
    )
  
  valid_alphas <- data$alpha[!is.na(data$alpha) & is.finite(data$alpha) & data$alpha > 0]
  
  if (length(valid_alphas) > 30) {
    log_alphas <- log(valid_alphas)
    y_bar <- mean(log_alphas, na.rm = TRUE)
    sigma_y <- sd(log_alphas, na.rm = TRUE)
    
    data <- data %>%
      mutate(
        RDIst = ifelse(alpha > 0, (log(alpha) - y_bar) / sigma_y, NA_real_)
      )
  } else {
    data$RDIst <- NA_real_
  }
  
  # Only return date and RDIst
  return(data |>  dplyr::select(date, RDIst))
}

```


```{r}
calculate_rdist_for_scales <- function(data, scales = c(3,6,9,12,24,36,48)) {
  require(dplyr)
  
  result <- data |> dplyr:: select(date)  # Start with date column
  
  for (k in scales) {
    rdi_k <- calculate_monthly_rdi(data, scale = k) %>%
      rename(!!paste0("RDI_", k) := RDIst)
    
    result <- left_join(result, rdi_k, by = "date")
  }
  
  return(result)
}

```
```{r}
monthly_rdist_tbl <- calculate_rdist_for_scales(data = monthly_tbl)
monthly_rdist_tbl |> head()
```
### RDI 12 monthly
```{r}
rdi_12tbl <- monthly_rdist_tbl |> 
  select(date,RDI_12) |> 
  drop_na()
```


```{r}
rdi_12tbl <- rdi_12tbl |> 
  mutate(sign = ifelse(RDI_12 >0,"pos","neg")) |> filter(date>"1989-12-01")
```
```{r}
#
# Enhanced data preparation with drought categories
rdi_12tbl <- monthly_rdist_tbl |> 
  select(date, RDI_12) |> 
  drop_na() |> 
  filter(date > "1989-12-01") |> 
  mutate(
    # Original sign classification
    sign = ifelse(RDI_12 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_12 >= 2.0 ~ "Extremely Wet",
      RDI_12 >= 1.5 & RDI_12 < 2.0 ~ "Very Wet", 
      RDI_12 >= 1.0 & RDI_12 < 1.5 ~ "Moderately Wet",
      RDI_12 >= 0.5 & RDI_12 < 1.0 ~ "Slightly Wet",
      RDI_12 >= -0.5 & RDI_12 < 0.5 ~ "Normal",
      RDI_12 >= -1.0 & RDI_12 < -0.5 ~ "Mild Drought",
      RDI_12 >= -1.5 & RDI_12 < -1.0 ~ "Moderate Drought",
      RDI_12 >= -2.0 & RDI_12 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_12 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_12tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_12 < -0.5),
    severe_drought_months = sum(RDI_12 < -1.5),
    wet_months = sum(RDI_12 > 0.5),
    min_rdi = min(RDI_12),
    max_rdi = max(RDI_12),
    min_date = date[which.min(RDI_12)],
    max_date = date[which.max(RDI_12)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)


drought_stats
drought_summary
stats_text
```

 
###  RDI 12 monthly
```{r fig.align='center',fig.height=10, fig.width=15,dpi=450}

ggplot(rdi_12tbl, aes(x = date, y = RDI_12, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_text(aes(x = drought_stats$min_date, y = drought_stats$min_rdi - 0.3), 
            label = paste0("Driest: ", format(drought_stats$min_date, "%b %Y")), 
            angle = 90, hjust = 1.1, size = 3.3, color = "#8B0000") +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +
  geom_text(aes(x = drought_stats$max_date, y = drought_stats$max_rdi + 0.3), 
            label = paste0("Wettest: ", format(drought_stats$max_date, "%b %Y")), 
            angle = 90, hjust = -0.1, size = 3.3, color = "#00008B") +

# KEY INSIGHTS annotation box
  annotate("rect", 
           xmin = as.Date("2004-05-01"), xmax = as.Date("2013-03-01"),
           ymin = 2.3, ymax = 3.2, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = as.Date("2005-01-01"), y = 3.0, 
           label = paste0("KEY INSIGHTS (", min(year(rdi_12tbl$date)), "-", max(year(rdi_12tbl$date)), ")"),
           size = 5, fontface = "bold", hjust = -0.1,) +

  annotate("text", x = as.Date("2004-05-01"), y = 2.8, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#CC0000") +

  annotate("text", x = as.Date("2004-05-01"), y = 2.6, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#8B0000",fontface="bold") +

  annotate("text", x = as.Date("2019-01-01"), y = -2.4, 
           label = paste0(" Event Extreme : ", format(drought_stats$min_date, "%b %Y"), 
                          " (RDI:",round(drought_stats$min_rdi, 1), ") •"),
           size = 5, hjust = 1, color = "grey30",fontface="bold") +
annotate("text", x = drought_stats$max_date, y = drought_stats$max_rdi + 0.2, 
         label = paste0("• Wettest event:",format(drought_stats$max_date, "%b %Y"),
                        " (RDI:",round(drought_stats$max_rdi, 1),")"),
         size = 5, hjust = 0, color = "grey30", fontface="bold") +
  # Start and end of analysis period
  annotate("text", x = min(rdi_12tbl$date), y = 3, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40",fontface="bold") +
  annotate("text", x = max(rdi_12tbl$date), y = 3, 
           label = "End of analysis", hjust = .9, size = 5, color = "grey50",fontface="bold") +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-05-28")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    breaks = seq(-3.5, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and theme
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1990–2025*)",
    subtitle = paste0("RDI (12-Month Scale) | ", stats_text),
    caption = "Source: DWD ",
    x = "", 
    y = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t=10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15, colour = "grey10",face = "bold"),
    axis.text.y = element_text(size = 15, colour = "gray10",face="bold"),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key.size = unit(1, "cm"),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )
```
The 12-month RDI analysis for Hamburg Fuhlsbüttel from 1990 to 2025 reveals a shifting pattern in annual drought dynamics. While the early 1990s experienced substantial wet conditions—including the wettest period in May 1994 (RDI = 2.6)—recent decades have seen more frequent and intense droughts, culminating in the extreme event of January 2019 (RDI = -3.0). Overall, 28.2% of the period was marked by drought (RDI < -0.5), and 7.8% experienced severe drought (RDI < -1.5), indicating a significant increase in water stress risk, particularly in the last decade. These findings emphasize the importance of ongoing monitoring and adaptation strategies, especially in light of potential climate-driven shifts in hydrological cycles.

```{r fig.align='center',fig.height=10, fig.width=15,dpi=450}


# Print the final plot
print(p1)

#ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/RDI_24_Hamburg.png", plot = p1, width = 15, height = 10, dpi = 450)
```
 
 





#### RDI 24 monthly 
```{r fig.align='center',fig.height=10,fig.width=15}

#
# Enhanced data preparation with drought categories
rdi_24tbl <- monthly_rdist_tbl |> 
  select(date, RDI_24) |> 
  drop_na() |> 
  filter(date > "1989-12-01") |> 
  mutate(
    # Original sign classification
    sign = ifelse(RDI_24 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_24 >= 2.0 ~ "Extremely Wet",
      RDI_24 >= 1.5 & RDI_24 < 2.0 ~ "Very Wet", 
      RDI_24 >= 1.0 & RDI_24 < 1.5 ~ "Moderately Wet",
      RDI_24 >= 0.5 & RDI_24 < 1.0 ~ "Slightly Wet",
      RDI_24 >= -0.5 & RDI_24 < 0.5 ~ "Normal",
      RDI_24 >= -1.0 & RDI_24 < -0.5 ~ "Mild Drought",
      RDI_24 >= -1.5 & RDI_24 < -1.0 ~ "Moderate Drought",
      RDI_24 >= -2.0 & RDI_24 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_24 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_24tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_24 < -0.5),
    severe_drought_months = sum(RDI_24 < -1.5),
    wet_months = sum(RDI_24 > 0.5),
    min_rdi = min(RDI_24),
    max_rdi = max(RDI_24),
    min_date = date[which.min(RDI_24)],
    max_date = date[which.max(RDI_24)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)


```

```{r fig.align='center',fig.height=10, fig.width=15}
library(ggplot2)
library(dplyr)
library(scales)
library(lubridate)

# Plot
p1 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_text(aes(x = drought_stats$min_date, y = drought_stats$min_rdi - 0.3), 
            label = paste0("Driest: ", format(drought_stats$min_date, "%b %Y")), 
            angle = 90, hjust = 1.1, size = 3.3, color = "#8B0000") +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +
  geom_text(aes(x = drought_stats$max_date, y = drought_stats$max_rdi + 0.3), 
            label = paste0("Wettest: ", format(drought_stats$max_date, "%b %Y")), 
            angle = 90, hjust = -0.1, size = 3.3, color = "#00008B") +

  # KEY INSIGHTS annotation box
  annotate("rect", 
           xmin = as.Date("2004-05-01"), xmax = as.Date("2013-03-01"),
           ymin = 2.3, ymax = 3.2, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = as.Date("2005-01-01"), y = 3.0, 
           label = paste0("KEY INSIGHTS (", min(year(rdi_24tbl$date)), "-", max(year(rdi_24tbl$date)), ")"),
           size = 5, fontface = "bold", hjust = -0.1,) +

  annotate("text", x = as.Date("2004-05-01"), y = 2.8, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#CC0000") +

  annotate("text", x = as.Date("2004-05-01"), y = 2.6, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#8B0000",fontface="bold") +

  annotate("text", x = as.Date("2020-01-01"), y = -2.4, 
           label = paste0(" Event Extreme : ", format(drought_stats$min_date, "%b %Y"), 
                          " (RDI:",round(drought_stats$min_rdi, 1), ") •"),
           size = 5, hjust = 1, color = "grey30",fontface="bold") +
annotate("text", x = drought_stats$max_date, y = drought_stats$max_rdi + 0.2, 
         label = paste0("• Wettest event:",format(drought_stats$max_date, "%b %Y"),
                        " (RDI:",round(drought_stats$max_rdi, 1),")"),
         size = 5, hjust = 0, color = "grey30", fontface="bold") +
  # Start and end of analysis period
  annotate("text", x = min(rdi_24tbl$date), y = 3.4, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40",fontface="bold") +
  annotate("text", x = max(rdi_24tbl$date), y = 3.4, 
           label = "End of analysis", hjust = .9, size = 5, color = "grey50",fontface="bold") +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-05-28")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    breaks = seq(-2.5, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and theme
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1990–2025*)",
    subtitle = stats_text,
    caption = "Source: DWD ",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t=10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15, colour = "grey10",face = "bold"),
    axis.text.y = element_text(size = 15, colour = "gray10",face="bold"),
    axis.title.y = element_text(size = 16, margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key.size = unit(1, "cm"),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  )

# Print the final plot
print(p1)

ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/RDI_24_Hamburg.png", plot = p1, width = 15, height = 10, dpi = 450)


# Create detailed summary table
drought_summary <- rdi_24tbl |>
  count(drought_category, name = "months") |>
  mutate(
    percentage = round(months / sum(months) * 100, 1),
    total_months = sum(months)
  ) |>
  arrange(match(drought_category, levels(rdi_24tbl$drought_category)))

# Print summary statistics
cat("\n=== DROUGHT ANALYSIS SUMMARY ===\n")
cat("Period:", min(rdi_24tbl$date), "to", max(rdi_24tbl$date), "\n")
cat("Total months analyzed:", nrow(rdi_24tbl), "\n\n")

cat("EXTREME EVENTS:\n")
cat("• Driest period:", format(drought_stats$min_date, "%B %Y"), 
    "with RDI =", round(drought_stats$min_rdi, 2), "\n")
cat("• Wettest period:", format(drought_stats$max_date, "%B %Y"), 
    "with RDI =", round(drought_stats$max_rdi, 2), "\n\n")

cat("CONDITION DISTRIBUTION:\n")
print(drought_summary)

cat("\nKEY INSIGHTS:\n")
cat("• Drought conditions (RDI < -0.5):", drought_stats$drought_pct, "% of the period\n")
cat("• Severe drought (RDI < -1.5):", drought_stats$severe_drought_pct, "% of the period\n")
cat("• Wet conditions (RDI > 0.5):", drought_stats$wet_pct, "% of the period\n")
```
#### Interpretation
The 24-month RDI trend for Hamburg Fuhlsbüttel from 1990 to 2024 reveals significant interannual variability in drought and wet conditions. While early decades (especially the 1990s) showed frequent and intense wet periods, the most recent decade has been characterized by prolonged dry conditions, culminating in the driest month observed in January 2020 (RDI = -2.4). Overall, drought conditions were present during 37.6% of the study period, with 4.7% classified as severe. This indicates increasing susceptibility to long-term water stress, especially post-2010.


# ```{r fig.align='center',fig.height=10,fig.width=15}
# # Alternative version with simplified color scheme (closer to your original)
# p2 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24)) +
#   geom_col(aes(fill = case_when(
#     RDI_24 >= 1.5 ~ "Very Wet",
#     RDI_24 >= 0.5 ~ "Wet", 
#     RDI_24 >= -0.5 ~ "Normal",
#     RDI_24 >= -1.5 ~ "Drought",
#     TRUE ~ "Severe Drought"
#   )), show.legend = TRUE, width = 20) +
#   
#   geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
#   geom_hline(yintercept = c(-1.5, -0.5, 0.5, 1.5), 
#              color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
#   
#   scale_fill_manual(
#     name = "Condition",
#     values = c(
#       "Severe Drought" = "#CC0000",
#       "Drought" = "#FF6633", 
#       "Normal" = "#CCCCCC",
#       "Wet" = "#3399FF",
#       "Very Wet" = "#0066CC"
#     )
#   ) +
#   
#   scale_x_date(
#     date_breaks = "5 years",
#     date_labels = "%Y", 
#     limits = as.Date(c("1990-01-01", "2025-12-31")),
#     expand = expansion(mult = c(0.01, 0.01))
#   ) +
#   
#   scale_y_continuous(
#     breaks = seq(-2.5, 3, 0.5),
#     expand = expansion(mult = c(0.05, 0.05))
#   ) +
#   
#   labs(
#     title = "24-Month Reconnaissance Drought Index (RDI)",
#     subtitle = "Hamburg Fuhlsbüttel (1990-2025) - Simplified Classification", 
#     caption = "Source: DWD | Negative values indicate drought conditions",
#     x = "",
#     y = "RDI Value"
#   ) +
#   
#   theme_minimal(base_size = 12) +
#   theme(
#     plot.title.position = "plot",
#     panel.grid.minor.x = element_blank(), 
#     panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
#     panel.grid.minor.y = element_blank(),
#     panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
#     plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 5)),
#     plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
#     plot.caption = element_text(color = "grey60", size = 12, hjust = 0),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
#     axis.text.y = element_text(size = 14),
#     axis.title.y = element_text(size = 16, margin = margin(r = 10)),
#     legend.position = "bottom", 
#     legend.title = element_text(size = 14, face = "bold"),
#     legend.text = element_text(size = 12),
#     plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
#   )
# 
# # Display alternative plot
# print(p2)

```


```{r fig.height=10,fig.width=15}
# # Alternative version with smoother date breaks for monthly data
# ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = sign)) +
#   geom_col(show.legend = FALSE, width = 25) +
#   geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
#   scale_x_date(
#     date_breaks = "2 years",  # Even finer resolution
#     date_labels = "'%y",      # Shortened year format
#     limits = as.Date(c("1990-01-01", "2025-12-31")),
#     expand = expansion(mult = c(0.01, 0.01))
#   ) + 
#   scale_y_continuous(
#     breaks = seq(-2.5, 3, 0.5),
#     expand = expansion(mult = c(0.05, 0.05))
#   ) +
#   scale_fill_manual(values = c("pos" = "#2f5952", "neg" = "#d95f02")) +
#   labs(
#     title = "24-Month Reconnaissance Drought Index (RDI)",
#     subtitle = "Hamburg Fuhlsbüttel (1990-2025)",
#     caption = "Source: DWD",
#     x = "", 
#     y = ""
#   ) +
#   theme_minimal(base_size = 12) +
#   theme(
#     plot.title.position = "plot",
#     panel.grid.minor = element_blank(),
#     panel.grid.major.x = element_line(color = "grey90", linewidth = 0.3),
#     panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
#     plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 5)),
#     plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 15)),
#     plot.caption = element_text(color = "grey60", size = 12, hjust = 0),
#     axis.text.x = element_text(size = 16),
#     axis.text.y = element_text(size = 18),
#     plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
#   )
```









```{r}
# Corrected SPEI calculation (remains the same)
monthly_tbl$BAL = monthly_tbl$PREC - monthly_tbl$PET
monthly_ts <- ts(monthly_tbl[,-c(1,2,3,5,6,7)],start = c(1936,01), end = c(2025,05), frequency = 12)

### SPEI Series
spei3 <- spei(monthly_ts[,"BAL"],3)
spei6 <- spei(monthly_ts[,"BAL"],6)
spei9 <- spei(monthly_ts[,"BAL"],9)
spei12 <- spei(monthly_ts[,"BAL"],12)
spei24 <- spei(monthly_ts[,"BAL"],24)
spei36 <- spei(monthly_ts[,"BAL"],36)
spei48 <- spei(monthly_ts[,"BAL"],48)
```


```{r}
date_seq <- seq.Date(from = as.Date("1936-01-01"), 
                     to = as.Date("2025-05-01"), 
                     by = "month")
spei3_tbl <- spei3$fitted |> as_tibble() |> 
  rename(spei_3 = x) 
spei6_tbl <- spei6$fitted |> as_tibble() |> 
  rename(spei_6 = x) 
spei9_tbl <- spei9$fitted |> as_tibble() |> 
  rename(spei_9 = x) 
spei12_tbl <- spei12$fitted |> as_tibble() |> 
  rename(spei_12 = x) 
spei24_tbl <- spei24$fitted |> as_tibble() |> 
  rename(spei_24 = x) 
spei36_tbl <- spei36$fitted |> as_tibble() |> 
  rename(spei_36 = x) 
spei48_tbl <- spei48$fitted |> as_tibble() |> 
  rename(spei_48 = x)

spei3 <- data.frame(date = date_seq,Value = spei3_tbl)
spei6 <- data.frame(date = date_seq,Value = spei6_tbl)
spei9 <- data.frame(date = date_seq,Value = spei9_tbl)
spei12 <- data.frame(date = date_seq,Value = spei12_tbl)
spei24 <- data.frame(date = date_seq,Value = spei24_tbl)
spei36 <- data.frame(date = date_seq,Value = spei36_tbl)
spei48 <- data.frame(date = date_seq,Value = spei48_tbl)

```









