---
title: "Hamburg_Precipitation_Analysis"
author: "MANDAR"
date: "2025-07-14"
output: 
  
  html_document: default
  pdf_document:
    latex_engine: xelatex
  word_document: default
---

```{r }
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)
```


```{r}
rm(list = ls())
```
```{r}
library(DT)
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
library(lubridate)

```



```{r}
# Reading the data 
data <- readRDS("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")

precipitation_tbl <- data |> 
  dplyr::select(date,prec)
## Monthly data
monthly_prec_tbl <- precipitation_tbl |> 
  mutate(year = year(date),
         month = month(date)) |> 
  group_by(year, month) |> 
  summarize(
    prec = sum(prec, na.rm = TRUE),
    .groups = "drop"
  )

annual_prec_tbl <- monthly_prec_tbl |>
  group_by(year, ) |> 
  summarize(
    prec = sum(prec, na.rm = TRUE),
    .groups = "drop"
  )


```

# Visualizing the Annual Precipitation (1936-2025)
```{r fig.height=10, fig.width=15}
# Annual Plot provides a good visualiaztion for over the period 
mean_prec <- mean(annual_prec_tbl$prec, na.rm = TRUE)
# Extreme years: max and min
extreme_years <- annual_prec_tbl %>%
  filter(prec == max(prec, na.rm = TRUE) | prec == min(prec, na.rm = TRUE))

annual_prec_tbl <- annual_prec_tbl |>
 dplyr::mutate(
    is_2025 = if_else(year == 2025, TRUE, FALSE)
  )

p0 <- annual_prec_tbl |> 
  ggplot(aes(x = year, y = prec)) +
  geom_line(color = "#0e7b0e") +

  # First: draw all years except 2025 with viridis fill
  geom_point(data = subset(annual_prec_tbl, !is_2025),
             aes(fill = prec),
             color = "black", shape = 21, size  = 5, stroke = 1.5) +

# Then: draw 2025 with a different fill (e.g., grey) and red outline
  geom_point(data = subset(annual_prec_tbl, is_2025),
             fill = "grey60", color = "red", shape = 21, size  = 5, stroke = 2) +

geom_hline(yintercept = mean_prec, linetype = "dotdash", color = "darkred", linewidth = 1) +
  annotate("text", x = 1938, y = mean_prec + 150,
           label = paste0("Mean = ", round(mean_prec, 1), " mm"),
           color = "darkred", fontface = "bold", size = 6, hjust = 0) +

  scale_fill_viridis_c(option = "C", direction = -1, name = "Precip. (mm)") +

  geom_text(data = extreme_years, aes(label = year), vjust = -1.5, fontface = "bold", size  = 6) +

  scale_y_continuous(
    limits = c(100, 1200),
    breaks = c(seq(100, 1200, by = 300), 1200),
    expand = c(0, 0)
  ) +
  scale_x_continuous(
    breaks = c(seq(1936, 2025, 10), 2025),
    expand = c(0.05, 0.05),
    limits = c(1936, 2025)
  ) +

  labs(
    title = "Annual Precipitation in Hamburg (1936–2025)",
    subtitle = "Based on daily aggregated data (January 1936 – June 2025)",
    caption = "Source:- Hamburg Fuhlsbüttel Deutscher Wetterdienst (DWD)",
    x = NULL,
    y = "Precipitation (mm)"
  ) +
  theme_minimal() +
theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0, face = "bold", colour = "black", size  = 30, margin = margin(b = 20)),
    plot.subtitle = element_text(hjust = 0, face = "italic", colour = "black", size = 20, margin = margin(b = 20)),
    plot.caption = element_text(hjust = 1, face = "italic", colour = "gray50", size  = 20),
    axis.line = element_line(color = "black", linewidth = 1),
    axis.title = element_text(face = "bold", color = "black", size = 22),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black", size  = 20),
    axis.text.y = element_text(face = "bold", color = "black", size = 20),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(size = 0.5),
    panel.background = element_blank(),
    plot.background = element_blank(),
    legend.position = "none"
  )
print(p0)
    ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/Annual_Precipitation(1936-2025).png", plot = p0, width = 15, height = 10, dpi = 450)
  
```

### 



```{r fig.align = 'center',fig.height=10, fig.width=15}
pci_annual <- monthly_prec_tbl%>%
  group_by(year) |> 
  summarise(
    pci = (sum(prec^2) / (sum(prec))^2) * 100,
    .groups = 'drop'
  )



pci_annual <- monthly_prec_tbl %>%
  group_by(year) %>%
  summarise(
    pci = (sum(prec^2) / (sum(prec))^2) * 100,
    .groups = 'drop'
  ) |> filter(year < 2025 & year > 1989)

p1 <- ggplot(pci_annual, aes(x = year, y = pci)) +
  geom_line(col = "darkblue", linewidth  = 1.2) +
  scale_x_continuous(
    breaks = seq(1990, 2024, by = 2), 
    limits = c(1990, 2024)
  ) +
  scale_y_continuous(
    breaks = seq(9, 13.5, by = 0.5),
    limits = c(9, 13.5)
  ) +
  geom_smooth(method = "loess", se = FALSE, col = "black", span = 0.2, linewidth  = 1.5) +
  labs(
    title = "PCI Trend in Hamburg  Fuhlsbüttel (1990-2024)",
    subtitle = "Precipitation Concentration Index with LOESS smoothing",
    caption ="Source: DWD ",
    y = "", 
    x = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, margin = margin(b=10)),
    plot.subtitle = element_text(size   = 20, hjust = 0.5, color = "gray40"),
    plot.caption = element_text(size   = 20, hjust = 1, color = "gray50"),
    axis.title.x = element_text(size   = 14, face = "bold"),
    axis.title.y = element_text(size =  14, face = "bold"),
    axis.text.x = element_text(size = 20, angle = 45, hjust = 1, colour = "gray50"),
    axis.text.y = element_text(size  = 20, colour = "grey10"),
    panel.grid.major = element_line(color = "gray90", linewidth  = 0.5),
    panel.grid.minor = element_line(color = "gray95", linewidth  = 0.3),
    plot.margin = margin(20, 20, 20, 20)
  )

  ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/Precipitation_Concentration_Index.png", plot = p1, width = 15, height = 10, dpi = 450)
```
### What we can infere
Key Findings:
1. Overall Stability:

PCI values consistently range between ~9.0-13.5 over 89 years
Mean appears stable around 10.5-10.6 (moderate concentration)
No significant long-term trend (black LOESS line is relatively flat)

2. Temporal Patterns:

1940s-1950s: Higher variability and more frequent spikes above 13
1960s-1980s: More stable period with fewer extreme values
1990s-2020s: Return to higher variability, similar to 1940s-1950s

3. Climate Implications:

Hamburg maintains moderate seasonal precipitation concentration throughout the record
No evidence of increasing concentration due to climate change
The variability suggests natural climate oscillations rather than systematic change

4. Extreme Years:

Several years show PCI >13 (irregular distribution): notably in 1940s, 1960s, and 2000s-2010s
These likely represent years with particularly wet or dry seasons
### Other intepretations
German Climate Context Analysis:

Regional Climate Influences: Hamburg, being in northern Germany near the North Sea, is particularly influenced by Atlantic weather systems and the North Atlantic Oscillation (NAO). The cyclical peaks you see (1999, 2011, 2018) likely correspond to periods when these Atlantic systems created more concentrated precipitation patterns.
Seasonal Concentration: In Germany's temperate oceanic climate, high PCI values often indicate:

Wet winters followed by dry summers (or vice versa)
Clustering of precipitation into fewer, more intense events
Potential impacts from blocking high-pressure systems over Central Europe


Climate Change Signals: The increasing volatility after 2010 aligns with observed climate change impacts in Germany, including:

More frequent extreme weather events
Shifts in seasonal precipitation patterns
Increased likelihood of both drought periods and heavy rainfall events


Recent Trends (2020-2024): The declining trend toward 2024 might reflect:

Changes in storm track patterns affecting northern Germany
Potential shifts in the timing or intensity of Atlantic low-pressure systems
Regional impacts of broader European climate variability


Agricultural/Hydrological Implications: For Hamburg and northern Germany, these PCI fluctuations have significant implications for water management, agriculture, and flood risk - particularly given the region's importance for German agriculture and its proximity to major river systems.

The data suggests Hamburg has experienced increasingly variable precipitation concentration patterns, which is consistent with climate projections for Northern European regions.
```{r,fig.align='center',fig.height=10,fig.width=15}
library(dplyr)
library(lubridate)
library(ggplot2)

# Step 1: Add year column
precipitation_tbl <- precipitation_tbl %>%
  mutate(year = year(date))

# Step 2: Create dry/wet flags
precip_flags <- precipitation_tbl %>%
  mutate(wet = replace_na(prec >= 1, FALSE)) %>%  # assume NA means dry
  pull(wet)
# Step 3: Run Length Encoding
rle_spells <- rle(precip_flags)

# Step 4: Convert RLE to spell table
spell_tbl <- tibble(
  length = rle_spells$lengths,
  type = ifelse(rle_spells$values, "wet", "dry")
)

# Step 5: Assign dates to spells (based on end index)
spell_tbl <- spell_tbl %>%
  mutate(end_index = cumsum(length),
         start_index = lag(end_index, default = 1),
         year = year(precipitation_tbl$date[end_index]))

# Step 6: Calculate average duration by year and type
spell_summary <- spell_tbl %>%
  filter(!is.na(year)) %>%  # prevent NAs from appearing
  group_by(year, type) %>%
  summarise(avg_duration = mean(length), .groups = "drop") %>%
  mutate(
    type = factor(type, levels = c("dry", "wet"))  # enforce clean factor levels
  ) 

# Optional: Check levels

# Step 7: Plot
p2 <- ggplot(spell_summary, aes(x = year, y = avg_duration, color = type)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("dry" = "#e41a1c", "wet" = "#1f78b4")) +
  labs(title = "Average Duration of Wet and Dry Spells",x="",
       y = "Average Spell Length (days)",
       color = "Spell Type") +
  theme_minimal(base_size = 14)


p2 <- p2 +
  geom_point(data = filter(spell_summary, year == 2025),
             aes(x = year, y = avg_duration, color = type),
             shape = 21, size = 4, fill = "black", stroke = 1.5) +
  geom_text(data = filter(spell_summary, year == 2025),
            aes(label = round(avg_duration, 1)),
            vjust = -1, fontface = "bold", color = "black", size = 6) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, margin = margin(b=10)),
    plot.subtitle = element_text(size   = 20, hjust = 0.5, color = "gray40"),
    plot.caption = element_text(size   = 20, hjust = 1, color = "gray50"),
    axis.title.y = element_text(size =  14, face = "bold"),
    axis.text.x = element_text(size = 20, angle = 45, hjust = 1, colour = "gray50"),
    axis.text.y = element_text(size  = 20, colour = "grey10"),
    legend.title = element_text(size = 16, face = "bold"),
     legend.text = element_text(size = 14, colour = "black"),
    legend.position = "bottom",
    panel.grid.major = element_line(color = "gray90", linewidth  = 0.5),
    panel.grid.minor = element_line(color = "gray95", linewidth  = 0.3),
    #plot.margin = margin(20, 20, 20, 20)
  )

 ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/Average_Duration_Spells.png", plot = p2, width = 15, height = 10, dpi = 450)
```















```{r}
library(ggplot2)

# Calculate monthly climatology (mean for each month)
monthly_anomalies <- monthly_prec_tbl %>%
  group_by(month) %>%
  mutate(monthly_mean = mean(prec, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(anomaly = prec - monthly_mean)

# Plot heatmap
ggplot(monthly_anomalies, aes(x = factor(month), y = factor(year), fill = anomaly)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "brown", mid = "white", high = "blue", midpoint = 0) +
  labs(title = "Monthly Precipitation Anomalies",
       x = "Month", y = "Year", fill = "Anomaly (mm)")

```
```{r}
library(changepoint)

# Convert to time series
annual_tbl <- annual_prec_tbl |> filter(year > 1989)

annual_ts <- ts(annual_tbl$prec, start = min(annual_tbl$year))

# Detect change point in mean
cpt_mean <- cpt.mean(annual_ts, method = "PELT", penalty  = "SIC")
# Plot
plot(cpt_mean, main = "Change Point in Annual Precipitation Mean")

```
```{r}
library(changepoint)

cpt_mean_binseg <- cpt.mean(annual_ts, method = "BinSeg", Q = 1)

# Extract changepoint index and year
cpts(cpt_mean_binseg)
time(annual_ts)[cpts(cpt_mean_binseg)]

# Plot
plot(cpt_mean_binseg, main = "Change Point in Annual Precipitation (BinSeg, Q=1)")

```


```{r}
 # Compute percentiles
quantiles <- quantile(annual_tbl$prec, probs = c(0.1, 0.25, 0.75, 0.9), na.rm = TRUE)

# Classify each year
annual_tbl <- annual_tbl %>%
  mutate(class = case_when(
    prec < quantiles[1] ~ "Very Dry",
    prec < quantiles[2] ~ "Dry",
    prec <= quantiles[3] ~ "Normal",
    prec <= quantiles[4] ~ "Wet",
    TRUE ~ "Very Wet"
  ))

# Plot
ggplot(annual_tbl, aes(x = year, y = prec, fill = class)) +
  geom_col(color = "black") +
  scale_fill_manual(values = c("Very Dry" = "#d73027", "Dry" = "#fc8d59",
                               "Normal" = "gray80", "Wet" = "#91bfdb",
                               "Very Wet" = "#4575b4")) +
  labs(title = "Annual Precipitation Classification",
       y = "Precipitation (mm)", fill = "Category")

```

```{r}
# Split data
before_1994 <- annual_prec_tbl |> filter(year <= 1994)
after_1994  <- annual_prec_tbl |> filter(year > 1994)

# Calculate means
mean_before <- mean(before_1994$prec)
mean_after  <- mean(after_1994$prec)

cat("Mean Precipitation Before 1994:", round(mean_before, 1), "mm\n")
cat("Mean Precipitation After 1994:", round(mean_after, 1), "mm\n")

```
## Seasonal Precipitation
```{r}
seasonal_prec <- monthly_prec_tbl |> 
  mutate(
    season = case_when(
      month == 12 | month == 1 | month == 2 ~ "Winter",
      month >= 3 & month <= 5 ~ "Spring",
      month >= 6 & month <= 8 ~ "Summer",
      month == 9 | month == 10 | month == 11 ~ "Fall",
      TRUE ~ NA_character_
    ),
    # Handle winter year grouping
    season_year = case_when(
      month == 12 ~ year + 1,
      TRUE ~ year
    )
  ) |> 
  group_by(season_year, season) |> 
  summarise(
    total_prec = sum(prec, na.rm = TRUE),
    mean_prec = mean(prec, na.rm = TRUE),
    n_months = n(),
    .groups = "drop"
  ) |> 
  filter(n_months == 3)
```

## 
```{r}
seasonal_prec_1990_2025 <- seasonal_prec |> filter(season_year > 1989)

# Calculate trends for each season
seasonal_trends <- seasonal_prec |> 
  group_by(season) |> 
  summarise(
    trend_slope = cor.test(season_year, total_prec)$estimate,
    p_value = cor.test(season_year, total_prec)$p.value,
    mann_kendall_p = trend::mk.test(total_prec)$p.value,
    # Changed this line
    .groups = "drop"
  )
seasonal_trends
```





```{r,fig.align='center',fig.width=15,fig.height=10}
# Create individual plots for each season
library(ggplot2)

seasonal_plots <- seasonal_prec |> 
  group_by(season) |> 
  group_map(~ {
    ggplot(.x, aes(x = season_year, y = total_prec)) +
      geom_line(color = "steelblue", alpha = 0.7) +
      geom_point(color = "steelblue", size = 0.8) +
      geom_smooth(method = "lm", se = TRUE, color = "red", size = 0.8) +
      labs(
        title = paste(.y$season, "Precipitation - Hamburg Fuhlsbüttel"),
        x = "Year",
        y = "Total Precipitation (mm)",
        subtitle = paste("1936-2025 | Trend:", 
                        ifelse(seasonal_trends$p_value[seasonal_trends$season == .y$season] < 0.05, 
                               "Significant", "Not significant"))
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 10)
      )
  })

# Print all plots
seasonal_plots
```


```{r}
# Or create a faceted plot
ggplot(seasonal_prec, aes(x = season_year, y = total_prec)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(color = "steelblue", size = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", size = 0.6) +
  facet_wrap(~season, scales = "free_y", ncol = 2) +
  labs(
    title = "Seasonal Precipitation Trends - Hamburg Fuhlsbüttel (1936-2025)",
    x = "Year",
    y = "Total Precipitation (mm)"
  ) +
  theme_minimal()
```
### Test´s for Trend for Winter Season:- Using 2 diffrent time frames 1936-1979 & 1980-2025

```{r}
# Filter winter data
winter_prec <- seasonal_prec |> 
  filter(season == "Winter")

# Detailed winter trend analysis
winter_analysis <- winter_prec |> 
  summarise(
    n_years = n(),
    mean_prec = mean(total_prec, na.rm = TRUE),
    median_prec = median(total_prec, na.rm = TRUE),
    sd_prec = sd(total_prec, na.rm = TRUE),
    min_prec = min(total_prec, na.rm = TRUE),
    max_prec = max(total_prec, na.rm = TRUE),
    trend_slope = cor.test(season_year, total_prec)$estimate,
    trend_p_value = cor.test(season_year, total_prec)$p.value,
    mann_kendall_p = trend::mk.test(total_prec)$p.value,
    # Calculate actual slope in mm/year
    lm_slope = lm(total_prec ~ season_year, data = winter_prec)$coefficients[2]
  )

print(winter_analysis)

# Create detailed winter plot
library(ggplot2)

winter_plot <- ggplot(winter_prec, aes(x = season_year, y = total_prec)) +
  geom_line(color = "steelblue", alpha = 0.7, size = 0.8) +
  geom_point(color = "steelblue", size = 1.2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "red", size = 1, alpha = 0.7) +
  labs(
    title = "Winter Precipitation Trend - Hamburg Fuhlsbüttel",
    subtitle = paste("1936-2025 | Slope:", round(winter_analysis$lm_slope, 2), "mm/year | p =", round(winter_analysis$trend_p_value, 3)),
    x = "Year",
    y = "Total Winter Precipitation (mm)",
    caption = "Winter = December-January-February | Trend line with 95% confidence interval"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(1940, 2020, 20))

print(winter_plot)

# Additional analysis: Before/after comparison
winter_change <- winter_prec |> 
  mutate(
    period = case_when(
      season_year < 1980 ~ "1936-1979",
      season_year >= 1980 ~ "1980-2025"
    )
  ) |> 
  group_by(period) |> 
  summarise(
    mean_prec = mean(total_prec, na.rm = TRUE),
    median_prec = median(total_prec, na.rm = TRUE),
    sd_prec = sd(total_prec, na.rm = TRUE),
    n_years = n(),
    .groups = "drop"
  ) |> 
  mutate(
    change_mean = mean_prec - lag(mean_prec),
    change_percent = (change_mean / lag(mean_prec)) * 100
  )

print(winter_change)

# Boxplot comparison
period_plot <- winter_prec |> 
  mutate(
    period = case_when(
      season_year < 1980 ~ "1936-1979",
      season_year >= 1980 ~ "1980-2025"
    )
  ) |> 
  ggplot(aes(x = period, y = total_prec, fill = period)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Winter Precipitation: Before vs After 1980",
    x = "Period",
    y = "Total Winter Precipitation (mm)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(period_plot)
```
### Test´s for Trend for Winter Season:- Using 2 diffrent time frames 1936-1989 & 1990-2025
```{r}
# Filter winter data
winter_prec <- seasonal_prec |> 
  filter(season == "Winter")

# Detailed winter trend analysis
winter_analysis <- winter_prec |> 
  summarise(
    n_years = n(),
    mean_prec = mean(total_prec, na.rm = TRUE),
    median_prec = median(total_prec, na.rm = TRUE),
    sd_prec = sd(total_prec, na.rm = TRUE),
    min_prec = min(total_prec, na.rm = TRUE),
    max_prec = max(total_prec, na.rm = TRUE),
    trend_slope = cor.test(season_year, total_prec)$estimate,
    trend_p_value = cor.test(season_year, total_prec)$p.value,
    mann_kendall_p = trend::mk.test(total_prec)$p.value,
    # Calculate actual slope in mm/year
    lm_slope = lm(total_prec ~ season_year, data = winter_prec)$coefficients[2]
  )

print(winter_analysis)

# Create detailed winter plot
library(ggplot2)

winter_plot <- ggplot(winter_prec, aes(x = season_year, y = total_prec)) +
  geom_line(color = "steelblue", alpha = 0.7, size = 0.8) +
  geom_point(color = "steelblue", size = 1.2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "red", size = 1, alpha = 0.7) +
  labs(
    title = "Winter Precipitation Trend - Hamburg Fuhlsbüttel",
    subtitle = paste("1936-2025 | Slope:", round(winter_analysis$lm_slope, 2), "mm/year | p =", round(winter_analysis$trend_p_value, 3)),
    x = "Year",
    y = "Total Winter Precipitation (mm)",
    caption = "Winter = December-January-February | Trend line with 95% confidence interval"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(1940, 2020, 20))

print(winter_plot)

# Additional analysis: Before/after comparison
winter_change <- winter_prec |> 
  mutate(
    period = case_when(
      season_year < 1990 ~ "1936-1989",
      season_year >= 1990 ~ "1990-2025"
    )
  ) |> 
  group_by(period) |> 
  summarise(
    mean_prec = mean(total_prec, na.rm = TRUE),
    median_prec = median(total_prec, na.rm = TRUE),
    sd_prec = sd(total_prec, na.rm = TRUE),
    n_years = n(),
    .groups = "drop"
  ) |> 
  mutate(
    change_mean = mean_prec - lag(mean_prec),
    change_percent = (change_mean / lag(mean_prec)) * 100
  )

print(winter_change)

# Boxplot comparison
period_plot <- winter_prec |> 
  mutate(
    period = case_when(
      season_year < 1990 ~ "1936-1989",
      season_year >= 1990 ~ "1990-2025"
    )
  ) |> 
  ggplot(aes(x = period, y = total_prec, fill = period)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Winter Precipitation: Before vs After 1990",
    x = "Period",
    y = "Total Winter Precipitation (mm)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(period_plot)
```
### Analysis:
1980 Breakpoint Analysis:

Mean increase: 30.8mm (166mm → 197mm)
Percentage increase: 18.6%
Median increase: 38mm (164mm → 202mm)
Variability: Standard deviation increased from 60.3mm to 67.2mm

1990 Breakpoint Analysis:

Mean increase: 27.3mm (171mm → 198mm)
Percentage increase: 16.0%
Median increase: 23mm (175mm → 198mm)
Variability: Standard deviation increased from 59.5mm to 71.2mm

Key observations:

Both breakpoints show substantial increases (~16-19%), confirming the trend is robust
The 1980 breakpoint shows a slightly larger absolute increase (30.8mm vs 27.3mm), suggesting the trend started earlier but was more gradual
The 1990 breakpoint shows the trend is more concentrated in the recent period (36 years vs 46 years)
Increased variability in both cases - winters are not only getting wetter on average, but also more variable (higher standard deviation)
The medians follow similar patterns to the means, indicating this isn't driven by a few extreme years

Climate implications:

An 18% increase in winter precipitation over ~45 years is climatologically significant
The increased variability suggests more extreme wet and dry winters
This aligns with climate projections for Northern Europe showing wetter winters

This is a clear signal of changing winter precipitation patterns at Hamburg Fuhlsbüttel!

#######

