---
title: "Reconnaissance Drought Index Analysis"
author: "MANDAR PRIYA PHATAK"
date: "2025-07-18"
output:
   word_document: default
   html_document: default
   pdf_document:
    latex_engine: xelatex
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)

```



```{r}
rm(list = ls())
```



```{r}
library(DT)
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
library(lubridate)
library(fitdistrplus)
```
# Data upload from the folder
```{r}
data <- readRDS("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")
plot_dir <- "/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots"
```
# Monthly frequency
```{r}
monthly_tbl <- data |> 
  dplyr::mutate(year = year(date),
         month = month(date)) |> 
  group_by(year, month) |> 
  summarize(
    prec = sum(prec, na.rm = TRUE),
    Tavg = mean(Tavg, na.rm = TRUE),
    Tmin = mean(Tmin, na.rm = TRUE),
    Tmax = mean(Tmax, na.rm = TRUE),
    
    .groups = "drop"
  ) |> 
 dplyr:: mutate(date = make_date(year, month)) |> 
  dplyr::select(year, month, date, prec:Tmax)
# removing NA's
monthly_tbl <- monthly_tbl |> drop_na()
```
## Confirming PET computaiton method:- choice between Hargraves and Thornthwaite 
```{r}

## PET computation 
PET_hargreaves <- hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)

PET_thornthwaite <- thornthwaite(Tave = monthly_tbl$Tavg,lat = 53.6332)

# PET_hargreaves |> as_tibble()
# PET_thornthwaite |> as_tibble()

#cbind(PET_hargreaves |> as_tibble() , PET_thornthwaite |> as_tibble())
```
### We select Hargreaves 
Key Observations from Your Data:

Thornthwaite shows many zero values - This is problematic as it indicates the method fails during winter months when temperatures are low
Hargreaves provides continuous values - More realistic for year-round analysis
Hargreaves generally shows higher PET values - This is typical for coastal areas

Recommendation for Hamburg Fühlsbuttel:
Use Hargreaves method for the following reasons:
Climate-Specific Advantages:

Coastal location: Hargreaves performs better in maritime climates like Hamburg
Temperature range: Better suited for the moderate temperature variations in northern Germany
Humidity considerations: While it doesn't directly use humidity, it's calibrated for regions with moderate humidity

Technical Advantages:

No zero values: Ensures continuous drought index calculation
Widely validated: Extensively tested in European climates
Robust performance: Less sensitive to extreme temperature variations

For Your Location Specifically:

Hamburg's maritime climate (Köppen Cfb) is well-suited for Hargreaves
The method works well for locations with moderate temperature ranges (like northern Germany)
Less prone to computational issues in winter 

## Selecting Hargraves for the  Computation of  PET
```{r}
monthly_tbl <- monthly_tbl |> 
  mutate(PET = hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)) |> 
  rename(PREC = prec)
```
# Drought Indices:- 
## Reconnaissance Drought Index (RDI) :- Computation is as follows
```{r}
# Computation function for RDI 
calculate_monthly_rdi <- function(data, scale) {
  require(dplyr)
  require(zoo)
  
  data <- data %>%
    arrange(date) %>%
    filter(PET > 0, PREC >= 0) %>%
    mutate(
      PREC_sum = rollsum(PREC, k = scale, fill = NA, align = "right"),
      PET_sum = rollsum(PET, k = scale, fill = NA, align = "right"),
      alpha = ifelse(PET_sum > 0, PREC_sum / PET_sum, NA_real_)
    )
  
  valid_alphas <- data$alpha[!is.na(data$alpha) & is.finite(data$alpha) & data$alpha > 0]
  
  if (length(valid_alphas) > 30) {
    log_alphas <- log(valid_alphas)
    y_bar <- mean(log_alphas, na.rm = TRUE)
    sigma_y <- sd(log_alphas, na.rm = TRUE)
    
    data <- data %>%
      mutate(
        RDIst = ifelse(alpha > 0, (log(alpha) - y_bar) / sigma_y, NA_real_)
      )
  } else {
    data$RDIst <- NA_real_
  }
  
  # Only return date and RDIst
  return(data |>  dplyr::select(date, RDIst))
}
# Funciton for differrent scales
calculate_rdist_for_scales <- function(data, scales = c(3,6,9,12,24,36,48)) {
  require(dplyr)
  
  result <- data |> dplyr:: select(date)  # Start with date column
  
  for (k in scales) {
    rdi_k <- calculate_monthly_rdi(data, scale = k) %>%
      rename(!!paste0("RDI_", k) := RDIst)
    
    result <- left_join(result, rdi_k, by = "date")
  }
  
  return(result)
}


monthly_rdist_tbl <- calculate_rdist_for_scales(data = monthly_tbl)

```
# We have 3,6,9,12,24,36,48 monthly scale. So then which scales are important for the analysis. Since the period is 1936-2025, so 3,6,9 scales will not be that informative as these will refelect very short term changes. So 12,24,36,48 are better choices

## RDI 12 1936-1989
```{r fig.align='center',fig.height=10,fig.width=15}
rdi_12tbl <- monthly_rdist_tbl |> 
  dplyr::select(date,RDI_12) |> 
  drop_na()

rdi_12tbl <- rdi_12tbl |> 
  dplyr::mutate(sign = ifelse(RDI_12 >0,"pos","neg")) |> 
# Enhanced data preparation with drought categories
 dplyr:: mutate(
    # Original sign classification
    sign = ifelse(RDI_12 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_12 >= 2.0 ~ "Extremely Wet",
      RDI_12 >= 1.5 & RDI_12 < 2.0 ~ "Very Wet", 
      RDI_12 >= 1.0 & RDI_12 < 1.5 ~ "Moderately Wet",
      RDI_12 >= 0.5 & RDI_12 < 1.0 ~ "Slightly Wet",
      RDI_12 >= -0.5 & RDI_12 < 0.5 ~ "Normal",
      RDI_12 >= -1.0 & RDI_12 < -0.5 ~ "Mild Drought",
      RDI_12 >= -1.5 & RDI_12 < -1.0 ~ "Moderate Drought",
      RDI_12 >= -2.0 & RDI_12 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_12 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )
rdi_12tbl <- rdi_12tbl |> filter(date <="1989-12-01")


# Calculate key statistics for plot annotations
drought_stats <- rdi_12tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_12 < -0.5),
    severe_drought_months = sum(RDI_12 < -1.5),
    wet_months = sum(RDI_12 > 0.5),
    min_rdi = min(RDI_12),
    max_rdi = max(RDI_12),
    min_date = date[which.min(RDI_12)],
    max_date = date[which.max(RDI_12)],
    drought_pct = round(drought_months / total_months * 100, 2),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 2),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 2), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 2), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Create the plot
p01 <- ggplot(rdi_12tbl, aes(x = date, y = RDI_12, fill = drought_category)) +
  
  # RDI columns
  geom_col(width = 18, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Reference lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Extreme drought (Oct 1959) and Wet event (May 1981)
  geom_vline(xintercept = as.Date("1959-10-01"), color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = as.Date("1981-05-01"), color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Annotations for Extreme droughts
 annotate("label",
         x = drought_stats$min_date,
         y = drought_stats$min_rdi,
         label = paste0("• Extreme Drought: ", format(drought_stats$min_date, "%b %Y"),
                        " (RDI:", round(drought_stats$min_rdi, 2), ")"),
         fill = "#8B0000", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 0) +
  
  annotate("label",
         x = drought_stats$max_date,
         y = drought_stats$max_rdi,
         label = paste0(" Wettest Event : ", format(drought_stats$max_date, "%b %Y"),
                        " (RDI:", round(drought_stats$max_rdi, 2), ")•"),
         fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 1) +
  
           
           

  # KEY INSIGHTS box
  annotate("text", x = as.Date("1945-01-01"), y = 3.85, 
           label = "KEY INSIGHTS (1936–1989)",
           size = 5, fontface = "bold", hjust = 0) +
  annotate("text", x = as.Date("1945-01-01"), y = 3.5, 
           label = "• Drought conditions: 23.5% of the period",
           size = 4.5, hjust = 0, color = "#CC0000") +
  annotate("text", x = as.Date("1945-01-01"), y = 3.25, 
           label = "• Severe drought: 6.8% of the period",
           size = 4.5, hjust = 0, color = "#8B0000", fontface = "bold") +
  annotate("text", x = as.Date("1945-01-01"), y = 2.95, 
           label = "• Wet Events: 31.7%",
           size = 4.5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Start and End Labels (adjusted positioning)
  annotate("text", x = as.Date("1936-01-01"), y = 1.7, 
           label = "Start Period", hjust = 0, size = 4, color = "grey40", fontface = "bold", angle=90) +
  annotate("text", x = as.Date("1989-12-01"), y = 1.7,  # Moved further left
           label = "Ending Period", hjust = 0, size = 4, color = "grey50", fontface = "bold",angle=90) +

  # X & Y axes - KEY CHANGES HERE
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1935-01-01", "1989-12-01")),  # Extended further on both sides
    expand = expansion(mult = c(0.01, 0.01))  # Increased expansion on both sides
  ) +
  scale_y_continuous(
    limits = c(-4.2, 4.2),  # Slightly extended Y limits
    breaks = seq(-4, 4, 1),
    expand = expansion(mult = c(0.0, 0.01))  # Added small expansion
  ) +

  # Fill colors for drought categories
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and Theme
  labs(
    title = "Drought Patterns in Hamburg Fuhlsbüttel (1936–1989)",
    subtitle = paste0("RDI (12-Month Scale)|",stats_text),
    caption = "Source: DWD",
    x = "", 
    y = "RDI Value"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 20, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 15, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 12, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 14, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 15, face = "bold", colour = "black"),
    legend.text = element_text(size = 10, face = "bold"),
    legend.key.size = unit(1, "cm"),
    # Add plot margins to prevent clipping
    #plot.margin = margin(t = 20, r = 30, b = 20, l = 20, unit = "pt")  # Increased right margin
  )

p01

```
🧾 Summary Statement
Between 1936 and 1989, Hamburg experienced a relatively wet-dominant climate with fewer droughts and more frequent moisture surpluses. While Extreme droughts such as the 1959 drought (RDI -3.9) and the 1981 wet event (RDI 2.7) did occur, overall drought conditions were less intense and less frequent compared to the post-1990 period. This era serves as a hydroclimatic baseline, highlighting the shift toward drier and more variable conditions in the modern climate.


## RDI 12 for 1990-2025
```{r fig.align='center',fig.width=15,fig.height=10}
rdi_12tbl <- monthly_rdist_tbl |> 
  dplyr::select(date,RDI_12) |> 
  drop_na()


rdi_12tbl <- rdi_12tbl |> 
  dplyr::mutate(sign = ifelse(RDI_12 >0,"pos","neg")) |> 
# Enhanced data preparation with drought categories
 dplyr:: mutate(
    # Original sign classification
    sign = ifelse(RDI_12 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_12 >= 2.0 ~ "Extremely Wet",
      RDI_12 >= 1.5 & RDI_12 < 2.0 ~ "Very Wet", 
      RDI_12 >= 1.0 & RDI_12 < 1.5 ~ "Moderately Wet",
      RDI_12 >= 0.5 & RDI_12 < 1.0 ~ "Slightly Wet",
      RDI_12 >= -0.5 & RDI_12 < 0.5 ~ "Normal",
      RDI_12 >= -1.0 & RDI_12 < -0.5 ~ "Mild Drought",
      RDI_12 >= -1.5 & RDI_12 < -1.0 ~ "Moderate Drought",
      RDI_12 >= -2.0 & RDI_12 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_12 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

rdi_12tbl <- rdi_12tbl |> filter(date>"1989-12-01")



# Calculate key statistics for plot annotations
drought_stats <- rdi_12tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_12 < -0.5),
    severe_drought_months = sum(RDI_12 < -1.5),
    wet_months = sum(RDI_12 > 0.5),
    min_rdi = min(RDI_12),
    max_rdi = max(RDI_12),
    min_date = date[which.min(RDI_12)],
    max_date = date[which.max(RDI_12)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)




# Filter data for 1990-2025 period


# Calculate modern period statistics (you already have these)
# stats_text <- "Drought conditions: 28.2% | Severe drought: 7.8% | Wet events: 27.1% | Wettest: May 1994 (2.6) | Driest: Jan 2019 (-3)"

# Create the plot
p02 <- ggplot(rdi_12tbl, aes(x = date, y = RDI_12, fill = drought_category)) +
  
  # RDI columns
  geom_col(width = 18, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Reference lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Extreme drought and wet events
  geom_vline(xintercept = as.Date("2019-01-01"), color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = as.Date("1994-05-01"), color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Annotation labels for Extreme droughts
  annotate("label",
           x = as.Date("2019-01-01"),
           y = -3.0,
           label = "Extreme Drought: Jan 2019(RDI:-3.0)",
           fill = "#8B0000", color = "white", fontface = "bold", size = 5, label.size = NA, hjust=1) +
  annotate("label",
           x = as.Date("1994-05-01"),
           y = 2.6,
           label = "Wettest Event: May 1994(RDI:2.6)",
           fill = "#00008B", color = "white", hjust = 0, fontface = "bold", size = 5, label.size = NA) +

  # KEY INSIGHTS box
  annotate("rect", 
           xmin = as.Date("2004-01-01"), xmax = as.Date("2014-01-01"),
           ymin = 2.8, ymax = 3.9, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = as.Date("2004-03-01"), y = 3.75, 
           label = "KEY INSIGHTS (1990-2025)",
           size = 5, fontface = "bold", hjust = 0) +

  annotate("text", x = as.Date("2004-03-01"), y = 3.5, 
           label = "• Drought conditions: 28.2%",
           size = 4.5, hjust = 0, color = "#CC0000") +

  annotate("text", x = as.Date("2004-03-01"), y = 3.25, 
           label = "• Severe drought: 7.8%",
           size = 4.5, hjust = 0, color = "#8B0000", fontface = "bold") +

  annotate("text", x = as.Date("2004-03-01"), y = 3.0, 
           label = "• Wet Events: 27.1%",
           size = 4.5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Start and end labels
  annotate("text", x = as.Date("1990-01-01"), y = 2.5, 
           label = "Period Start", hjust = 0, size = 4, color = "grey40", fontface = "bold", angle=90) +
  annotate("text", x = as.Date("2025-01-01"), y = 2.5, 
           label = "Current", hjust = 1, size = 4, color = "grey50", fontface = "bold",angle=90) +

  # Axis and color scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1989-12-01", "2025-06-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-4, 4),
    breaks = seq(-4, 4, 1),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
   scale_fill_manual(
  name = "Condition",
  values = drought_colors,
  guide = guide_legend(
    title.position = "top",
    title.hjust = 0.5,
    nrow = 1,  # Change from 1 to 2 rows
    reverse = TRUE
  )
) +
  
  
  # Titles and themes
  labs(
    title = "Modern Drought Patterns in Hamburg Fuhlsbüttel (1990–2025*)",
    subtitle = paste0("RDI (12-Month Scale) |", stats_text),
    caption = "Source: DWD | *Data through May 2025",
    x = "", 
    y = "RDI Value"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 20, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 15, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 12, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 14, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 15, face = "bold", colour = "black"),
    legend.text = element_text(size = 10, face = "bold"),
    legend.key.size = unit(1, "cm"),
    
  )
p02
```

## RDI 12 for  1936-2025
```{r fig.align='center',fig.height=10,fig.width=15}
library(dplyr)

rdi_12tbl <- monthly_rdist_tbl |> 
  dplyr::select(date,RDI_12) |> 
  drop_na()
# Recent period chosen 1990-2025
rdi_12tbl <- rdi_12tbl |> 
  dplyr::mutate(sign = ifelse(RDI_12 >0,"pos","neg")) |> 
# Enhanced data preparation with drought categories
 dplyr:: mutate(
    # Original sign classification
    sign = ifelse(RDI_12 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_12 >= 2.0 ~ "Extremely Wet",
      RDI_12 >= 1.5 & RDI_12 < 2.0 ~ "Very Wet", 
      RDI_12 >= 1.0 & RDI_12 < 1.5 ~ "Moderately Wet",
      RDI_12 >= 0.5 & RDI_12 < 1.0 ~ "Slightly Wet",
      RDI_12 >= -0.5 & RDI_12 < 0.5 ~ "Normal",
      RDI_12 >= -1.0 & RDI_12 < -0.5 ~ "Mild Drought",
      RDI_12 >= -1.5 & RDI_12 < -1.0 ~ "Moderate Drought",
      RDI_12 >= -2.0 & RDI_12 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_12 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_12tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_12 < -0.5),
    severe_drought_months = sum(RDI_12 < -1.5),
    wet_months = sum(RDI_12 > 0.5),
    min_rdi = min(RDI_12),
    max_rdi = max(RDI_12),
    min_date = date[which.min(RDI_12)],
    max_date = date[which.max(RDI_12)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)
p0 <- ggplot(rdi_12tbl, aes(x = date, y = RDI_12, fill = drought_category)) +
  
  # RDI columns
  geom_col(width = 18, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Reference lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Extreme drought and wet events
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Annotation labels for Extreme droughts
 annotate("label",
         x = drought_stats$min_date,
         y = drought_stats$min_rdi,
         label = paste0("Extreme Drought:",format(drought_stats$min_date, "%b %Y"),
                        "(RDI: ",round(drought_stats$min_rdi, 2), ")"),
         fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA,hjust=0) +
  annotate("label",
         x = drought_stats$max_date,
         y = drought_stats$max_rdi,
         label = paste0("Wettest Event:",format(drought_stats$max_date, "%b %Y"),
                        "(RDI:",round(drought_stats$max_rdi, 2), ")"),
         fill = "#00008B", color = "white",hjust = 1, fontface = "bold", size = 5, label.size = NA) +

  # KEY INSIGHTS box
  annotate("rect", 
           xmin = as.Date("1977-04-01"), xmax = as.Date("1999-05-01"),
           ymin = 2.9, ymax = 3.9, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = as.Date("1977-05-01"), y = 3.75, 
           label = paste0("KEY INSIGHTS (", min(year(rdi_12tbl$date)), "-", max(year(rdi_12tbl$date)), ")"),
           size = 5, fontface = "bold", hjust = -0.1) +

  annotate("text", x = as.Date("1977-05-01"), y = 3.5, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +

  annotate("text", x = as.Date("1977-05-01"), y = 3.25, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +

  annotate("text", x = as.Date("1977-05-01"), y = 3, 
           label = paste0("• Wet Events:",drought_stats$wet_pct, "%"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Start and end labels
  annotate("text", x = min(rdi_12tbl$date), y = 1.5, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40", fontface = "bold",angle=90) +
  annotate("text", x = max(rdi_12tbl$date), y = 1.5, 
           label = "End of analysis", hjust = 0, size = 5, color = "grey50", fontface = "bold",angle=90) +

  # Axis and color scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1936-12-01", "2025-05-28")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-4, 4),
    breaks = seq(-4, 4, 1),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
   scale_fill_manual(
  name = "Condition",
  values = drought_colors,
  guide = guide_legend(
    title.position = "top",
    title.hjust = 0.5,
    nrow = 1,  # Change from 1 to 2 rows
    reverse = TRUE
  )
) +
  # Titles and themes
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1936–2025*)",
    subtitle = paste0("RDI (12-Month Scale)|",stats_text),
    caption = "Source: DWD",
    x = "", 
    y = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10,r=20)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 16, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 11, face = "bold"),  # Changed from size = 10 to size = 14 and added bold
    legend.key.size = unit(1, "cm"),
  )

```

#### Historical Climate Characteristics (1936-2025)
Overall Water Balance:

Hamburg experiences a slightly wet-biased climate - wet events (29.8%) are more common than drought conditions (25.4%)
Normal conditions dominate (~45% of the time)
Severe droughts are relatively rare (only 7.2% of the period)

Major Climate Periods Identified
1940s-1950s: The Dry Era

Sustained drought conditions, culminating in the extreme drought of October 1959 (RDI: -3.9)
This was the driest period in Hamburg's modern record

1960s-1980s: The Wet Period

Markedly wetter conditions with frequent positive RDI values
Peak wetness in May 1981 (RDI: 2.7) - the wettest event on record
This represents a major climate shift from the previous dry period

1990s-Present: Increased Variability

More volatile conditions with alternating wet/dry cycles
Notable drought events in early 2000s, around 2018-2020
Suggests potentially changing climate patterns

Climate Resilience Insights
Natural Variability:

Hamburg's climate shows strong decadal-scale oscillations
Extreme droughts (both wet and dry) are part of the natural system
The city has historically recovered from severe droughts

Water Security:

3 out of 4 years have adequate to excess moisture
Severe water stress is infrequent but can be prolonged when it occurs
The 1950s drought lasted several years, showing persistence is possible

Implications for Planning
Infrastructure & Water Management:

Design systems to handle RDI values from -4 to +3
Plan for drought events that could last 2-3 years (like 1950s pattern)
Expect wet periods that could cause flooding concerns

Agriculture & Ecosystems:

Hamburg's ecosystems are adapted to this wet-dry variability
Crop planning should account for roughly 1-in-4 chance of drought conditions
Irrigation needs are moderate compared to more arid regions

Climate Change Context:

Recent decades show increasing volatility rather than a clear trend toward wetter/drier
The question is whether future extremes will exceed the historical -4 to +3 range
Current variability may be the "new normal" rather than transitional

Bottom Line
Hamburg has a naturally variable but generally water-secure climate with distinct multi-decadal wet and dry phases. While severe droughts occur, they're historically followed by recovery periods. The main concern for the future may be increased unpredictability rather than systematic drying or wetting.



| Metric                        | 1936–1989            | 1990–2025            | 1936–2025            |
| ----------------------------- | -------------------- | -------------------- | -------------------- |
| **Drought Conditions**        | 23.5% of months      | **28.2%** ↑          | 25.4% of months      |
| **Severe Drought Conditions** | 6.8% of months       | **7.8%** ↑           | 7.2% of months       |
| **Wet Events**                | **31.7%** of months  | 27.1% ↓              | 29.8% of months      |
| **Wettest Event**             | May 1981 (RDI: 2.7)  | May 1994 (RDI: 2.6)  | May 1981 (RDI: 2.7)  |
| **Driest Event**              | Oct 1959 (RDI: -3.9) | Jan 2019 (RDI: -3.0) | Oct 1959 (RDI: -3.9) |

🌡️ Shift Toward More Drought-Prone Conditions
Drought conditions increased to 28.2% of months, up from 25.4% historically.

Severe droughts rose to 7.8%, compared to 7.2% in the full-period record.

Wet events declined to 27.1%, down from 29.8% during 1936–2025.

➡️ Implication: Hamburg's climate has become slightly more drought-prone in recent decades, with increasing frequency and intensity of dry conditions.

📉 New Extremes and Reduced Wet Surpluses
Driest event: January 2019 (RDI: -3.0) — most severe drought of the modern period.

Wettest event: May 1994 (RDI: 2.6) — strong but slightly below the 1981 historical maximum (RDI: 2.7).

No wet events above +2.7 RDI observed since 1981.

➡️ Implication: While extreme wet events have diminished in intensity, extreme droughts still occur, and their frequency is increasing.

🔁 Changing Balance in Climate Variability
The modern RDI distribution is nearly balanced:

28.2% drought vs. 27.1% wet events.

In contrast, the historical period showed a wet bias (29.8% wet vs. 25.4% drought).

The absence of extremely wet years after 1994 suggests a loss of moisture surplus periods.

➡️ Implication: The long-term wet-dominated climate is giving way to a more neutral or drought-inclined regime.

🧠 Interpretation and Trends
1. Drought Intensification
The rise in both total and severe drought occurrences during 1990–2025 highlights a shift toward more persistent and impactful dry spells, consistent with warming-related hydroclimatic stress.

2. Persistence and Clustering
Clusters of droughts, particularly in the early 2000s and 2015–2020, point to multi-year drought patterns rather than isolated events.

3. Declining Wet Extremes
No extremely wet events exceeding RDI > 2.7 occurred post-1981, indicating that extreme recharge periods have not matched past levels.

4. Modern Drought Benchmark
The 2019 drought (RDI: -3.0) serves as a new benchmark for planning and adaptation, nearly matching the legendary 1959 drought (RDI: -3.9).

📌 Climate Change Implications for Hamburg
Evidence of Regime Shift:
More frequent droughts.

Fewer, less intense wet events.

Balanced but more variable hydroclimate.

Possible Drivers:
Weakened Atlantic (maritime) moisture influence.

Increased continentality, favoring drier extremes.

🛠️ Practical Recommendations
Water resource planning should be based on a 28% drought frequency, not the outdated 25%.

The 2019 drought should be used as a realistic extreme scenario in infrastructure and policy assessments.

Agriculture and ecosystems must be managed with the expectation of symmetrical risk of both drought and wet periods — unlike the past wet-dominated conditions.

Bottom Line
Between 1990 and 2025, Hamburg has transitioned from a historically wet-biased climate to a more drought-sensitive one. The rising frequency of dry conditions and absence of extreme wet years underline the need for updated climate adaptation strategies.


## RDI 24 for 1936-1989
```{r fig.align='center', fig.height=10, fig.width=15}
# Enhanced data preparation with drought categories
rdi_24tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_24) |> 
  drop_na() |> 
  filter(date <= "1989-12-01") |> 
 dplyr:: mutate(
    # Original sign classification
    sign = ifelse(RDI_24 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_24 >= 2.0 ~ "Extremely Wet",
      RDI_24 >= 1.5 & RDI_24 < 2.0 ~ "Very Wet", 
      RDI_24 >= 1.0 & RDI_24 < 1.5 ~ "Moderately Wet",
      RDI_24 >= 0.5 & RDI_24 < 1.0 ~ "Slightly Wet",
      RDI_24 >= -0.5 & RDI_24 < 0.5 ~ "Normal",
      RDI_24 >= -1.0 & RDI_24 < -0.5 ~ "Mild Drought",
      RDI_24 >= -1.5 & RDI_24 < -1.0 ~ "Moderate Drought",
      RDI_24 >= -2.0 & RDI_24 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_24 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_24tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_24 < -0.5),
    severe_drought_months = sum(RDI_24 < -1.5),
    wet_months = sum(RDI_24 > 0.5),
    min_rdi = min(RDI_24),
    max_rdi = max(RDI_24),
    min_date = date[which.min(RDI_24)],
    max_date = date[which.max(RDI_24)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events:",drought_stats$wet_pct,"% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

date_range <- range(rdi_24tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date %m-% years(20)
annotation_end <- mid_date %m-% years(5)
```


```{r fig.align='center', fig.height=10, fig.width=15}
ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  # geom_text(aes(x = drought_stats$min_date, y = drought_stats$min_rdi - 0.3), 
  #           label = paste0("Driest: ", format(drought_stats$min_date, "%b %Y")), 
  #           angle = 90, hjust = 1.1, size = 3.3, color = "#8B0000") +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +
  # geom_text(aes(x = drought_stats$max_date, y = drought_stats$max_rdi + 0.3), 
  #           label = paste0("Wettest: ", format(drought_stats$max_date, "%b %Y")), 
  #           angle = 90, hjust = -0.1, size = 3.3, color = "#00008B") +

  # KEY INSIGHTS annotation box
 # Annotation box
  annotate("rect", xmin = annotation_start , xmax = annotation_end,
           ymin = 2.3, ymax = 3.45, fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +
  annotate("text", x = annotation_start, y = 3.3, 
           label = paste0("KEY INSIGHTS (", year(date_range[1]) - 1, "-", year(date_range[2]), ")"),
           size = 5, fontface = "bold", hjust = 0) +
  annotate("text", x = annotation_start, y = 3., 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +
  annotate("text", x = annotation_start, y = 2.75, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +
  annotate("text", x = annotation_start, y = 2.45, 
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

   # Annotation labels for Extreme droughts
 annotate("label",
         x = drought_stats$min_date,
         y = drought_stats$min_rdi,
         label = paste0("Extreme Drought:",format(drought_stats$min_date, "%b %Y"),
                        "(RDI: ",round(drought_stats$min_rdi, 2), ")"),
         fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA,hjust=1) +
  
  annotate("label",
         x = drought_stats$max_date,
         y = drought_stats$max_rdi,
         label = paste0("Wettest Event:",format(drought_stats$max_date, "%b %Y"),
                        "(RDI:",round(drought_stats$max_rdi, 2), ")"),
         fill = "#00008B", color = "white",hjust = 1, fontface = "bold", size = 5, label.size = NA) +

  # Start and end of analysis period
  annotate("text", x = min(rdi_24tbl$date), y = 1.75, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40",fontface="bold", angle=90) +
  annotate("text", x = max(rdi_24tbl$date), y = 1.75, 
           label = "End of analysis", hjust = 0, size = 5, color = "grey50",fontface="bold",angle=90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1936-01-01", "1992-03-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    limits = c(-3,3.5),
    breaks = seq(-3, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and theme
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1936–1989)",
     subtitle = paste0("RDI (24-Month Scale) | ", stats_text),
    caption = "Source: DWD ",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t=10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10",face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10",face="bold"),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 12),
    legend.key.size = unit(1, "cm"),
    #plot.margin = margin(t = 10, r = 20, b = 10, l = 20)
  )



```

## RDI 24 for 1990-2025


```{r fig.align='center',fig.height=10,fig.width=15}

#
# Enhanced data preparation with drought categories
rdi_24tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_24) |> 
  drop_na() |> 
  filter(date > "1989-12-01") |> 
 dplyr:: mutate(
    # Original sign classification
    sign = ifelse(RDI_24 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_24 >= 2.0 ~ "Extremely Wet",
      RDI_24 >= 1.5 & RDI_24 < 2.0 ~ "Very Wet", 
      RDI_24 >= 1.0 & RDI_24 < 1.5 ~ "Moderately Wet",
      RDI_24 >= 0.5 & RDI_24 < 1.0 ~ "Slightly Wet",
      RDI_24 >= -0.5 & RDI_24 < 0.5 ~ "Normal",
      RDI_24 >= -1.0 & RDI_24 < -0.5 ~ "Mild Drought",
      RDI_24 >= -1.5 & RDI_24 < -1.0 ~ "Moderate Drought",
      RDI_24 >= -2.0 & RDI_24 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_24 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_24tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_24 < -0.5),
    severe_drought_months = sum(RDI_24 < -1.5),
    wet_months = sum(RDI_24 > 0.5),
    min_rdi = min(RDI_24),
    max_rdi = max(RDI_24),
    min_date = date[which.min(RDI_24)],
    max_date = date[which.max(RDI_24)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)


# Plot
p1 <- ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  # geom_text(aes(x = drought_stats$min_date, y = drought_stats$min_rdi - 0.3), 
  #           label = paste0("Driest: ", format(drought_stats$min_date, "%b %Y")), 
  #           angle = 90, hjust = 1.1, size = 3.3, color = "#8B0000") +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +
  # geom_text(aes(x = drought_stats$max_date, y = drought_stats$max_rdi + 0.3), 
  #           label = paste0("Wettest: ", format(drought_stats$max_date, "%b %Y")), 
  #           angle = 90, hjust = -0.1, size = 3.3, color = "#00008B") +

  # KEY INSIGHTS annotation box
  annotate("rect", 
           xmin = as.Date("2004-05-01"), xmax = as.Date("2013-08-01"),
           ymin = 2, ymax = 3.2, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = as.Date("2005-01-01"), y = 3.0, 
           label = paste0("KEY INSIGHTS (", min(year(rdi_24tbl$date)), "-", max(year(rdi_24tbl$date)), ")"),
           size = 5, fontface = "bold", hjust = -0.1,) +

  annotate("text", x = as.Date("2004-05-01"), y = 2.75, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = -0.1, color = "#CC0000") +

  annotate("text", x = as.Date("2004-05-01"), y = 2.5, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = -0.1, color = "#8B0000",fontface="bold") +
  annotate("text", x = as.Date("2004-05-01"), y = 2.25, 
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
           size = 5, hjust = -0.1, color = "#00008B",fontface="bold") +

   # Annotation labels for Extreme droughts
 annotate("label",
         x = drought_stats$min_date,
         y = drought_stats$min_rdi,
         label = paste0("Extreme Drought:",format(drought_stats$min_date, "%b %Y"),
                        "(RDI: ",round(drought_stats$min_rdi, 2), ")"),
         fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA,hjust=1) +
  
  annotate("label",
         x = drought_stats$max_date,
         y = drought_stats$max_rdi,
         label = paste0("Wettest Event:",format(drought_stats$max_date, "%b %Y"),
                        "(RDI:",round(drought_stats$max_rdi, 2), ")"),
         fill = "#00008B", color = "white",hjust = 0, fontface = "bold", size = 5, label.size = NA) +

  # Start and end of analysis period
  annotate("text", x = min(rdi_24tbl$date), y = 3, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40",fontface="bold") +
  annotate("text", x = max(rdi_24tbl$date), y = 3, 
           label = "End of analysis", hjust = .9, size = 5, color = "grey50",fontface="bold") +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1989-12-01", "2026-01-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    limits = c(-3,3.5),
    breaks = seq(-3, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and theme
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1990–2025*)",
     subtitle = paste0("RDI (24-Month Scale) | ", stats_text),
    caption = "Source: DWD ",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t=10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10",face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10",face="bold"),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key.size = unit(1, "cm"),
    #plot.margin = margin(t = 10, r = 20, b = 10, l = 20)
  )

# Print the final plot
print(p1)

# ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/RDI_24_Hamburg.png", plot = p1, width = 15, height = 10, dpi = 450)



```


## RDI 24 for 1936-2025
```{r fig.align='center', fig.height=10,fig.width=15}
# Create RDI table and classify
rdi_24tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_24) |> 
  drop_na() |> 
  mutate(
    sign = ifelse(RDI_24 > 0, "pos", "neg"),
    drought_category = case_when(
      RDI_24 >= 2.0 ~ "Extremely Wet",
      RDI_24 >= 1.5 ~ "Very Wet", 
      RDI_24 >= 1.0 ~ "Moderately Wet",
      RDI_24 >= 0.5 ~ "Slightly Wet",
      RDI_24 >= -0.5 ~ "Normal",
      RDI_24 >= -1.0 ~ "Mild Drought",
      RDI_24 >= -1.5 ~ "Moderate Drought",
      RDI_24 >= -2.0 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"
    ),
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Compute statistics
drought_stats <- rdi_24tbl |> summarise(
  total_months = n(),
  drought_months = sum(RDI_24 < -0.5),
  severe_drought_months = sum(RDI_24 < -1.5),
  wet_months = sum(RDI_24 > 0.5),
  min_rdi = min(RDI_24),
  max_rdi = max(RDI_24),
  min_date = date[which.min(RDI_24)],
  max_date = date[which.max(RDI_24)],
  drought_pct = round(drought_months / total_months * 100, 1),
  severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
  wet_pct = round(wet_months / total_months * 100, 1)
)

# Prepare dynamic text
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  " Wet events : ", drought_stats$wet_pct, "%|",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette
drought_colors <- c(
  "Extreme Drought" = "#8B0000",
  "Severe Drought" = "#FF0000",
  "Moderate Drought" = "#FF4500",
  "Mild Drought" = "#FFA500",
  "Normal" = "#CCCCCC",
  "Slightly Wet" = "#87CEEB",
  "Moderately Wet" = "#4682B4",
  "Very Wet" = "#1E90FF",
  "Extremely Wet" = "#0000CD"
)

# Define dynamic x-axis and annotation window
date_range <- range(rdi_24tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date %m-% years(20)
annotation_end <- mid_date %m-% years(1)
```


```{r fig.align='center', fig.height=10,fig.width=15}
# Plot
ggplot(rdi_24tbl, aes(x = date, y = RDI_24, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  geom_hline(yintercept = seq(-2, 2, 0.5), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Annotation box
  annotate("rect", xmin = annotation_start , xmax = annotation_end,
           ymin = 2.25, ymax = 3.5, fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +
  annotate("text", x = annotation_start, y = 3.4,
           label = paste0("KEY INSIGHTS (", year(date_range[1]) - 1, "-", year(date_range[2]), ")"),
           size = 5, fontface = "bold", hjust = 0) +
  annotate("text", x = annotation_start, y = 3,
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +
  annotate("text", x = annotation_start, y = 2.75,
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +
  annotate("text", x = annotation_start, y = 2.4,
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Labels for events
  annotate("label", x = drought_stats$min_date, y = drought_stats$min_rdi,
           label = paste0("Extreme Drought: ", format(drought_stats$min_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$min_rdi, 2), ")"),
           fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA, hjust = 1) +
  annotate("label", x = drought_stats$max_date, y = drought_stats$max_rdi,
           label = paste0("Wettest Event: ", format(drought_stats$max_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$max_rdi, 2), ")"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 0) +

  # Start/End labels
  annotate("text", x = date_range[1], y = 3, 
           label = "Start of analysis", hjust = 1, size = 5, color = "grey40", fontface = "bold",angle = 90) +
  annotate("text", x = date_range[2], y = 3, 
           label = "End of analysis", hjust = 1, size = 5, color = "grey50", fontface = "bold", angle = 90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = c(as.Date("1936-01-01"),as.Date("2025-05-28")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-3.5, 3.5),
    breaks = seq(-3.5, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel(1936-2025*)",
    subtitle = paste0("RDI (24-Month Scale)|",stats_text),
    caption = "Source: DWD",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.grid.major.y = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14,face = "bold"),
    legend.key.size = unit(.5, "cm")
  )

```

Inference and Analysis:
The 12-month RDI analysis for Hamburg Fuhlsbüttel from 1990 to 2025 reveals a shifting pattern in annual drought dynamics. While the early 1990s experienced substantial wet conditions—including the wettest period in May 1994 (RDI = 2.6)—recent decades have seen more frequent and intense droughts, culminating in the Extreme drought of January 2019 (RDI = -3.0). Overall, 28.2% of the period was marked by drought (RDI < -0.5), and 7.8% experienced severe drought (RDI < -1.5), indicating a significant increase in water stress risk, particularly in the last decade. These findings emphasize the importance of ongoing monitoring and adaptation strategies, especially in light of potential climate-driven shifts in hydrological cycles.

 
## Interpretation & Analysis
The 24-month RDI trend for Hamburg Fuhlsbüttel from 1990 to 2024 reveals significant interannual variability in drought and wet conditions. While early decades (especially the 1990s) showed frequent and intense wet periods, the most recent decade has been characterized by prolonged dry conditions, culminating in the driest month observed in January 2020 (RDI = -2.4). Overall, drought conditions were present during 37.6% of the study period, with 4.7% classified as severe. This indicates increasing susceptibility to long-term water stress, especially post-2010.

 
## RDI 36 (1936-2025)
```{r fig.align='center', fig.height=10,fig.width=15}
# Create RDI table and classify
rdi_36tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_36) |> 
  drop_na() |> 
  dplyr::mutate(
    # Original sign classification
    sign = ifelse(RDI_36 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_36 >= 2.0 ~ "Extremely Wet",
      RDI_36 >= 1.5 & RDI_36 < 2.0 ~ "Very Wet", 
      RDI_36 >= 1.0 & RDI_36 < 1.5 ~ "Moderately Wet",
      RDI_36 >= 0.5 & RDI_36 < 1.0 ~ "Slightly Wet",
      RDI_36 >= -0.5 & RDI_36 < 0.5 ~ "Normal",
      RDI_36 >= -1.0 & RDI_36 < -0.5 ~ "Mild Drought",
      RDI_36 >= -1.5 & RDI_36 < -1.0 ~ "Moderate Drought",
      RDI_36 >= -2.0 & RDI_36 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_36 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_36tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_36 < -0.5),
    severe_drought_months = sum(RDI_36 < -1.5),
    wet_months = sum(RDI_36 > 0.5),
    min_rdi = min(RDI_36),
    max_rdi = max(RDI_36),
    min_date = date[which.min(RDI_36)],
    max_date = date[which.max(RDI_36)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ",drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Define dynamic x-axis and annotation window
date_range <- range(rdi_36tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date %m-% years(20)
annotation_end <- mid_date %m+% years(7)
# Plot
ggplot(rdi_36tbl, aes(x = date, y = RDI_36, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  geom_hline(yintercept = seq(-2, 2, 0.5), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Annotation box
  annotate("rect", xmin = annotation_start , xmax = annotation_end,
           ymin = 2.5, ymax = 3.5, fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +
  annotate("text", x = annotation_start, y = 3.35, 
           label = paste0("KEY INSIGHTS (", year(date_range[1]) - 2, "-", year(date_range[2]), ")"),
           size = 5, fontface = "bold", hjust = 0) +
  annotate("text", x = annotation_start, y = 3.15, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +
  annotate("text", x = annotation_start, y = 2.9, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +
  annotate("text", x = annotation_start, y = 2.65, 
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Labels for events
  annotate("label", x = drought_stats$min_date, y = drought_stats$min_rdi,
           label = paste0("Extreme Drought: ", format(drought_stats$min_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$min_rdi, 2), ")"),
           fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA, hjust = 1) +
  annotate("label", x = drought_stats$max_date, y = drought_stats$max_rdi,
           label = paste0("Wettest Event: ", format(drought_stats$max_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$max_rdi, 2), ")"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 0) +

  # Start/End labels
  annotate("text", x = date_range[1], y = 3, 
           label = "Start of analysis", hjust = 1, size = 5, color = "grey40", fontface = "bold",angle = 90) +
  annotate("text", x = date_range[2], y = 3, 
           label = "End of analysis", hjust = 1, size = 5, color = "grey50", fontface = "bold", angle = 90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = c(as.Date("1936-01-01"),as.Date("2025-05-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-3, 3.5),
    breaks = seq(-3, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel(1936-2025*)",
    subtitle = paste0("RDI (36-Month Scale)|",stats_text),
    caption = "Source: DWD",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.grid.major.y = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 20)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14,face = "bold"),
    legend.key.size = unit(.5, "cm")
  )
```
## RDI 36 (1936-1989)
```{r fig.align='center', fig.height=10, fig.width=15}
rdi_36tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_36) |> 
  drop_na() |> 
  dplyr::mutate(
    # Original sign classification
    sign = ifelse(RDI_36 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_36 >= 2.0 ~ "Extremely Wet",
      RDI_36 >= 1.5 & RDI_36 < 2.0 ~ "Very Wet", 
      RDI_36 >= 1.0 & RDI_36 < 1.5 ~ "Moderately Wet",
      RDI_36 >= 0.5 & RDI_36 < 1.0 ~ "Slightly Wet",
      RDI_36 >= -0.5 & RDI_36 < 0.5 ~ "Normal",
      RDI_36 >= -1.0 & RDI_36 < -0.5 ~ "Mild Drought",
      RDI_36 >= -1.5 & RDI_36 < -1.0 ~ "Moderate Drought",
      RDI_36 >= -2.0 & RDI_36 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_36 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  ) |> filter(date<"1990-01-01")

# Calculate key statistics for plot annotations
drought_stats <- rdi_36tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_36 < -0.5),
    severe_drought_months = sum(RDI_36 < -1.5),
    wet_months = sum(RDI_36 > 0.5),
    min_rdi = min(RDI_36),
    max_rdi = max(RDI_36),
    min_date = date[which.min(RDI_36)],
    max_date = date[which.max(RDI_36)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ",drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Define dynamic x-axis and annotation window
date_range <- range(rdi_36tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date %m-% years(20)
annotation_end <- mid_date %m-% years(4)

# Plot
ggplot(rdi_36tbl, aes(x = date, y = RDI_36, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  geom_hline(yintercept = seq(-2, 2, 0.5), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Annotation box
  annotate("rect", xmin = annotation_start , xmax = annotation_end,
           ymin = 2.5, ymax = 3.5, fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +
  annotate("text", x = annotation_start, y = 3.35, 
           label = paste0("KEY INSIGHTS (", year(date_range[1]) - 4, "-", year(date_range[2]), ")"),
           size = 5, fontface = "bold", hjust = 0) +
  annotate("text", x = annotation_start, y = 3.15, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +
  annotate("text", x = annotation_start, y = 2.9, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +
  annotate("text", x = annotation_start, y = 2.65, 
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Labels for events
  annotate("label", x = drought_stats$min_date, y = drought_stats$min_rdi,
           label = paste0("Extreme Drought: ", format(drought_stats$min_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$min_rdi, 2), ")"),
           fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA, hjust = 0) +
  annotate("label", x = drought_stats$max_date, y = drought_stats$max_rdi,
           label = paste0("Wettest Event: ", format(drought_stats$max_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$max_rdi, 2), ")"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 1) +

  # Start/End labels
  annotate("text", x = date_range[1], y = 3, 
           label = "Start of analysis", hjust = 1, size = 5, color = "grey40", fontface = "bold",angle = 90) +
  annotate("text", x = date_range[2], y = 3, 
           label = "End of analysis", hjust = 1, size = 5, color = "grey50", fontface = "bold", angle = 90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = c(as.Date("1936-01-01"),as.Date("1990-01-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-2.5, 3.5),
    breaks = seq(-2.5, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel(1936-1989)",
    subtitle = paste0("RDI (36-Month Scale)|",stats_text),
    caption = "Source: DWD",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.grid.major.y = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 20)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14,face = "bold"),
    legend.key.size = unit(.5, "cm")
  )
```
## RDI 36 (1990-2025)

```{r fig.align='center', fig.height=10, fig.width=15}
# Data preparation
rdi_36tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_36) |> 
  drop_na() |> 
  filter(date > "1989-12-01") |> 
  dplyr::mutate(
    # Original sign classification
    sign = ifelse(RDI_36 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_36 >= 2.0 ~ "Extremely Wet",
      RDI_36 >= 1.5 & RDI_36 < 2.0 ~ "Very Wet", 
      RDI_36 >= 1.0 & RDI_36 < 1.5 ~ "Moderately Wet",
      RDI_36 >= 0.5 & RDI_36 < 1.0 ~ "Slightly Wet",
      RDI_36 >= -0.5 & RDI_36 < 0.5 ~ "Normal",
      RDI_36 >= -1.0 & RDI_36 < -0.5 ~ "Mild Drought",
      RDI_36 >= -1.5 & RDI_36 < -1.0 ~ "Moderate Drought",
      RDI_36 >= -2.0 & RDI_36 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_36 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_36tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_36 < -0.5),
    severe_drought_months = sum(RDI_36 < -1.5),
    wet_months = sum(RDI_36 > 0.5),
    min_rdi = min(RDI_36),
    max_rdi = max(RDI_36),
    min_date = date[which.min(RDI_36)],
    max_date = date[which.max(RDI_36)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ", drought_stats$wet_pct,"% |",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Define dynamic x-axis and annotation window
date_range <- range(rdi_36tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date %m-% years(2)
annotation_end <- mid_date %m+% years(7)
```


```{r fig.align='center', fig.height=10, fig.width=15}
# Create the plot
ggplot(rdi_36tbl, aes(x = date, y = RDI_36, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

   # Annotation box
  annotate("rect", xmin = annotation_start , xmax = annotation_end,
           ymin = 2.5, ymax = 3.5, fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +
  annotate("text", x = annotation_start, y = 3.35, 
           label = paste0("KEY INSIGHTS (", year(date_range[1]) , "-", year(date_range[2]), ")"),
           size = 5, fontface = "bold", hjust = 0) +
  annotate("text", x = annotation_start, y = 3.15, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +
  annotate("text", x = annotation_start, y = 2.9, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +
  annotate("text", x = annotation_start, y = 2.65, 
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Labels for events
  annotate("label", x = drought_stats$min_date, y = drought_stats$min_rdi,
           label = paste0("Extreme Drought: ", format(drought_stats$min_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$min_rdi, 2), ")"),
           fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA, hjust = 1) +
  annotate("label", x = drought_stats$max_date, y = drought_stats$max_rdi,
           label = paste0("Wettest Event: ", format(drought_stats$max_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$max_rdi, 2), ")"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 0) +

  # Start/End labels
  annotate("text", x = date_range[1], y = 3, 
           label = "Start of analysis", hjust = 1, size = 5, color = "grey40", fontface = "bold",angle = 90) +
  annotate("text", x = date_range[2], y = 3, 
           label = "End of analysis", hjust = 1, size = 5, color = "grey50", fontface = "bold", angle = 90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = c(as.Date("1989-12-01"),as.Date("2025-06-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-3, 3.5),
    breaks = seq(-3, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel(1990-2025)",
    subtitle = paste0("RDI (36-Month Scale)|",stats_text),
    caption = "Source: DWD",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.grid.major.y = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 20)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14,face = "bold"),
    legend.key.size = unit(.5, "cm")
  )


# ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/RDI_36_Hamburg.png", plot = p2, width = 15, height = 10, dpi = 450)

```

## RDI 48 monthly RDI (1936-1989)
```{r fig.align='center', fig.height=10, fig.width=15}
# Data preparation
rdi_48tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_48) |> 
  drop_na() |> 
  filter(date <"1990-01-01") |> 
  dplyr::mutate(
    # Original sign classification
    sign = ifelse(RDI_48 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_48 >= 2.0 ~ "Extremely Wet",
      RDI_48 >= 1.5 & RDI_48 < 2.0 ~ "Very Wet", 
      RDI_48 >= 1.0 & RDI_48 < 1.5 ~ "Moderately Wet",
      RDI_48 >= 0.5 & RDI_48 < 1.0 ~ "Slightly Wet",
      RDI_48 >= -0.5 & RDI_48 < 0.5 ~ "Normal",
      RDI_48 >= -1.0 & RDI_48 < -0.5 ~ "Mild Drought",
      RDI_48 >= -1.5 & RDI_48 < -1.0 ~ "Moderate Drought",
      RDI_48 >= -2.0 & RDI_48 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_48 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_48tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_48 < -0.5),
    severe_drought_months = sum(RDI_48 < -1.5),
    wet_months = sum(RDI_48 > 0.5),
    min_rdi = min(RDI_48),
    max_rdi = max(RDI_48),
    min_date = date[which.min(RDI_48)],
    max_date = date[which.max(RDI_48)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
   "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Define dynamic x-axis and annotation window
date_range <- range(rdi_48tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date %m-% years(20)
annotation_end <- mid_date %m-% years(5)
# Plot
ggplot(rdi_48tbl, aes(x = date, y = RDI_48, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  geom_hline(yintercept = seq(-2, 2, 0.5), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # # Annotation box
   annotate("rect", xmin = annotation_start , xmax = annotation_end,
           ymin = 2, ymax = 3, fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +
   annotate("text", x = annotation_start, y = 2.85, 
           label = paste0("KEY INSIGHTS (", year(date_range[1]) - 3, "-", year(date_range[2]), ")"),
            size = 5, fontface = "bold", hjust = 0) +
   annotate("text", x = annotation_start, y = 2.65, 
            label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +
   annotate("text", x = annotation_start, y = 2.45, 
            label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
            size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +
   annotate("text", x = annotation_start, y = 2.25, 
          label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
            size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Labels for events
  annotate("label", x = drought_stats$min_date, y = drought_stats$min_rdi,
           label = paste0("Severe Drought: ", format(drought_stats$min_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$min_rdi, 2), ")"),
           fill = "#FF0000", color = "white", fontface = "bold", size = 4, label.size = NA, hjust = 0) +
  annotate("label", x = drought_stats$max_date, y = drought_stats$max_rdi,
           label = paste0("Wettest Event: ", format(drought_stats$max_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$max_rdi, 2), ")"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 1) +

  # Start/End labels
  annotate("text", x = date_range[1], y = 2.5, 
           label = "Start of analysis", hjust = 1, size = 5, color = "grey40", fontface = "bold",angle = 90) +
  annotate("text", x = date_range[2], y = 2.5, 
           label = "End of analysis", hjust = 1, size = 5, color = "grey50", fontface = "bold", angle = 90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = c(as.Date("1936-01-01"),as.Date("1990-01-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-2, 3.5),
    breaks = seq(-2, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel(1936-1989)",
    subtitle = paste0("RDI (48-Month Scale)|",stats_text),
    caption = "Source: DWD",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey90"),
    panel.grid.major.y = element_line(color = "grey90"),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 20)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 17, face = "bold"),
    legend.text = element_text(size = 15,face = "bold"),
    legend.key.size = unit(.5, "cm")
  )
```
## RDI 48 (1990-2025)

```{r fig.align='center', fig.height=10, fig.width=15}
rdi_48tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_48) |> 
  drop_na() |> 
  filter(date >= "1990-01-01") |> 
  dplyr::mutate(
    # Original sign classification
    sign = ifelse(RDI_48 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_48 >= 2.0 ~ "Extremely Wet",
      RDI_48 >= 1.5 & RDI_48 < 2.0 ~ "Very Wet", 
      RDI_48 >= 1.0 & RDI_48 < 1.5 ~ "Moderately Wet",
      RDI_48 >= 0.5 & RDI_48 < 1.0 ~ "Slightly Wet",
      RDI_48 >= -0.5 & RDI_48 < 0.5 ~ "Normal",
      RDI_48 >= -1.0 & RDI_48 < -0.5 ~ "Mild Drought",
      RDI_48 >= -1.5 & RDI_48 < -1.0 ~ "Moderate Drought",
      RDI_48 >= -2.0 & RDI_48 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_48 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_48tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_48 < -0.5),
    severe_drought_months = sum(RDI_48 < -1.5),
    wet_months = sum(RDI_48 > 0.5),
    min_rdi = min(RDI_48),
    max_rdi = max(RDI_48),
    min_date = date[which.min(RDI_48)],
    max_date = date[which.max(RDI_48)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
   "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Define dynamic x-axis and annotation window
date_range <- range(rdi_48tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date
annotation_end <- mid_date %m+% years(5)

ggplot(rdi_48tbl, aes(x = date, y = RDI_48, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # # KEY INSIGHTS annotation box
  annotate("rect",
           xmin = annotation_start, xmax = annotation_end,
           ymin = 2, ymax = 3,
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = annotation_start, y = 2.85,
         label = paste0("KEY INSIGHTS (", min(year(rdi_48tbl$date)), "-", max(year(rdi_48tbl$date)), ")"),
         size = 6, fontface = "bold", hjust = 0) +


  annotate("text", x = annotation_start, y = 2.65,
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 6, hjust = 0, color = "#CC0000") +

  annotate("text", x = annotation_start, y = 2.45,
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "%"),
           size = 6, hjust = 0, color = "#8B0000", fontface = "bold")  +
  annotate("text", x = annotation_start, y = 2.25, 
          label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
            size = 6, hjust = 0, color = "#00008B", fontface = "bold") +

 
  
  annotate("label", 
         x = drought_stats$min_date, 
         y = -2.9,
         label = paste0(format(drought_stats$min_date, " %Y %b "), 
                        "Event Extreme (RDI: ", round(drought_stats$min_rdi, 1), ")"),
         fill = "#8B0000",              # Background color of the label
         size = 5, 
         hjust = 1, 
         color = "white",       
         fontface = "bold", label.size = NA ) +
         
  annotate("label", x = drought_stats$max_date, y = drought_stats$max_rdi,
           label = paste0("Wettest Event: ", format(drought_stats$max_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$max_rdi, 2), ") •"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 0) +

  # Start and end of analysis period
  annotate("text", x = min(rdi_48tbl$date), y = 3, 
           label = "Start of analysis", hjust = 0.8, size = 5, color = "grey40", fontface = "bold", angle = 90) +
  annotate("text", x = max(rdi_48tbl$date), y = 3, 
           label = "End of analysis", hjust = 0.8, size = 5, color = "grey50", fontface = "bold", angle=90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1989-12-01", "2025-12-31")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    limits = c(-3.5, 3.5),
    breaks = seq(-3.5, 3, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and theme
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1990–2025)",
    subtitle = paste0("RDI (48-Month Scale) | ",stats_text),
    caption = "Source: DWD ",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 12),
    legend.key.size = unit(1, "cm")
  )

```

## RDI 48 (1936-2025)

```{r fig.align='center',fig.height=10, fig.width=15}
# Data preparation
rdi_48tbl <- monthly_rdist_tbl |> 
  dplyr::select(date, RDI_48) |> 
  drop_na() |> 
  dplyr::mutate(
    # Original sign classification
    sign = ifelse(RDI_48 > 0, "pos", "neg"),
    
    drought_category = case_when(
      RDI_48 >= 2.0 ~ "Extremely Wet",
      RDI_48 >= 1.5 & RDI_48 < 2.0 ~ "Very Wet", 
      RDI_48 >= 1.0 & RDI_48 < 1.5 ~ "Moderately Wet",
      RDI_48 >= 0.5 & RDI_48 < 1.0 ~ "Slightly Wet",
      RDI_48 >= -0.5 & RDI_48 < 0.5 ~ "Normal",
      RDI_48 >= -1.0 & RDI_48 < -0.5 ~ "Mild Drought",
      RDI_48 >= -1.5 & RDI_48 < -1.0 ~ "Moderate Drought",
      RDI_48 >= -2.0 & RDI_48 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought"  # RDI_48 < -2.0
    ),
    
    # Create ordered factor for proper legend ordering
    drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
    ))
  )

# Calculate key statistics for plot annotations
drought_stats <- rdi_48tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(RDI_48 < -0.5),
    severe_drought_months = sum(RDI_48 < -1.5),
    wet_months = sum(RDI_48 > 0.5),
    min_rdi = min(RDI_48),
    max_rdi = max(RDI_48),
    min_date = date[which.min(RDI_48)],
    max_date = date[which.max(RDI_48)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_rdi, 1), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_rdi, 1), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

# Define dynamic x-axis and annotation window
date_range <- range(rdi_48tbl$date)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date  %m-% years(15)
annotation_end <- mid_date %m+% years(2)
```


```{r fig.align='center',fig.height=10, fig.width=15}
# Create the plot


ggplot(rdi_48tbl, aes(x = date, y = RDI_48, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  # Add reference dashed lines
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  # Highlight driest and wettest periods
  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +

  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # # KEY INSIGHTS annotation box
  annotate("rect",
           xmin = annotation_start, xmax = annotation_end,
           ymin = 2, ymax = 3,
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = annotation_start, y = 2.85,
         label = paste0("KEY INSIGHTS (", min(year(rdi_48tbl$date)) - 3, "-", max(year(rdi_48tbl$date)), ")"),
         size = 6, fontface = "bold", hjust = 0) +


  annotate("text", x = annotation_start, y = 2.65,
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 6, hjust = 0, color = "#CC0000") +

  annotate("text", x = annotation_start, y = 2.45,
           label = paste0("•Severe drought:",drought_stats$severe_drought_pct,"%"),
           size = 6, hjust = 0, color = "#8B0000", fontface = "bold")  +
  annotate("text", x = annotation_start, y = 2.25, 
          label = paste0("•Wet Events:",drought_stats$wet_pct, "%"),
            size = 6, hjust = 0, color = "#00008B", fontface = "bold") +

 
  
  annotate("label", 
         x = drought_stats$min_date, 
         y = -2.9,
         label = paste0(format(drought_stats$min_date, "%Y%b "), 
                        "droughtExtreme (RDI: ", round(drought_stats$min_rdi, 1), ")⬤ "),
         fill = "#8B0000",              # Background color of the label
         size = 5, 
         hjust = 1, 
         color = "white",       
         fontface = "bold", label.size = NA ) +
         
  annotate("label", x = drought_stats$max_date, y = drought_stats$max_rdi,
           label = paste0("•Wettest Event:",format(drought_stats$max_date, "%b %Y"),
                          " (RDI: ", round(drought_stats$max_rdi, 3),")"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 0) +

  # Start and end of analysis period
  annotate("text", x = min(rdi_48tbl$date), y = 3, 
           label = "Start of analysis", hjust = 0.8, size = 5, color = "grey40", fontface = "bold", angle = 90) +
  annotate("text", x = max(rdi_48tbl$date), y = 3, 
           label = "End of analysis", hjust = 0.8, size = 5, color = "grey50", fontface = "bold", angle=90) +

  # Scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1936-01-01", "2025-12-31")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    limits = c(-3, 3.5),
    breaks = seq(-3, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  # Labels and theme
  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1936–2025)",
    subtitle = paste0("RDI (48-Month Scale) | ",stats_text),
    caption = "Source: DWD ",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "gray10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 12),
    legend.key.size = unit(1, "cm")
  )


#ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/RDI_48_Hamburg.png", plot = p3, width = 15, height = 10, dpi = 450)
```
```


###### Comparison of RDI 12 & RDI 24 & RDI 36 & RDI 48


Metric	RDI 12-Month	RDI 24-Month	Interpretation
Period	1990–2025*	1990–2025*	Same base period
Drought Conditions (< -0.5)	28.2%	37.6%	24-month shows more persistent dryness
Severe Drought (< -1.5)	7.8%	4.7%	12-month captures sharper, shorter extremes
Wet Conditions (> 0.5)	27.1%	28.9%	Fairly similar in wet distribution
Driest Event	Jan 2019 (RDI = -3.0)	Jan 2020 (RDI = -2.4)	12-month shows sharper negative anomaly
Wettest Event	May 1994 (RDI = 2.6)	May 1995 (RDI = 3.0)	Similar, offset by 1 year

✍️ Interpretation Summary:
While the 24-month RDI highlights more persistent drought conditions across multi-year scales, the 12-month RDI is more sensitive to year-to-year variability, capturing sharper extremes like the 2019 drought. The higher frequency of severe drought in the 12-month scale (7.8% vs 4.7%) suggests greater short-term hydroclimatic variability, even though total wet conditions are fairly consistent across both indices.

✅ Part 2: Trend Analysis – Mann-Kendall Test
```{r}
library(Kendall)

# Ensure the data is in time order and remove NAs
rdi_12_values <- rdi_12tbl |> arrange(date) |> pull(RDI_12) |> na.omit()

# Run Mann-Kendall test
mk_test_12 <- MannKendall(rdi_12_values)
print(mk_test_12)
#Do the same for RDI 24-month:


rdi_24_values <- rdi_24tbl |> arrange(date) |> pull(RDI_24) |> na.omit()
mk_test_24 <- MannKendall(rdi_24_values)
print(mk_test_24)

### 
library(Kendall)

# Ensure the data is in time order and remove NAs
rdi_36_values <- rdi_36tbl |> arrange(date) |> pull(RDI_36) |> na.omit()

# Run Mann-Kendall test
mk_test_36 <- MannKendall(rdi_36_values)
print(mk_test_36)

rdi_48_values <- rdi_48tbl |> arrange(date) |> pull(RDI_48) |> na.omit()

# Run Mann-Kendall test
mk_test_48 <- MannKendall(rdi_48_values)
print(mk_test_48)
```
```
The Mann-Kendall trend analysis reveals significant negative trends in both the 12-month and 24-month RDI series for Hamburg Fuhlsbüttel (1990–2025). The 12-month RDI exhibits a moderate downward trend (τ = -0.163, p < 0.001), indicating increasing frequency of dry years. More notably, the 24-month RDI shows a strong and highly significant downward trend (τ = -0.282, p < 0.0001), suggesting an intensification of long-term drought conditions. These results support the evidence of a drying trend in the region, particularly after 2000.
The Mann-Kendall trend analysis of the 36-month RDI series reveals a strong and statistically significant downward trend (τ = -0.438, p < 0.0001), indicating a persistent and intensifying shift toward long-term drought conditions in Hamburg Fuhlsbüttel over the period 1990–2025. This scale, which reflects hydrological droughts and multi-year water deficits, highlights a marked departure from the relatively wet conditions of the early 1990s. The trend corroborates observed extreme drought events, particularly the one peaking in January 2021 (RDI = -2.9), and suggests a long-term decline in regional water availability.

The 48-month RDI shows a persistent and intensifying long-term drought trend in Hamburg Fuhlsbüttel from 1990 to 2025. Nearly 37% of the period exhibits negative RDI values below -0.5, and 8.2% qualifies as severe drought. The most extreme drought occurred in January 2022 (RDI = -2.7), following a decade of steadily declining multi-year moisture conditions. The Mann-Kendall test confirms a strong and statistically significant drying trend (τ = -0.563, p < 0.0001), signaling major long-term hydrological stress in the region. In contrast, the early 1990s, particularly April 1995, marked a period of sustained wet conditions (RDI = 2.7), a level that has not been approached since.


##### RDI Summmary #####
```{r}
rdi_summary_tbl <- tibble::tibble(
  `RDI Scale` = c("12-Month", "24-Month", "36-Month", "48-Month"),
  `Drought (%)` = c("28.2%", "37.6%", "37.2%", "36.9%"),
  `Severe Drought (%)` = c("7.8%", "4.7%", "6.1%", "8.2%"),
  `Wettest Event` = c("May 1994 (2.6)", "May 1995 (3.0)", "Jun 1995 (2.7)", "Apr 1995 (2.7)"),
  `Driest Event` = c("Jan 2019 (-3.0)", "Jan 2020 (-2.4)", "Jan 2021 (-2.9)", "Jan 2022 (-2.7)"),
  `Kendall’s Tau` = c("-0.163", "-0.282", "-0.438", "-0.563"),
  `Trend Significance` = c("✓", "✓", "✓✓", "✓✓✓")
)



knitr::kable(rdi_summary_tbl, caption = "Comparative Summary of RDI Indices (1990–2025)")

```


```{r}
library(knitr)

# Save LaTeX-formatted table to .tex file
kable(rdi_summary_tbl, format = "latex", 
      caption = "Comparative Summary of RDI Indices (1990–2025)") |> 
  writeLines("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Tables/RDI_Summary_Table.tex")
```


```{r}
readr::write_csv(
  rdi_summary_tbl,
  "/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Tables/RDI_Summary_Table.csv"
)

```


##### End of RDI Index ######













