---
title: "SPEI_Index_Analysis"
author: "MANDAR"
date: "2025-07-22"
output: html_document
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 300, 
                      fig.width = 15, fig.height = 10)

```

```{r}
rm(list = ls())
```


```{r}
library(DT)
library(tidyverse)
library(SPEI)
library(drought)
library(extRemes)
library(heatwaveR)
library(lubridate)
library(fitdistrplus)
```
# Loading the Data


```{r}
data <- readRDS("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Data_DWD/Hamburg_Fuhlsbüttel.rds")
data <- data |> drop_na()
monthly_tbl <- data |> 
  dplyr::mutate(year = year(date),
         month = month(date)) |> 
  group_by(year, month) |> 
  summarize(
    prec = sum(prec, na.rm = TRUE),
    Tavg = mean(Tavg, na.rm = TRUE),
    Tmin = mean(Tmin, na.rm = TRUE),
    Tmax = mean(Tmax, na.rm = TRUE),
    
    .groups = "drop"
  ) |> 
 dplyr:: mutate(date = make_date(year, month)) |> 
  dplyr::select(year, month, date, prec:Tmax)

monthly_tbl <- monthly_tbl |> drop_na()
```
## From previous analysis we already know that Hargreaves method is suitable for PET
```{r}
## PEt calculation
monthly_tbl <- monthly_tbl |> 
  mutate(PET = hargreaves(Tmin = monthly_tbl$Tmin,Tmax = monthly_tbl$Tmax,Pre = monthly_tbl$prec,lat = 53.6332)) |> 
  rename(PREC = prec)
## Balance Computation
monthly_tbl$BAL <- monthly_tbl$PREC - monthly_tbl$PET
# Converting the monthly to ts format for convenience 
monthly_tbl <- ts(monthly_tbl[,-c(1,2,3)], start = c(1936,01), end = c(2025,05), frequency = 12)

```

#### SPEI Index 
```{r}

# SPEI series 3,6,9,12,24,36,48
spei_3 <- spei(monthly_tbl[,"BAL"],3)
spei_6 <- spei(monthly_tbl[,"BAL"],6)
spei_9 <- spei(monthly_tbl[,"BAL"],9)
spei_12 <- spei(monthly_tbl[,"BAL"],12)
spei_24 <- spei(monthly_tbl[,"BAL"],24)
spei_36 <- spei(monthly_tbl[,"BAL"],36)
spei_48 <- spei(monthly_tbl[,"BAL"],48)

spei_48$fitted |> as_tibble()
```
### Plot of SPEI Index
```{r}
date_seq <- seq.Date(from = as.Date("1936-01-01"), 
                     to = as.Date("2025-05-01"), 
                     by = "month")

spei_3 <- spei_3$fitted |> as_tibble()
spei_6 <- spei_6$fitted |> as_tibble()
spei_9 <- spei_9$fitted |> as_tibble()
spei_12 <- spei_12$fitted |> as_tibble()
spei_24<- spei_24$fitted |> as_tibble()
spei_36 <- spei_36$fitted |> as_tibble()
spei_48 <- spei_48$fitted |> as_tibble()
spei_3tbl <- data.frame(date = date_seq, Value = spei_3)
spei_6tbl <- data.frame(date = date_seq, Value = spei_6)
spei_9tbl <- data.frame(date = date_seq, Value = spei_9)
spei_12tbl <- data.frame(date = date_seq, Value = spei_12)
spei_24tbl <- data.frame(date = date_seq, Value = spei_24)
spei_36tbl <- data.frame(date = date_seq, Value = spei_36)
spei_48tbl <- data.frame(date = date_seq, Value = spei_48)

colnames(spei_3tbl) <- c("date","spei_3")
colnames(spei_6tbl) <- c("date","spei_6")
colnames(spei_9tbl) <- c("date","spei_9")
colnames(spei_12tbl) <- c("date","spei_12")
colnames(spei_24tbl) <- c("date","spei_24")
colnames(spei_36tbl) <- c("date","spei_36")
colnames(spei_48tbl) <- c("date","spei_48")

spei_12tbl <- spei_12tbl |> drop_na()
spei_24tbl <- spei_24tbl |> drop_na()
spei_36tbl <- spei_36tbl |> drop_na()
spei_48tbl <- spei_48tbl |> drop_na()
```

# spei 12 1936-2025
```{r fig.align='center', fig.width=15, fig.height=10}
library(shadowtext)
library(ggplot2)
library(lubridate)

# Fixed classification with proper handling of NA values and non-overlapping ranges
spei_12tbl <- spei_12tbl |> 
  mutate(
    sign = ifelse(spei_12 > 0, "pos", "neg"),
    
    drought_category = case_when(
      is.na(spei_12)                      ~ NA_character_,  # Handle NA values explicitly
      spei_12 >= 2.0                      ~ "Extremely Wet",
      spei_12 >= 1.5                      ~ "Very Wet", 
      spei_12 >= 1.0                      ~ "Moderate Wet",
      spei_12 > 0.99                      ~ "Normal",        # Changed to > to avoid overlap
      spei_12 >= -0.99                    ~ "Normal",
      spei_12 >= -1.0                     ~ "Moderate Dry",  # Simplified range
      spei_12 >= -1.5                     ~ "Very Dry",      # Simplified range
      spei_12 < -2.0                      ~ "Extremely Dry", # Changed to < for clarity
      TRUE                                ~ "Normal"         # Catch-all for any missed values
    ),
    
    drought_category = factor(
      drought_category,
      levels = c(
        "Extremely Dry", "Very Dry", "Moderate Dry",
        "Normal",
        "Moderate Wet", "Very Wet", "Extremely Wet"
      )
    )
  )

# Check for any remaining NA values in drought_category
na_check <- sum(is.na(spei_12tbl$drought_category))
if(na_check > 0) {
  cat("Warning: There are", na_check, "NA values in drought_category\n")
  
  # Option 1: Remove rows with NA
  # spei_12tbl <- spei_12tbl |> filter(!is.na(drought_category))
  
  # Option 2: Replace NA with "Normal" (if that makes sense for your data)
  # spei_12tbl$drought_category[is.na(spei_12tbl$drought_category)] <- "Normal"
}

# Rest of your code remains the same...
# Summary statistics
drought_stats <- spei_12tbl |> 
  summarise(
    total_months = n(),
    drought_months = sum(spei_12 < -0.5, na.rm = TRUE),
    severe_drought_months = sum(spei_12 <= -1.5 & spei_12 >= -1.99, na.rm = TRUE),
    extreme_drought_months = sum(spei_12 <= -2, na.rm = TRUE),
    wet_months = sum(spei_12 > 0.5, na.rm = TRUE),
    min_spei = min(spei_12, na.rm = TRUE),
    max_spei = max(spei_12, na.rm = TRUE),
    min_date = date[which.min(spei_12)],
    max_date = date[which.max(spei_12)],
    drought_pct = round(drought_months / total_months * 100, 2),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 2),
    extreme_drought_pct = round(extreme_drought_months / total_months * 100, 2),
    wet_pct = round(wet_months / total_months * 100, 1)
  )

# Subtitle
stats_text <- paste0(
  "Drought pct: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Extreme drought: ", drought_stats$extreme_drought_pct,"% | ",
  "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), 
  " (SPEI: ", round(drought_stats$max_spei, 2), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), 
  " (SPEI: ", round(drought_stats$min_spei, 2), ")"
)

drought_colors <- c(
  "Extremely Dry"  = "#8B0000",
  "Very Dry"       = "#FF0000",
  "Moderate Dry"   = "#FFA07A",
  "Normal"         = "#CCCCCC",
  "Moderate Wet"   = "#87CEEB",
  "Very Wet"       = "#4682B4",
  "Extremely Wet"  = "#00008B"
)

date_range <- range(spei_12tbl$date, na.rm = TRUE)
mid_date <- mean(as.numeric(date_range)) |> as.Date(origin = "1970-01-01")
annotation_start <- mid_date %m-% years(20)
annotation_end   <- mid_date %m-% years(3)

# Plot with additional safeguards
ggplot(spei_12tbl |> filter(!is.na(drought_category)), # Filter out NA values for plotting
       aes(x = date, y = spei_12, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Key Insights box
  annotate("rect", xmin = annotation_start , xmax = annotation_end,
           ymin = 2.3, ymax = 3.45, fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +
  annotate("text", x = annotation_start, y = 3.3, 
           label = paste0("KEY INSIGHTS (", year(date_range[1]), "-", year(date_range[2]), ")"),
           size = 5, fontface = "bold", hjust = 0) +
  annotate("text", x = annotation_start, y = 3.1, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "%"),
           size = 5, hjust = 0, color = "#CC0000") +
  annotate("text", x = annotation_start, y = 2.9, 
           label = paste0("• Extreme Drought: ", drought_stats$extreme_drought_pct, "%"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +
  annotate("text", x = annotation_start, y = 2.65, 
           label = paste0("• Severe Drought: ", drought_stats$severe_drought_pct, "%"),
           size = 5, hjust = 0, color = "#B22222", fontface = "bold") +
  annotate("text", x = annotation_start, y = 2.45, 
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "%"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Extreme events annotation
  annotate("label",
           x = drought_stats$min_date,
           y = drought_stats$min_spei,
           label = paste0("Extreme Drought: ", format(drought_stats$min_date, "%b %Y"),
                          " (SPEI: ", round(drought_stats$min_spei, 2), ")"),
           fill = "#8B0000", color = "white", fontface = "bold", size = 4, label.size = NA, hjust = 1) +

  annotate("label",
           x = drought_stats$max_date,
           y = drought_stats$max_spei,
           label = paste0("Wettest Event: ", format(drought_stats$max_date, "%b %Y"),
                          " (SPEI: ", round(drought_stats$max_spei, 2), ")"),
           fill = "#00008B", color = "white", fontface = "bold", size = 5, label.size = NA, hjust = 0) +

  # Start/End
  annotate("text", x = min(spei_12tbl$date, na.rm = TRUE), y = 3, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40", fontface = "bold") +
  annotate("text", x = max(spei_12tbl$date, na.rm = TRUE), y = 3, 
           label = "End of analysis", hjust = 0.9, size = 5, color = "grey50", fontface = "bold") +

  # Axis and scales
  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1936-01-01", "2026-01-01")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    limits = c(-3.5, 3.5),
    breaks = seq(-3.5, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    na.translate = FALSE,  # This prevents NA from appearing in legend
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1936–2025*)",
    subtitle = paste0("SPEI (12-Month Scale) | ", stats_text),
    caption = "Source: DWD",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 16),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "grey10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 12),
    legend.key.size = unit(1, "cm"),
    #plot.margin = margin(t = 10, r = 5, b = 10, l = 5)
  )

```







```{r fig.align='center',fig.height=10,fig.width=15}
library(ggplot2)
library(lubridate)
library(shadowtext)
p12 <- 
p12
# Saving the plot
 # ggsave("/Users/mandarphatak/Documents/GitHub/Climate-Analysis/DWD_Folder/Plots/SPEI_12_Hamburg.png", plot = p12, width = 15, height = 10, dpi = 450)
```
```{r}
spei_12tbl <- spei_12tbl |> filter(date >"1989-12-01")

spei_12tbl <- spei_12tbl |> 
  mutate(
    sign = ifelse(spei_12 >0,"pos","neg"),
    drought_category=case_when(
      spei_12 >= 2.0 ~ "Extremely Wet",
      spei_12 >= 1.5 & spei_12 < 1.99 ~ "Very Wet", 
      spei_12 >= 1.0 & spei_12 < 1.49 ~ "Moderately Wet",
      spei_12 >= 0.5 & spei_12 < 0.99 ~ "Slightly Wet",
      spei_12 >= -0.5 & spei_12 < 0.5 ~ "Normal",
      spei_12 >= -1.0 & spei_12 < -0.5 ~ "Mild Drought",
      spei_12 >= -1.5 & spei_12 < -1.0 ~ "Moderate Drought",
      spei_12 >= -2.0 & spei_12 < -1.5 ~ "Severe Drought",
      TRUE ~ "Extreme Drought" 
    ),
     drought_category = factor(drought_category, levels = c(
      "Extreme Drought", "Severe Drought", "Moderate Drought", 
      "Mild Drought", "Normal", "Slightly Wet", 
      "Moderately Wet", "Very Wet", "Extremely Wet"
  ))
  
  )

drought_stats <- spei_12tbl |>
  summarise(
    total_months = n(),
    drought_months = sum(spei_12tbl$spei_12 < -0.5),
    severe_drought_months = sum(spei_12 < -1.5),
    wet_months = sum(spei_12tbl$spei_12 > 0.5),
    min_spei = min(spei_12),
    max_spei = max(spei_12tbl$spei_12),
    min_date = date[which.min(spei_12tbl$spei_12)],
    max_date = date[which.max(spei_12tbl$spei_12)],
    drought_pct = round(drought_months / total_months * 100, 1),
    severe_drought_pct = round(severe_drought_months / total_months * 100, 1),
    wet_pct = round(wet_months / total_months * 100, 1)
  )
# Create statistical summary for subtitle
stats_text <- paste0(
  "Drought conditions: ", drought_stats$drought_pct, "% | ",
  "Severe drought: ", drought_stats$severe_drought_pct, "% | ",
  "Wet events: ", drought_stats$wet_pct, "% | ",
  "Wettest: ", format(drought_stats$max_date, "%b %Y"), " (", round(drought_stats$max_spei, 2), ") | ",
  "Driest: ", format(drought_stats$min_date, "%b %Y"), " (", round(drought_stats$min_spei, 2), ")"
)

# Define color palette for drought categories
drought_colors <- c(
  "Extreme Drought" = "#8B0000",     # Dark Red
  "Severe Drought" = "#FF0000",      # Red  
  "Moderate Drought" = "#FF4500",    # Orange Red
  "Mild Drought" = "#FFA500",        # Orange
  "Normal" = "#CCCCCC",              # Light Gray
  "Slightly Wet" = "#87CEEB",        # Sky Blue
  "Moderately Wet" = "#4682B4",      # Steel Blue
  "Very Wet" = "#1E90FF",            # Dodger Blue
  "Extremely Wet" = "#0000CD"        # Medium Blue
)

#r fig.align='center', fig.height=12,fig.width=18}
 ggplot(spei_12tbl, aes(x = date, y = spei_12, fill = drought_category)) +
  geom_col(show.legend = TRUE, width = 20) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +

  geom_hline(yintercept = c(-2, -1.5, -1, -0.5, 0.5, 1, 1.5, 2), 
             color = "grey50", linewidth = 0.3, linetype = "dashed", alpha = 0.6) +

  geom_vline(xintercept = drought_stats$min_date, color = "#8B0000", linetype = "dotdash", linewidth = 0.8) +
  geom_vline(xintercept = drought_stats$max_date, color = "#00008B", linetype = "dotdash", linewidth = 0.8) +

  # Key Insights for 1990–2025
  annotate("rect", 
           xmin = as.Date("2008-01-01"), xmax = as.Date("2017-01-01"),
           ymin = 2.5, ymax = 3.45, 
           fill = "white", color = "grey60", alpha = 0.9, linewidth = 0.5) +

  annotate("text", x = as.Date("2008-01-01"), y = 3.3, 
           label = "KEY INSIGHTS (1990–2025)", size = 5, fontface = "bold", hjust = 0) +

  annotate("text", x = as.Date("2008-01-01"), y = 3.1, 
           label = paste0("• Drought conditions: ", drought_stats$drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#CC0000") +

  annotate("text", x = as.Date("2008-01-01"), y = 2.9, 
           label = paste0("• Severe drought: ", drought_stats$severe_drought_pct, "% of period"),
           size = 5, hjust = 0, color = "#8B0000", fontface = "bold") +

  annotate("text", x = as.Date("2008-01-01"), y = 2.7, 
           label = paste0("• Wet Events: ", drought_stats$wet_pct, "% of period"),
           size = 5, hjust = 0, color = "#00008B", fontface = "bold") +

  # Extreme dry event in recent period
  geom_shadowtext(
    data = data.frame(
      x = drought_stats$min_date,
      y = drought_stats$min_spei - 0.2,
      label = paste0("  Recent Drought ●\n",format(drought_stats$min_date, "%b %Y"),
                     " (SPEI: ", round(drought_stats$min_spei, 2), ")")
    ),
    mapping = aes(x = x, y = y, label = label),
    color = "white",
    bg.colour = "#8B0000",
    fontface = "bold",
    size = 4.5,
    hjust = 1,
    inherit.aes = FALSE
  ) +

  # Wettest event in recent period
  geom_shadowtext(
    data = data.frame(
      x = drought_stats$max_date,
      y = drought_stats$max_spei + 0.2,
      label = paste0("● Recent Wet\n",format(drought_stats$max_date,"%b %Y"),
                     " (SPEI:",round(drought_stats$max_spei, 2), ")")
    ),
    mapping = aes(x = x, y = y, label = label),
    color = "white",
    bg.colour = "#00008B",
    fontface = "bold",
    size = 4.5,
    hjust = 0,
    inherit.aes = FALSE
  ) +

  annotate("text", x = min(spei_12tbl$date), y = 3, 
           label = "Start of analysis", hjust = 0, size = 5, color = "grey40", fontface = "bold") +

  annotate("text", x = max(spei_12tbl$date), y = 3, 
           label = "End of analysis", hjust = 0.9, size = 5, color = "grey50", fontface = "bold") +

  scale_x_date(
    date_breaks = "5 years",
    date_labels = "%Y",
    limits = as.Date(c("1990-01-01", "2025-05-28")),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_y_continuous(
    limits = c(-3.5, 3.5),
    breaks = seq(-3.5, 3.5, 0.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +

  scale_fill_manual(
    name = "Condition",
    values = drought_colors,
    guide = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      nrow = 1,
      reverse = TRUE
    )
  ) +

  labs(
    title = "Drought Severity in Hamburg Fuhlsbüttel (1990–2025*)",
    subtitle = paste0("SPEI (12-Month Scale) | ", stats_text),
    caption = "Source: DWD",
    x = "", 
    y = ""
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.5),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 22, margin = margin(t = 10, b = 15)),
    plot.subtitle = element_text(color = "grey60", size = 18, margin = margin(b = 10)),
    plot.caption = element_text(color = "grey60", size = 15, hjust = 1, margin = margin(t = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, colour = "grey10", face = "bold"),
    axis.text.y = element_text(size = 18, colour = "grey10", face = "bold"),
    legend.position = "top",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key.size = unit(1, "cm")
  )

```



Drought Severity Trends in Hamburg Fuhlsbüttel (1936–2025 vs. 1990–2025)
A comparative analysis of the Standardized Precipitation Evapotranspiration Index (SPEI-12) reveals a marked shift in drought dynamics over Hamburg Fuhlsbüttel. During the full historical period (1936–2025), drought conditions were present in 28.5% of months, with severe drought occurring in 7.0%. In contrast, the recent period (1990–2025) experienced drought in 31.5% of months and severe drought in 7.1%, indicating a clear uptick in overall dryness.
Notably, the frequency of wet events has declined slightly (from 30.3% to 28.7%), and recent decades have seen a cluster of persistent dry spells, particularly after 2015. The driest event of the full record occurred in March 1960 (SPEI = –3.02), while the most severe recent drought was in April 2019 (SPEI = –2.74). This temporal shift toward more frequent and prolonged droughts signals increasing hydroclimatic stress, likely influenced by anthropogenic climate change.


```{r}
period_stats <- data.frame(
  period = c("1936–2025", "1990–2025"),
  drought_pct = c(28.5, 31.5),
  severe_drought_pct = c(7.0, 7.1),
  wet_pct = c(30.3, 28.7)
)

library(tidyr)
library(ggplot2)

# Pivot to long format
period_long <- pivot_longer(period_stats, 
                            cols = c(drought_pct, severe_drought_pct, wet_pct),
                            names_to = "Category", values_to = "Percent")

# Clean labels
period_long$Category <- recode(period_long$Category,
                               drought_pct = "Drought",
                               severe_drought_pct = "Severe Drought",
                               wet_pct = "Wet Events")

# Plot
ggplot(period_long, aes(x = Category, y = Percent, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("1936–2025" = "#a6bddb", "1990–2025" = "#2b8cbe")) +
  labs(title = "Comparison of Drought and Wet Event Frequencies",
       subtitle = "Hamburg Fuhlsbüttel (SPEI-12)",
       x = "", y = "Percent of Months",
       fill = "Period") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(face = "bold"),
        plot.title = element_text(face = "bold", size = 18),
        legend.position = "top")
```

```{r}
library(trend)

# Monthly binary drought indicator (e.g., SPEI < -0.5)
spei_12tbl <- spei_12tbl %>%
  mutate(is_drought = ifelse(spei_12 < -0.5, 1, 0),
         year = year(date)) %>%
  group_by(year) %>%
  summarise(drought_months = sum(is_drought),
            total_months = n(),
            drought_pct = 100 * drought_months / total_months)

# Mann–Kendall test on annual % of drought months
mk_test <- mk.test(spei_12tbl$drought_pct)
print(mk_test)

```
```{r}

```


```{r}
spei_12tbl <- spei_12tbl |> filter(year > 1989 )

# Mann–Kendall test on annual % of drought months
mk_test <- mk.test(spei_12tbl$drought_pct)
print(mk_test)

```

A Mann–Kendall trend test was conducted to evaluate changes in drought frequency over time based on annual percentages of months with SPEI-12 below –0.5.
For the full period (1936–2025), the trend is statistically insignificant (τ = 0.0057, p = 0.943), indicating no long-term change in drought frequency.
However, for the recent period (1990–2025), a moderate increasing trend is observed (τ = 0.195, p = 0.115), suggesting a potential rise in drought occurrence—though not statistically significant at the 5% level. This aligns with visual evidence and changing climatic dynamics post-1990.

