---
title: "SPEI"
author: "Tina"
date: "2024-09-08"
output: html_document
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 360, warning = FALSE,
                      fig.width = 8, fig.height = 6)

```
# libraries


```{r}
library(dplyr)
library(SPEI)
library(kableExtra)
library(readxl)
library(readr)
library(writexl)
library(tidyr)
library(tidyverse)
library(scales)
library(ggplot2)
library(plotly)
library(timetk)
library(broom)
library(trend)
library(lubridate)
library(trend)
library(Kendall)
library(ggtext)
```



```{r}
observational_data <- observational_data <- read.delim("/Users/mandarphatak/Downloads/Tina_Thesis/SPI_RDI_Data/Obseravtaionalfordroughtanalysis.txt", sep = ";") 
observational_data |> head()
```
```{r}
observational_tbl <- observational_data |> 
  rename(date= MESS_DATUM) |> 
  mutate(date=ymd(date)) |>
  dplyr::select(date,RSK,TMK,TNK,TXK,TGK)


```
```{r}
observational_tbl <- observational_tbl |> 
    mutate(RSK = ifelse(RSK == -999, NA, RSK)) 
    
```
```{r}


observation_tbl <- observational_tbl |> 
  rename(PREC = RSK)

observation_tbl
```

#### FPR SPEI we need PET and BAL . We has  used hargreaves method as it is more robust and suitable for our station
, although thornwaite could be a better alternative as our station is near the sea. Thornwaite may be suitable for our station. In our PET calculations using hargreaves , it seems that it is inflating the values.

```{r}
## We use the hargreaves method for PET, although
observation_tbl <- observation_tbl %>% 
    mutate(PET = hargreaves(Tmax = TXK, Tmin = TNK, Pre = PREC ,lat = 53.6332, na.rm = TRUE))
observation_tbl <- observation_tbl %>%
    mutate(BAL = PREC - PET)
```

## Annual Data Conversion

```{r}
# Step 3: Convert to annual data (aggregating sums of PREC and BAL, and mean of temperatures)
observation_annual_tbl <- observation_tbl %>%
    mutate(year = year(date)) %>%
    group_by(year) %>%
    summarize(
        PREC_annual = sum(PREC, na.rm = TRUE),
        BAL_annual = sum(BAL, na.rm = TRUE),
        Tavg_annual = mean(TMK, na.rm = TRUE),
        PET_annual = sum(PET, na.rm = TRUE),
        .groups = 'drop'
    )

bal_ts_annual <- ts(observation_annual_tbl$BAL_annual, start = 1936, frequency = 1)
```




```{r}
# Step 5: Compute SPEI with scales 12,9,6,3
spei_3 <- spei(bal_ts_annual, scale = 3)
spei_6 <- spei(bal_ts_annual, scale = 6)
spei_9 <- spei(bal_ts_annual, scale = 9)
spei_12 <- spei(bal_ts_annual, scale = 12)

# Step 6: Add SPEI results back to the annual data frame

spei_12 <- spei_12$fitted |> as_tibble()
spei_9 <- spei_9$fitted |> as_tibble()
spei_3 <- spei_3$fitted |> as_tibble()
spei_6 <- spei_6$fitted |> as_tibble()




colnames(spei_12) <- "spei_12"
colnames(spei_9) <- "spei_9"
colnames(spei_6) <- "spei_6"
colnames(spei_3) <- "spei_3"
years <- 1936:2023
# View the results
spei_12 <- data.frame(Year = years,Value = spei_12)
spei_6 <- data.frame(Year = years,Value = spei_6)
spei_3 <- data.frame(Year = years,Value = spei_3)
spei_9 <- data.frame(Year = years,Value =spei_9)
```

###. Plotting SPEI 
```{r}
spei_6 <- spei_6 |> filter(Year >= 1990)
spei_6 <- spei_6 |> mutate(sign = ifelse(spei_6 >=0, "pos","neg"))



trend_model <- lm(spei_6 ~ Year, data = spei_6 )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 3)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)



trend_model <- lm(spei_6 ~ Year, data = spei_6 )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 3)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)

spei_6_annual_plot <- spei_6 |> 
    group_by(sign) |> 
    ggplot() +
    geom_col(aes(x = Year, y = spei_6, fill = sign), show.legend = FALSE, width = 0.65) + 
    geom_smooth(aes(x = Year, y = spei_6), color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_continuous(breaks = seq(1990, 2023, by = 3), expand = c(0,0)) +
    scale_y_continuous(
        breaks = seq(-2.5, 2, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-2.5, 2, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 6 Annual",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) +
    theme_minimal() +
    coord_cartesian(xlim = c(1990, 2023), ylim = c(-2.5, 2)) +
   theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),
        plot.background = element_rect(fill = "grey", color = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    )  +
    annotate("text", x = 2017, y = 1.5,
             label = sprintf("Trend: %.3f", trend_value_rounded),
             hjust = 0, vjust = -0.5, size = 3.5, fontface = "bold", color = "black")
ggsave("spei_6_annual_plot.jpeg", plot = spei_6_annual_plot, width = 8, height = 6, dpi = 300, device = "jpeg")
```






```{r}


spei_12 <- spei_12 |> filter(Year >= 1990)
spei_12 <- spei_12 |> mutate(sign = ifelse(spei_12 >=0, "pos","neg"))



trend_model <- lm(spei_12 ~ Year, data = spei_12)
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 3)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)
spei_12_annual_plot <- spei_12 |> 
    group_by(sign) |> 
    ggplot() +
    geom_col(aes(x = Year, y = spei_12, fill = sign), show.legend = FALSE, width = 0.65) + 
    geom_smooth(aes(x = Year, y = spei_12), color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_continuous(breaks = seq(1990, 2023, by = 3), expand = c(0,0)) +
    scale_y_continuous(
        breaks = seq(-2.5, 2, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-2.5,2, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5)+
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 12 Annual",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) +
    theme_minimal() +
    coord_cartesian(xlim = c(1990, 2023), ylim = c(-2.5, 2)) +
   theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),  # Set panel background to light grey
        plot.background = element_rect(fill = "grey", color = NA),   # Set plot background to light grey
        panel.grid.minor = element_blank(),  # Keep minor grid lines off
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),  # Set major grid lines to black for visibility
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    annotate("text", x = 2017, y = 1.5,
             label = sprintf("Trend: %.3f", trend_value_rounded),
             hjust = 0, vjust = -1.5, size = 3.5, fontface = "bold", color = "black")

ggsave("spei_12_annual_plot.jpeg", plot = spei_12_annual_plot, width = 8, height = 6, dpi = 300, device = "jpeg")

```
```{r}

spei_9 <- spei_9 |> filter(Year >= 1990)
spei_9 <- spei_9 |> mutate(sign = ifelse(spei_9 >=0, "pos","neg"))



trend_model <- lm(spei_9 ~ Year, data = spei_9 )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 3)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)
spei_9_annual_plot <- spei_9 |> 
     group_by(sign) |> 
    ggplot() +
    geom_col(aes(x = Year, y = spei_9, fill = sign), show.legend = FALSE, width = 0.65) + 
    geom_smooth(aes(x = Year, y = spei_9), color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_continuous(breaks = seq(1990, 2023, by = 3), expand = c(0,0)) +
    scale_y_continuous(
        breaks = seq(-2.5, 2, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-2.5,2, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5)+
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 9 Annual ",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) +
    theme_minimal() +
    coord_cartesian(xlim = c(1990, 2023), ylim = c(-2.5, 2)) +
   theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),  # Set panel background to light grey
        plot.background = element_rect(fill = "grey", color = NA),   # Set plot background to light grey
        panel.grid.minor = element_blank(),  # Keep minor grid lines off
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),  # Set major grid lines to black for visibility
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    )+
    annotate("text", x = 2017, y = 1.5,
             label = sprintf("Trend: %.3f", trend_value_rounded),
             hjust = 0, vjust = -1.5, size = 3.5, fontface = "bold", color = "black")

ggsave("spei_9_annual_plot.jpeg", plot = spei_9_annual_plot, width = 8, height = 6, dpi = 300, device = "jpeg")
```
```{r}

spei_3 <- spei_3 |> filter(Year > 1989)
spei_3 <- spei_3 |> mutate(sign = ifelse(spei_3 >=0, "pos","neg"))



trend_model <- lm(spei_3 ~ Year, data = spei_3 )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 3)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)

spei_3_annual_plot <- spei_3 |> 
     group_by(sign) |> 
    ggplot() +
    geom_col(aes(x = Year, y = spei_3, fill = sign), show.legend = FALSE, width = 0.65) + 
    geom_smooth(aes(x = Year, y = spei_3), color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_continuous(breaks = seq(1990, 2023, by = 3), expand = c(0,0)) +
    scale_y_continuous(
        breaks = seq(-2.5, 2, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-2.5,2, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5)+
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 3 Annual",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) +
    theme_minimal() +
    coord_cartesian(xlim = c(1990, 2023), ylim = c(-2.5, 2)) +
   theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),  # Set panel background to light grey
        plot.background = element_rect(fill = "grey", color = NA),   # Set plot background to light grey
        panel.grid.minor = element_blank(),  # Keep minor grid lines off
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),  # Set major grid lines to black for visibility
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    )+
    annotate("text", x = 2017, y = 1.5,
             label = sprintf("Trend: %.3f", trend_value_rounded),
             hjust = 0, vjust = -1.5, size = 3.5, fontface = "bold", color = "black")

ggsave("spei_3_annual_plot.jpeg", plot = spei_3_annual_plot, width = 8, height = 6, dpi = 300, device = "jpeg")
```
### Monthly Plots for SPEI


## Monthly Data Conversion

```{r}
observation_monthly_tbl <- observation_tbl %>%
    mutate(year = year(date), month = month(date)) %>%
    group_by(year,month) %>%
    summarize(
        PREC_monthly = sum(PREC, na.rm = TRUE),
        BAL_monthly = sum(BAL, na.rm = TRUE),
        Tavg_month = mean(TMK, na.rm = TRUE),
        PET_month = sum(PET, na.rm = TRUE),
        .groups = 'drop'
    )

bal_ts_monthly <- ts(observation_monthly_tbl$BAL_monthly, start = 1936, frequency = 12)

```

#### monthly calculations for SPEI 

```{r}
# Step 5: Compute SPEI with scale 12
spei_3m <- spei(bal_ts_monthly, scale = 3)
spei_6m <- spei(bal_ts_monthly, scale = 6)
spei_9m <- spei(bal_ts_monthly, scale = 9)
spei_12m <- spei(bal_ts_monthly, scale = 12)

# Step 6: Add SPEI results back to the annual data frame

spei_12m <- spei_12m$fitted |> as_tibble()
spei_9m <- spei_9m$fitted |> as_tibble()
spei_3m <- spei_3m$fitted |> as_tibble()
spei_6m <- spei_6m$fitted |> as_tibble()




colnames(spei_12m) <- "spei_12"
colnames(spei_9m) <- "spei_9"
colnames(spei_6m) <- "spei_6"
colnames(spei_3m) <- "spei_3"
# Create the sequence of dates
date_seq <- seq(from = as.Date("1936-01-01"), to = as.Date("2023-12-01"), by = "month")

# View the first few dates to confirm
head(date_seq)

# View the last few dates to confirm
tail(date_seq)

# Check the length of the sequence
length(date_seq)
# View the results
spei_12m <- data.frame(date = date_seq,Value = spei_12m) 
spei_9m <- data.frame(date = date_seq,Value = spei_9m) 
spei_6m<- data.frame(date = date_seq,Value = spei_6m) 
spei_3m<- data.frame(date = date_seq,Value = spei_3m) 
```
## plots for the monthly spei


```{r}
spei_3m <- spei_3m |> filter(date > "1989-12-01")
spei_3m <- spei_3m |> mutate(sign = ifelse(spei_3 >=0, "pos","neg"))



trend_model <- lm(spei_3 ~ date, data = spei_3m )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 5)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)






library(ggplot2)
library(lubridate)

# Assuming your date column is already in Date format. If not, convert it:
# spei_3m$date <- as.Date(spei_3m$date)

spei_3_monthly_plot <- spei_3m |> 
    ggplot(aes(x = date, y = spei_3)) +
    geom_col(aes(fill = sign), position = position_identity(), show.legend = FALSE, width = 60) + 
    geom_smooth(color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_date(
        limits = c(as.Date("1990-01-01"), as.Date("2023-12-01")),
        date_breaks = "3 years",
        date_labels = "%Y",
        expand = c(0,0)
    ) +
    scale_y_continuous(
        breaks = seq(-3, 3, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-3, 3, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 3 Monthly",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) +
  coord_cartesian(ylim = c(-3,3)) +
    theme_minimal()  +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),
        plot.background = element_rect(fill = "grey", color = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    annotate("text", x = as.Date("2017-01-01"), y = 2,
             label = sprintf("Trend: %.0e", trend_value_rounded),
             hjust = 0, vjust = -1.5, size = 3.5, fontface = "bold", color = "black")

ggsave("spei_3_monthly_plot.jpeg", plot = spei_3_monthly_plot, width = 8, height = 6, dpi = 300, device = "jpeg")
```
```{r}
spei_6m <- spei_6m |> filter(date > "1989-12-01")
spei_6m <- spei_6m |> mutate(sign = ifelse(spei_6 >=0, "pos","neg"))
trend_model <- lm(spei_6 ~ date, data = spei_6m )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 5)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)


spei_6_monthly_plot <- spei_6m |> 
ggplot(aes(x = date, y = spei_6)) +
    geom_col(aes(fill = sign), position = position_identity(), show.legend = FALSE, width = 60) + 
    geom_smooth(color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_date(
        limits = c(as.Date("1990-01-01"), as.Date("2023-12-01")),
        date_breaks = "3 years",
        date_labels = "%Y",
        expand = c(0,0)
    ) +
    scale_y_continuous(
        breaks = seq(-3, 3, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-3, 3, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 6 Monthly",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) + coord_cartesian(ylim = c(-3,3)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),
        plot.background = element_rect(fill = "grey", color = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    annotate("text", x = as.Date("2017-01-01"), y = 2,
             label = sprintf("Trend: %.0e", trend_value_rounded),
             hjust = 0, vjust = -1.5, size = 3.5, fontface = "bold", color = "black")
ggsave("spei_6_monthly_plot.jpeg", plot = spei_6_monthly_plot, width = 8, height = 6, dpi = 300, device = "jpeg")
```
```{r}
spei_9m <- spei_9m |> filter(date > "1989-12-01")
spei_9m <- spei_9m |> mutate(sign = ifelse(spei_9 >=0, "pos","neg"))
trend_model <- lm(spei_9 ~ date, data = spei_9m )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 5)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)




 spei_9_monthly_plot <- spei_9m |> 
ggplot(aes(x = date, y = spei_9)) +
    geom_col(aes(fill = sign), position = position_identity(), show.legend = FALSE, width = 60) + 
    geom_smooth(color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_date(
        limits = c(as.Date("1990-01-01"), as.Date("2023-12-01")),
        date_breaks = "3 years",
        date_labels = "%Y",
        expand = c(0,0)
    ) +
    scale_y_continuous(
        breaks = seq(-3, 3, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-3, 3, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 9 Monthly",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) + coord_cartesian(ylim = c(-3,3)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),
        plot.background = element_rect(fill = "grey", color = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    annotate("text", x = as.Date("2017-01-01"), y = 2,
             label = sprintf("Trend: %.0e", trend_value_rounded),
             hjust = 0, vjust = -1.5, size = 3.5, fontface = "bold", color = "black")

ggsave("spei_9_monthly_plot.jpeg", plot = spei_9_monthly_plot, width = 8, height = 6, dpi = 300, device = "jpeg")


```
```{r}
spei_12m <- spei_12m |> filter(date > "1989-12-01")
spei_12m <- spei_12m |> mutate(sign = ifelse(spei_12 >=0, "pos","neg"))
trend_model <- lm(spei_12 ~ date, data = spei_12m )
trend_value <- coef(trend_model)[2]  # This is the slope
trend_value_rounded <- round(trend_value, 7)  # Round to 3 decimal places
trend_text <- paste0("Trend: ", trend_value_rounded)


spei_12_monthly_plot <- spei_12m |> 
ggplot(aes(x = date, y = spei_12)) +
    geom_col(aes(fill = sign), position = position_identity(), show.legend = FALSE, width = 60) + 
    geom_smooth(color = "red", method = "lm", se = FALSE, linewidth = 1) +
    scale_fill_manual(values = c("pos" = "darkgreen", "neg" = "#f5896b")) +
    scale_x_date(
        limits = c(as.Date("1990-01-01"), as.Date("2023-12-01")),
        date_breaks = "3 years",
        date_labels = "%Y",
        expand = c(0,0)
    ) +
    scale_y_continuous(
        breaks = seq(-3, 3, by = 0.5),
        sec.axis = sec_axis(~ ., name = "", breaks = seq(-3, 3, by = 0.5)),
        expand = c(0, 0)
    ) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    labs(
        title = "Hamburg Fuhlsbüttel Standardised Precipitation Evapotranspiration Index\nSPEI 12 Monthly",
        subtitle = "1990-2023",
        x = "", 
        y = ""
    ) + coord_cartesian(ylim = c(-3,3)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
        axis.text.y = element_text(face = "bold", color = "black"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "black", size = 1),
        panel.background = element_rect(fill = "grey", color = NA),  # Set panel background to light grey
        plot.background = element_rect(fill = "grey", color = NA),   # Set plot background to light grey
        panel.grid.minor = element_blank(),  # Keep minor grid lines off
        panel.grid.major = element_line(color = "lightgrey", size = 0.5),  # Set major grid lines to black for visibility
        axis.line.y.right = element_line(color = "black", size = 1),
        axis.text.y.right = element_text(face = "bold", color = "black"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    annotate("text", x = as.Date("2017-01-01"), y = 2,
             label = sprintf("Trend: %.2e", trend_value_rounded),
             hjust = 0, vjust = -1.5, size = 3.5, fontface = "bold", color = "black")
ggsave("spei_12_monthly_plot.jpeg", plot = spei_12_monthly_plot, width = 8, height = 6, dpi = 300, device = "jpeg")
```

