---
title: "Average_Temperature_Hamburg"
author: "Tina"
date: "2024-09-09"
output:
  word_document: default
  html_document: default
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 400, warning = FALSE,
                      fig.width = 15, fig.height = 10)

```



```{r}
library(dplyr)
library(SPEI)
library(kableExtra)
library(readxl)
library(readr)
library(writexl)
library(tidyr)
library(tidyverse)
library(scales)
library(ggplot2)
library(plotly)
library(timetk)
library(broom)
library(trend)
library(lubridate)
library(trend)
library(Kendall)
library(ggtext)

```



```{r}
observational_data <- observational_data <- read.delim("/Users/mandarphatak/Downloads/Tina_Thesis/SPI_RDI_Data/Obseravtaionalfordroughtanalysis.txt", sep = ";") 
observational_data |> head()
```
```{r}
observational_tbl <- observational_data |> 
  rename(date= MESS_DATUM) |> 
  mutate(date=ymd(date)) |>
  dplyr::select(date,RSK,TMK,TNK,TXK,TGK)
observational_tbl <- observational_tbl |> 
    mutate(RSK = ifelse(RSK == -999, NA, RSK)) 
    

observation_tbl <- observational_tbl |> 
  rename(PREC = RSK)
```
```{r}
observation_annual_tbl <- observation_tbl %>%
        mutate(year = year(date)) %>%
        group_by(year) %>%
        summarize(
            PREC = sum(PREC, na.rm = TRUE),  # Total annual precipitation
            Tavg = mean(TMK, na.rm = TRUE),  # Average annual temperature
            Tmax = mean(TXK, na.rm = TRUE),  # Average annual maximum temperature
            Tmin = mean(TNK, na.rm = TRUE),  # Average annual minimum temperature
            .groups = 'drop'
        ) 

    
```


## Average Temperatures: 

Plot is as follows



```{r}

# Calculate mean temperature and trend
mean_temp <- mean(observation_annual_tbl$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ year, data = observation_annual_tbl)
trend_value <- coef(trend_model)[2]  # Slope of the trend line

# Create the plot
ggplot(observation_annual_tbl, aes(x = year, y = Tavg)) +
    geom_ribbon(aes(ymin = 6.5, ymax = Tavg), fill = "#ffc288", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#FF8109", size = 2) +
    geom_line(color = "#FDA666") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.5) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = .75) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1936-2023%s —1936-2023 Trend\n%s(+0.0239°C/year)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    )  +
    scale_x_continuous(breaks = c(seq(1936, 2023, by = 10), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(6.5, 11.5),
        breaks = seq(6.5, 11.5, by = 0.5),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(43.7, 52.7, by = 1),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    coord_cartesian(xlim = c(1936,2023)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", linewidth  = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
   
    
    # Add label for mean temperature
    annotate("text", x = 1999, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 3, vjust = -1.5, color = "black", fontface="bold")
```
# We can use a plot in the following way
```{r}
library(dplyr)
library(gridExtra)
library(cowplot)

# Assuming observation_annual_tbl and mean_temp are already defined

# Calculate the trend
trend <- lm(Tavg ~ year, data = observation_annual_tbl)
trend_per_year <- coef(trend)[2]
mean_temp <- mean(observation_annual_tbl$Tavg, na.rm = TRUE)

# Create the main plot
p <- ggplot(observation_annual_tbl, aes(x = year, y = Tavg)) +
    geom_ribbon(aes(ymin = 6.5, ymax = Tavg), fill = "#ffc288", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#FF8109", size = 2) +
    geom_line(color = "#FDA666") +
    geom_smooth(method = "lm", color = "blue", se = FALSE, size = 1) +
    geom_smooth(method = "loess", color = "darkgreen", se = FALSE, size = 1, span = .2) +
    geom_hline(yintercept = mean_temp, color = "black", size = .75) +
    scale_x_continuous(breaks = c(seq(1936, 2023, by = 10), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(6.5, 11.5),
        breaks = seq(6.5, 11.5, by = 0.5),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(43.7, 52.7, by = 1),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    coord_cartesian(xlim = c(1936,2023)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", linewidth  = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    labs(x = "", y = "") +
    annotate("text", x = 1999, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 2.5, vjust = -1.5, color = "black", fontface="bold") 


# Create the title with legend elements
title <- ggdraw() + 
    draw_label("Hamburg Fuhlsbüttel Average Temperature", 
               fontface = "bold", x = 0, y = 0.10, hjust = 0, size = 14) +
    draw_label("1936-2023", x = 0.06, y = -.6, hjust = -0.3, size = 10, fontface = "bold") +
    draw_line(x = c(0.5, 0.53), y = c(-.6, -.6), color = "darkgreen", size = 1) +
    draw_text("LOESS", x = 0.531, y = -.6, hjust = 0, size = 10,fontface="bold") +
    draw_line(x = c(0.68, 0.72), y = c(-0.6, -.6), color = "blue", size = 1) +
    draw_text(paste0("1936-2023 Trend \n(", sprintf("%+.4f", trend_per_year), "°C/year)"), 
              x = 0.69, y = -.6, hjust = -0.3, size = 10,fontface="bold") 

# Combine the title and the main plot
final_plot <- plot_grid(title, p, ncol = 1, rel_heights = c(0.1, 0.9))

# Display the plot
print(final_plot)

```



### Analysis:- 

From the above plot We  see that the Average Temperature witness an increasing trend of 0.0239°C for each year. Thus we can summarize t that based on the trend value each year for the period 1936-2023 (88 years) the average temperature  might risen by an approximate 2°C. 




In order to confirm our prediction we divide the data in two periods and check for the trend


### Checking the Trend from 1936-1990

```{r}
data_36_90 <- observation_annual_tbl |> filter(year < 1990)

data_36_90$Tavg |> max()

mean_temp <- mean(data_36_90$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ year, data = data_36_90)
trend_value <- coef(trend_model)[2]  # S

data_36_90 |> 
ggplot( aes(x = year, y = Tavg)) +
    geom_ribbon(aes(ymin = 6.5, ymax = Tavg), fill = "#ffc288", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#ff9733", size = 3, shape = 20) +
    geom_line(color = "#ff942f") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.5) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = .75) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1936-1989%s —1936-1989 Trend\n%s(+ 0.0073°C/year)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    ) +
   scale_x_continuous(breaks = c(seq(1936, 1989, by = 5), 1989), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(6.5, 10.5),
        breaks = seq(6.5, 10.5, by = 0.5),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(43.7, 50.9, by = .9),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    coord_cartesian(xlim = c(1936,1989)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", size = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
  
    annotate("text", x = 1946, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 0.5, vjust = -1, color = "black", fontface="bold")
```
## For the second period 1990-2023

```{r}
data_90_2023 <- observation_annual_tbl |> filter(year >= 1990)

data_90_2023$Tavg |> max()

mean_temp <- mean(data_90_2023$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ year, data = data_90_2023)
trend_value <- coef(trend_model)[2]  # S

data_90_2023 |> 
ggplot( aes(x = year, y = Tavg)) +
    geom_ribbon(aes(ymin = 7, ymax = Tavg), fill = "#ffc288", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#ff9733", size = 3, shape = 20) +
    geom_line(color = "#ff942f") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.5) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = .75) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1990-2023%s —1990-2023 Trend\n%s(+ 0.032606°C/year)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    ) +
   scale_x_continuous(breaks = c(seq(1990, 2023, by = 5), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(7, 11),
        breaks = seq(6.5, 11, by = 0.5),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(44.6, 51.8, by = .9),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    coord_cartesian(xlim = c(1990,2023)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", size = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
  
    annotate("text", x = 1995, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 0.5, vjust = -1, color = "black", fontface="bold")
```


## Clearly we can see that the trend has increased 4.5 times in the period 1990-2023 in comparison to 1936-1989. The key result is also shows that the trend for 1990-2023 is even higher than for the whole time period 1936-2023.
 We can  summarize our findings that  as it is expected that  in last 34years(1990-2023) on an average the temperature has increased by 34 * 0.0326 = 1.1°C 

## In order to verify our analysis we perform he following calculations

```{r}


mean_1936_1990 <- data_36_90 |> 
  summarise(
    mean = mean(Tavg)
  )
mean_1990_2023 <- data_90_2023 |> 
  summarise(
    mean = mean(Tavg)
  )

dff = mean_1990_2023 - mean_1936_1990

print(dff)
```
## Our analysis is  correct that the average annual temperature has increased by 1.257°C  



We can also plot our Average temperature as follows
```{r}
Average_Temperature_Plot <- observation_annual_tbl |> 
    ggplot(aes(year,Tavg)) +
    geom_line(aes(color = "1"), size = 1,show.legend = FALSE) +
    geom_point(fill = "gray",aes(color = "1"), shape = 21, show.legend = TRUE)+
    geom_smooth(method = "lm", color = "blue", se = FALSE, size = 0.5,show.legend = FALSE) +
    theme_light() +
    geom_smooth(method = "loess",se = FALSE, aes(color ="2"), size = 1, span = 0.15,show.legend = FALSE) +
    scale_x_continuous(breaks = c(seq(1936,2023,10),2023), expand = c(0,0), limits = c(1936, 2023)) +
    scale_y_continuous(limits = c(6.5,11),
                       breaks = c(seq(6.5,11,.5),11.5), expand = c(0,0),
                       labels = function(x) paste0(format(x, nsmall = 1), " °C")) +
    scale_color_manual(name = NULL, 
                       breaks = c(1,2),
                       values = c("gray","black"),
                       labels = c("Annual Mean","Loess"),
                       guide = guide_legend(override.aes = list(shape =15,size =5))) +
    labs(
        x = "",
        y = "Temperature(C)°",
        title = "Average Temperature ",
        subtitle = "Hamburg Fuhlsbüttel "
    ) +
    theme(
        axis.ticks = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(margin = margin(b=10),colour  = "black",face="bold"),
        plot.subtitle = element_text(size = 8, margin = margin(b=10),face = "bold"),
        legend.position = c(0.10,0.90),
        legend.title = element_text(size = 0),
        legend.key.height = unit(10,"pt"),
        legend.margin = margin(0,0,0,0),
        axis.text.x = element_text(angle = 45, hjust = 1)
    )


print(Average_Temperature_Plot)
```











## We now try to see if similar case exists for the monthly frequency

```{r}
# Calculate mean temperature and trend

observation_monthly_tbl <- observation_tbl |> 
  mutate(
    year = year(date),
    month = month(date)
  ) |> 
  group_by(year, month) |> 
    summarize(
            PREC = sum(PREC, na.rm = TRUE),  # Total annual precipitation
            Tavg = mean(TMK, na.rm = TRUE),  # Average annual temperature
            Tmax = mean(TXK, na.rm = TRUE),  # Average annual maximum temperature
            Tmin = mean(TNK, na.rm = TRUE),  # Average annual minimum temperature
            .groups = 'drop'
        ) |> 
  mutate(date = make_date(year,month))

mean_temp <- mean(observation_monthly_tbl$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ date, data = observation_monthly_tbl)
trend_value <- coef(trend_model)[2] 

max(observation_monthly_tbl$Tavg)
```
## Plot for Average Temperature for monthly data

```{r}

observation_monthly_tbl |> 
  ggplot(aes(date,Tavg)) +
  geom_ribbon(aes(ymin = -8.5, ymax = Tavg), fill = "#fac0a5", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#FF8109", size = 1) +
    geom_line(color = "#FDA666") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.75) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = 1) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1936-2023%s —1936-2023 Trend\n%s(+6.741e-05°C/monthly)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    ) +
  scale_x_date(
    date_breaks = "3 years",  # Set breaks every 10 years
    date_labels = "%Y",
    expand = c(0, 0)
  ) +
  coord_cartesian(xlim = c(as.Date("1936-01-01"), as.Date("2023-12-31"))) +
  #scale_x_continuous(breaks = c(seq(1936, 2023, by = 10), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(-8.5, 25.5),
        breaks = seq(-8.5, 25.5, by = 2),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(16.7, 77.9, by = 5),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    theme_minimal() + 
    theme(
      
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", size = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "#3D4551", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
   # Add label for mean temperature
    annotate("text", x = as.Date("1999-12-01"), y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 3, vjust = -1.5, color = "black", fontface="bold")
  
  
```

## Analysis We can observe that there is a postive trend of 0.00006741746 for each month.So in 88 years  around an increase of 0.07°C in average temperature. Which is acceptable change given the monthly data. 


# Our focus will now shifts to  the Seasonal Variation in the Average Temperature

