---
title: "Temperature_final"
author: "Tina"
date: "2024-09-09"
output: html_document
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 400, warning = FALSE,
                      fig.width = 15, fig.height = 10)

```



```{r}
library(dplyr)
library(SPEI)
library(kableExtra)
library(readxl)
library(readr)
library(writexl)
library(tidyr)
library(tidyverse)
library(scales)
library(ggplot2)
library(plotly)
library(timetk)
library(broom)
library(trend)
library(lubridate)
library(trend)
library(Kendall)
library(ggtext)

```



```{r}
observational_data <- observational_data <- read.delim("/Users/mandarphatak/Downloads/Tina_Thesis/SPI_RDI_Data/Obseravtaionalfordroughtanalysis.txt", sep = ";") 
observational_data |> head()
```
```{r}
observational_tbl <- observational_data |> 
  rename(date= MESS_DATUM) |> 
  mutate(date=ymd(date)) |>
  dplyr::select(date,RSK,TMK,TNK,TXK,TGK)
observational_tbl <- observational_tbl |> 
    mutate(RSK = ifelse(RSK == -999, NA, RSK)) 
    

observation_tbl <- observational_tbl |> 
  rename(PREC = RSK)
```
```{r}
observation_annual_tbl <- observation_tbl %>%
        mutate(year = year(date)) %>%
        group_by(year) %>%
        summarize(
            PREC = sum(PREC, na.rm = TRUE),  # Total annual precipitation
            Tavg = mean(TMK, na.rm = TRUE),  # Average annual temperature
            Tmax = mean(TXK, na.rm = TRUE),  # Average annual maximum temperature
            Tmin = mean(TNK, na.rm = TRUE),  # Average annual minimum temperature
            .groups = 'drop'
        ) 

    
```


## Average Temperatures: Plot is as follows



```{r}

# Calculate mean temperature and trend
mean_temp <- mean(observation_annual_tbl$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ year, data = observation_annual_tbl)
trend_value <- coef(trend_model)[2]  # Slope of the trend line
```


```{r}
# Create the plot
ggplot(observation_annual_tbl, aes(x = year, y = Tavg)) +
    geom_ribbon(aes(ymin = 6.5, ymax = Tavg), fill = "#ffc288", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#FF8109", size = 2) +
    geom_line(color = "#FDA666") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.5) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = .75) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1936-2023%s —1936-2023 Trend\n%s(+0.0239°C/year)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    )  +
    scale_x_continuous(breaks = c(seq(1936, 2023, by = 10), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(6.5, 11.5),
        breaks = seq(6.5, 11.5, by = 0.5),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(43.7, 52.7, by = 1),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    coord_cartesian(xlim = c(1936,2023)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", size = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
   
    
    # Add label for mean temperature
    annotate("text", x = 1999, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 3, vjust = -1.5, color = "black", fontface="bold")
```



### Analysis:- 

From the above plot We  see that the Average Temperature witness an increasing trend of 0.0239°C for each year. Thus we can sum this to say that based on the trend for the period 1936-2023 (88 years), the average temperature has risen by an approximate 2°C. 






### Checking the Trend from 1936-1990

```{r}
data_36_90 <- observation_annual_tbl |> filter(year < 1990)

data_36_90$Tavg |> max()

mean_temp <- mean(data_36_90$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ year, data = data_36_90)
trend_value <- coef(trend_model)[2]  # S
```
```{r}
data_36_90 |> 
ggplot( aes(x = year, y = Tavg)) +
    geom_ribbon(aes(ymin = 6.5, ymax = Tavg), fill = "#ffc288", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#ff9733", size = 3, shape = 20) +
    geom_line(color = "#ff942f") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.5) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = .75) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1936-1989%s —1936-1989 Trend\n%s(+ 0.0073°C/year)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    ) +
   scale_x_continuous(breaks = c(seq(1936, 1989, by = 5), 1989), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(6.5, 10.5),
        breaks = seq(6.5, 10.5, by = 0.5),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(43.7, 50.9, by = .9),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    coord_cartesian(xlim = c(1936,1989)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", size = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
  
    annotate("text", x = 1946, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 0.5, vjust = -1, color = "black", fontface="bold")
```
Analysis:- From the above plot We can observe that there is a very small +ve trend, so over the period the change in temperature would be around = 54 * 0.0073



## Checking for Trend for last 30 years :- From 1990-2023

```{r}
data <- observation_annual_tbl |> 
  filter(year >=1990)
# Calculate mean temperature and trend
mean_temp <- mean(data$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ year, data = data)
trend_value <- coef(trend_model)[2]  # Slope of
```


```{r}
data |> 
ggplot( aes(x = year, y = Tavg)) +
    geom_ribbon(aes(ymin = 7, ymax = Tavg), fill = "#ffc288", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#FF8109", size = 2) +
    geom_line(color = "#FDA666") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.5) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = .75) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1990-2023%s —1990-2023 Trend\n%s(+ 0.0326°C/year)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    )  +
    scale_x_continuous(breaks = c(seq(1990, 2023, by = 5), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(7, 11),
        breaks = seq(7, 11, by = 0.5),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(44.6, 51.8, by = 1),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    coord_cartesian(xlim = c(1990,2023)) +
    theme_minimal() +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", size = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
  
    annotate("text", x = 1995, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 0.5, vjust = -1.5, color = "black", fontface="bold")
```
## From the above plot we can clearly observe tha trend for average temperature for last 30 Years has increased in comparsion the  to the whole time period.  What  inference can be drawn from the above plot is that Average Temperature has been rising at fast pace. Since the trend is on an average increase each year by 0.0326°C , so it means that there has been an increase in temperature by 1°C in last 30 Years. 






## We now try to see if similar case exists for the monthly frequency

```{r}
# Calculate mean temperature and trend

observation_monthly_tbl <- observation_tbl |> 
  mutate(
    year = year(date),
    month = month(date)
  ) |> 
  group_by(year, month) |> 
    summarize(
            PREC = sum(PREC, na.rm = TRUE),  # Total annual precipitation
            Tavg = mean(TMK, na.rm = TRUE),  # Average annual temperature
            Tmax = mean(TXK, na.rm = TRUE),  # Average annual maximum temperature
            Tmin = mean(TNK, na.rm = TRUE),  # Average annual minimum temperature
            .groups = 'drop'
        ) |> 
  mutate(date = make_date(year,month))

mean_temp <- mean(observation_monthly_tbl$Tavg, na.rm = TRUE)
trend_model <- lm(Tavg ~ date, data = observation_monthly_tbl)
trend_value <- coef(trend_model)[2] 

max(observation_monthly_tbl$Tavg)
```
## Plot for Average Temperature for monthly data

```{r}

observation_monthly_tbl |> 
  ggplot(aes(date,Tavg)) +
  geom_ribbon(aes(ymin = -8.5, ymax = Tavg), fill = "#fac0a5", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#FF8109", size = 1) +
    geom_line(color = "#FDA666") +
    geom_smooth(method = "lm", color = "black", se = FALSE, size = 0.75) +
    geom_hline(yintercept = mean_temp, color = "#818589", size = 1) +
    labs(
        title = "Hamburg Fuhlsbüttel Average Temperature",
        subtitle = sprintf(
            "1936-2023%s —1936-2023 Trend\n%s(+6.741e-05°C/monthly)",
            sprintf("%*s", 80, ""),  # Adjust 30 to change spacing
            sprintf("%*s", 100, "")   # Adjust 55 to change alignment
        ),
        x = "",
        y = ""
    ) +
  scale_x_date(
    date_breaks = "3 years",  # Set breaks every 10 years
    date_labels = "%Y",
    expand = c(0, 0)
  ) +
  coord_cartesian(xlim = c(as.Date("1936-01-01"), as.Date("2023-12-31"))) +
  #scale_x_continuous(breaks = c(seq(1936, 2023, by = 10), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(-8.5, 25.5),
        breaks = seq(-8.5, 25.5, by = 2),
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = seq(16.7, 77.9, by = 5),  
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))
    ) +
    theme_minimal() + 
    theme(
      
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", size = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "#3D4551", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
   # Add label for mean temperature
    annotate("text", x = as.Date("1999-12-01"), y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 3, vjust = -1.5, color = "black", fontface="bold")
  
  
```

## Analysis We can observe that there is a postive trend of 0.00006741746 for each month.So in 88 years  around an increase of 0.07°C in average temperature. Which is acceptable change given the monthly data. 


# Our focus now shifts to  the Seasonal Variation in the Average Temperature:- We would be using the annual data as less data points this making the trend clearly visible.




