---
title: "Data"
author: "Mandar"
date: "`r Sys.Date()`"
output: html_document
---

```{r }
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 360, warning = FALSE,
                      fig.width = 10, fig.height = 8)
```

# Data source:- NCEI National Center for Environmental Information

We want to get data for  HAMBURG FUHLSBUETTEL :- 
Step 1:- We need to find the inventory list 
url = "https://www.ncei.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt"
```{r}
library(tidyverse)
url <- "https://www.ncei.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt"
## We can extract the information  for availability of data for our desired station :
### GSN means that it is dedicated station , this data is available for most of the variables and for a long period of time
table <- read_fwf(url, 
          fwf_positions(
            start = c(1, 13, 21, 31, 39, 69,  78),  # Corrected start positions
            end   = c(11, 20, 30, 38, 65, 77, 93),  # Corrected end positions
            col_names = c("Station_ID", "Latitude", "Longitude", "Elevation", "Station_Name", "GSN_Flag", "WMO_ID")
          ),
          col_types = "cdddccc")  # Define column types

## We can see the id for our desired station
table |> 
  filter(Station_Name == "HAMBURG FUHLSBUETTEL")
```

## We are trying to clean the data and also ensure accessibilty is there. 
```{r}
## We observe that that the last column WMO_ID is character but it is an id. 
table$WMO_ID <- suppressWarnings(   as.numeric(table$WMO_ID))

## We find that only 4 stations are there which are under GCOS Surface Network
table |> 
  filter(GSN_Flag == "GSN" & startsWith(Station_ID, "GM"))

## In total we have 1,123 stations for Germany
table |> 
  filter(startsWith(Station_ID, "GM"))


```


Step2:- Extracting the data  for HAMBURG FUHLSBUETTEL

```{r}
# Downloading the file directly from the url:- for HAMBURG FUHLSBUETTEL
#url <- "https://www.ncei.noaa.gov/pub/data/ghcn/daily/by_station/GM000010147.csv.gz"


url <- "https://www.ncei.noaa.gov/pub/data/ghcn/daily/by_station/GM000010147.csv.gz"
destfile <- "GM000010147.csv.gz"

# Download the file
download.file(url, destfile, mode = "wb")

# Read the file from the downloaded location
data_hamburg <- read_csv(
  destfile,
  col_names = c("station_id", "date", "element", "value", "quality_flag", "source_flag", "observation_flag", "time_flag"),
  col_types = cols(
    station_id = col_character(),
    date = col_date(format = "%Y%m%d"),
    element = col_character(),
    value = col_double(),
    quality_flag = col_character(),
    source_flag = col_character(),
    observation_flag = col_character(),
    time_flag = col_character()
  ),
  na = c("NA", "")
)

saveRDS(object = data_hamburg, file = "/Users/mandarphatak/Climate-Analysis/data/hamburg.rds")
#unique(data_hamburg$element)

```




Step1:- We cleaned the data,and  scaled the data as the variables are measured as tenths of respective scales.



```{r}
# reading the rds file from the folder 
data <- readRDS("/Users/mandarphatak/Climate-Analysis/data/hamburg.rds")
# glimpse of the data

# data understanding , we don't need the station_id, and SNWD for our analysis
data <- data |> 
  select(station_id,date,element,value) |> 
  pivot_wider(names_from = element, values_from = value) |> 
  select(date:TAVG) |> 
  # data cleaning and manipulation
  mutate(
    # Calculate TAVG when we have both TMAX and TMIN but no TAVG
    TAVG = case_when(
      !is.na(TMAX) & !is.na(TMIN) & is.na(TAVG) ~ (TMAX + TMIN) / 2,
      TRUE ~ TAVG
    ),
     TMAX = case_when(
      is.na(TMAX) & !is.na(TMIN) & !is.na(TAVG) ~ 2 * TAVG - TMIN,
      TRUE ~ TMAX
    ),
    # Calculate TMIN when we have TMAX and TAVG but no TMIN
    TMIN= case_when(
      !is.na(TMAX) & is.na(TMIN) & !is.na(TAVG) ~ 2 * TAVG - TMAX,
      TRUE ~ TMIN
    )
  ) |> 
  mutate(across(c(TMAX, TAVG, TMIN, PRCP), ~ ./10))

# Saving the cleaned data
saveRDS(object = data, file = "/Users/mandarphatak/Climate-Analysis/data/hamburg_clean.rds")



```
Step2:-  Monthly data conversion. Since our data is of daily frequnecy we need to convert to monthly frequency

```{r}
data <- readRDS("/Users/mandarphatak/Climate-Analysis/data/hamburg_clean.rds") 
monthly_tbl <- data |> 
  mutate(year = year(date), 
         month = month(date)) |> 
  group_by(year, month) |> 
  summarise(
    TMAX = mean(TMAX, na.rm = TRUE),
    TMIN = mean(TMIN, na.rm = TRUE),
    TAVG = mean(TAVG, na.rm = TRUE),
    PRCP = sum(PRCP, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(date = make_date(year, month))
monthly_tbl

# monthly data
saveRDS(monthly_tbl, file = "/Users/mandarphatak/Climate-Analysis/data/monthly_tbl.rds")
```
### End of Part 1 :- Data Compiling ###


