---
title: "Maximum Temperature"
author: "Mandar"
date: "2024-09-26"
output: 
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: '2'
  word_document:
    toc: yes
    toc_depth: '2'
---

```{r}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 360, warning = FALSE,
                      fig.width = 8, fig.height = 6)
```


```{r}
library(tidyverse)
library(dplyr)
library(cowplot)
library(lubridate)
library(ggplot2)
library(ggtext)
library(readr)

```


```{r }
 observational_data <- read.delim("/Users/mandarphatak/Downloads/Tina_Thesis/SPI_RDI_Data/Obseravtaionalfordroughtanalysis.txt", sep = ";") 
observational_data |> head()
```
## Daily Data
```{r}
observational_tbl <- observational_data |> 
  rename(date= MESS_DATUM) |> 
  mutate(date=ymd(date)) |>
  dplyr::select(date,RSK,TMK,TNK,TXK,TGK)
observational_tbl <- observational_tbl |> 
    mutate(RSK = ifelse(RSK == -999, NA, RSK)) 
    

observation_tbl <- observational_tbl |> 
  rename(PREC = RSK)
```
```{r}
max_temp_tbl <- observation_tbl |> 
    mutate(year = year(date)) |> 
    group_by(year) |> 
    summarize(
        Tmax = mean(TXK)
    )
max_temp_tbl |> head()
```

## Plots for max temperatures
```{r}
# Calculate mean temperature and trend
mean_temp <- mean(max_temp_tbl$Tmax, na.rm = TRUE)
trend <- lm(Tmax ~ year, data = max_temp_tbl)
trend_per_year <- coef(trend)[2]  # Slope of the trend line

max_temp_tbl$Tmax |> max()

celsius_breaks <- seq(10.5, 15.5, by = 0.5)

fahrenheit_breaks <- celsius_breaks * 1.8 + 32

p <- ggplot(max_temp_tbl, aes(x = year, y = Tmax)) +
    geom_ribbon(aes(ymin = 10.5, ymax = Tmax), fill = "#e68b88", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#e17773", size = 2, fill = "#dd635e", shape = 21) +
    geom_line(color = "#C50003") +
    geom_smooth(method = "lm", color = "blue", se = FALSE, size = 1) +
    geom_smooth(method = "loess", color = "#03A66A", se = FALSE, size = 1, span = .15) +
    geom_hline(yintercept = mean_temp, color = "black", size = .75) +
    scale_x_continuous(breaks = c(seq(1936, 2023, by = 10), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(10.5, 15.5),
        breaks = celsius_breaks,
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = fahrenheit_breaks, 
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))) +
    coord_cartesian(xlim = c(1936,2023)) +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", linewidth  = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    labs(x = "", y = "") +
    annotate("text", x = 1999, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 2.5, vjust = -1.5, color = "black", fontface="bold") 
title <- ggdraw() + 
    draw_label("Hamburg Fuhlsbüttel Maximum Temperature", 
               fontface = "bold", x = 0, y = 0.10, hjust = 0, size = 14) +
    draw_label("1936-2023", x = 0.06, y = -.6, hjust = -0.3, size = 10, fontface = "bold") +
    draw_line(x = c(0.5, 0.53), y = c(-.6, -.6), color = "#03A66A", size = 1) +
    draw_text("LOESS", x = 0.531, y = -.6, hjust = 0, size = 10,fontface="bold") +
    draw_line(x = c(0.68, 0.72), y = c(-0.6, -.6), color = "blue", size = 1) +
    draw_text(paste0("1936-2023 Trend \n(", sprintf("%+.4f", trend_per_year), "°C/year)"), 
              x = 0.69, y = -.6, hjust = -0.3, size = 10,fontface="bold") 
# Combine the title and the main plot
final_plot <- plot_grid(title, p, ncol = 1, rel_heights = c(0.1, 0.9))

# Display the plot
print(final_plot)
```

### Comparison of  two time frames 

#### From 1936-1989
```{r}
max_temp_1936_1989_tbl <- max_temp_tbl |> filter (year < 1990)

mean_temp <- mean(max_temp_1936_1989_tbl$Tmax, na.rm = TRUE)
trend <- lm(Tmax ~ year, data = max_temp_1936_1989_tbl)
trend_per_year <- coef(trend)[2]  # Slope of the trend line

celsius_breaks <- seq(10.5, 14.5, by = 0.5)

fahrenheit_breaks <- celsius_breaks * 1.8 + 32

p <- ggplot(max_temp_1936_1989_tbl, aes(x = year, y = Tmax)) +
    geom_ribbon(aes(ymin = 10.5, ymax = Tmax), fill = "#e68b88", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#e17773", size = 2, fill = "#dd635e", shape = 21) +
    geom_line(color = "#C50003") +
    geom_smooth(method = "lm", color = "blue", se = FALSE, size = 1) +
    geom_smooth(method = "loess", color = "#03A66A", se = FALSE, size = 1, span = .15) +
    geom_hline(yintercept = mean_temp, color = "black", size = .75) +
    scale_x_continuous(breaks = c(seq(1936, 1989, by = 5), 1989), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(10.5, 15.5),
        breaks = celsius_breaks,
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = fahrenheit_breaks, 
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))) +
    coord_cartesian(xlim = c(1936,1989)) +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", linewidth  = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    labs(x = "", y = "") +
    annotate("text", x = 1969, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 2.5, vjust = -1.5, color = "black", fontface="bold") 
title <- ggdraw() + 
    draw_label("Hamburg Fuhlsbüttel Maximum Temperature", 
               fontface = "bold", x = 0, y = 0.10, hjust = 0, size = 14) +
    draw_label("1936-1989", x = 0.06, y = -.6, hjust = -0.3, size = 10, fontface = "bold") +
    draw_line(x = c(0.5, 0.53), y = c(-.6, -.6), color = "#03A66A", size = 1) +
    draw_text("LOESS", x = 0.531, y = -.6, hjust = 0, size = 10,fontface="bold") +
    draw_line(x = c(0.68, 0.72), y = c(-0.6, -.6), color = "blue", size = 1) +
    draw_text(paste0("1936-1989 Trend \n(", sprintf("%+.4f", trend_per_year), "°C/year)"), 
              x = 0.69, y = -.6, hjust = -0.3, size = 10,fontface="bold") 
# Combine the title and the main plot
final_plot <- plot_grid(title, p, ncol = 1, rel_heights = c(0.1, 0.9))

# Display the plot
print(final_plot)
```


### from 1990-2023
```{r}
max_temp_1990_2023_tbl <- max_temp_tbl |> filter(year>=1990)

mean_temp <- mean(max_temp_1990_2023_tbl$Tmax, na.rm = TRUE)
trend <- lm(Tmax ~ year, data = max_temp_1990_2023_tbl)
trend_per_year <- coef(trend)[2]  # Slope of the trend line

celsius_breaks <- seq(11, 16, by = .5)

fahrenheit_breaks <- celsius_breaks * 1.8 + 32

max_temp_1990_2023_tbl$Tmax |> max()
```


```{r}
p <- ggplot(max_temp_1990_2023_tbl, aes(x = year, y = Tmax)) +
    geom_ribbon(aes(ymin = 11, ymax = Tmax), fill = "#e68b88", alpha = 0.3, na.rm = TRUE) +
    geom_point(color = "#e17773", size = 2, fill = "#dd635e", shape = 21) +
    geom_line(color = "#C50003") +
    geom_smooth(method = "lm", color = "blue", se = FALSE, size = 1,na.rm = TRUE) +
    geom_smooth(method = "loess", color = "#03A66A", se = FALSE, size = 1, span = 0.25, na.rm = TRUE) +
    geom_hline(yintercept = mean_temp, color = "black", size = .75) +
    scale_x_continuous(breaks = c(seq(1990, 2023, by = 5), 2023), expand = c(0,0)) +
    scale_y_continuous(
        limits = c(11, 16),
        breaks = celsius_breaks,
        labels = function(x) paste0(format(x, nsmall = 1), " °C"),
        expand = c(0, 0),
        sec.axis = sec_axis(~ . * 1.8 + 32, 
                            breaks = fahrenheit_breaks, 
                            labels = function(x) paste0(format(x, nsmall = 1), " °F"))) +
    coord_cartesian(xlim = c(1990,2023)) +
    theme(
        plot.title = element_text(hjust = 0, face = "bold", colour = "black"),
        plot.subtitle = element_text(hjust = 0, face = "bold", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", color = "#3D4551"),
        axis.text.y = element_text(face = "bold", color = "#3D4551"),
        axis.title = element_text(face = "bold", color = "black"),
        axis.line = element_line(color = "#7393B3", linewidth  = 0.5),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "lightgrey", linewidth  = 0.5),
        axis.line.y.right = element_line(color = "#7393B3", linewidth = 0.5),
        axis.text.y.right = element_text(face = "bold", color = "#3D4551"),
        axis.title.y.right = element_text(face = "bold", color = "black", angle = 90),
        plot.margin = margin(t = 40, r = 40, b = 20, l = 20, unit = "pt")
    ) +
    labs(x = "", y = "") +
    annotate("text", x = 1999, y = mean_temp, 
             label = paste("Mean: ", round(mean_temp, 3), "°C"),
             hjust = 1, vjust = -1.5, color = "black", fontface="bold") 
title <- ggdraw() + 
    draw_label("Hamburg Fuhlsbüttel Maximum Temperature", 
               fontface = "bold", x = 0, y = 0.10, hjust = 0, size = 14) +
    draw_label("1990-2023", x = 0.06, y = -.6, hjust = -0.3, size = 10, fontface = "bold") +
    draw_line(x = c(0.5, 0.53), y = c(-.6, -.6), color = "#03A66A", size = 1) +
    draw_text("LOESS", x = 0.531, y = -.6, hjust = 0, size = 10,fontface="bold") +
    draw_line(x = c(0.68, 0.72), y = c(-0.6, -.6), color = "blue", size = 1) +
    draw_text(paste0("1990-2023 Trend \n(", sprintf("%+.4f", trend_per_year), "°C/year)"), 
              x = 0.69, y = -.6, hjust = -0.3, size = 10,fontface="bold") 
# Combine the title and the main plot
final_plot <- plot_grid(title, p, ncol = 1, rel_heights = c(0.1, 0.9))

# Display the plot
print(final_plot)
```

Analysis:-
For Hamburg Fuhlbüttel the maximum temperature for whole period 1936-2023 shows a trend of +0.0226, which means that temperature has increased by 1.9°C , while in last 34 years it seems to havr increased by 1.5°C based on the Trend. So we can say that last 34 years Hamburg Fuhlsbüttel has witnessed an increase in maximum temperatures.



